---
title: "processing_datasets"
author: "Rojashree Jayakumar"
format: html
editor: visual
---

## Processing Datasets

This script is used for processing the nine publicly available 10X Xenium subcellular spatial transcriptomics datasets over a range of different tissue types from human samples

A. Brain (AD),

B. Brain (GL),

C. Breast Cancer,

D. Lung,

E. Lung (High-Res),

F. Lymph Node,

G. Pancreas,

H. Ovarian,

I. Skin.

```{r}
#| message: false
#| include: false
#| warning: false
knitr::opts_knit$set(root.dir = here::here())
source("bin/functions.R")
library(scRNAseq)
library(data.table)
library(CellChat)
library(scran)
library(scuttle)
library(SingleCellExperiment)
library(tidyverse)
```

#### Loading the input paths and gene panels

```{r}
#| message: false
#| include: false
#| warning: false

ddir = "data/spatial_xenium/"
input_paths = paste0(ddir,list.files(ddir)) # note the file order
genes_file = paste0(input_paths, "/gene_panel.json")

print("Loaded input file paths and gene panels")
```

#### Loading and processing reference datasets

-   The reference datasets were downloaded and processed for annotation using singleR package for label transfer

-   The brain_ad reference is downloaded from CellXgene website. https://cellxgene.cziscience.com/collections/7c4552fd-8a6d-4da3-9854-2dfa8baca8bf

-   The brain_gl reference is downloaded from CellXgene website. https://cellxgene.cziscience.com/collections/999f2a15-3d7e-440b-96ae-2c806799c08c

-   The cancer reference is downloaded from CellXgene website. https://cellxgene.cziscience.com/collections/3f7c572c-cd73-4b51-a313-207c7f20f188

-   The lymph node reference is downloaded from the scRNAseq package using the version 2023-12-21 and path lymph_node.

-   The pancreas reference is downloaded from Suerat data.pancreasref.SeuratData Azimuth reference was used.

```{r}
#| message: false
#| include: true
#| warning: false

# Reference datasets
# brain_ad
brain_ad_ref = anndata::read_h5ad(file.path("data/references/brain_ad_ref.h5ad"))
brain_ad_ref = AnnData2SCE(brain_ad_ref)
n_cells <- nrow(brain_ad_ref)
set.seed(123)
subset_idx <- sample(n_cells, size = floor(0.1 * n_cells)) # for speed
brain_ad_ref <- brain_ad_ref[subset_idx, ]
brain_ad_ref <- mapping_function(ref_dataset = brain_ad_ref)

# brain_gl
brain_gl_ref = anndata::read_h5ad(file.path("data/references/brain_gl_ref.h5ad"))
brain_gl_ref = AnnData2SCE(brain_gl_ref)
brain_gl_ref <- mapping_function(ref_dataset = brain_gl_ref)

# breast cancer
cancer_ref = readRDS("data/references/cancer_reference.rds")
breat_cancer_ref = cancer_ref[,cancer_ref$disease == "breast cancer"]

# lung and lung_high_res
lung_ref = lung_high_res_ref =  cancer_ref[,cancer_ref$disease == "lung cancer"]

# lympnode
lymph_node_ref = fetchDataset("he-organs-2020", "2023-12-21", path = 'lymph_node')
lymph_node_ref <- lymph_node_ref[,!is.na(lymph_node_ref$Cell_type_in_merged_data)]

# ovarian
ovarian_ref = cancer_ref[,cancer_ref$disease == "ovarian cancer"]

#pancreas
pancreas_ref = readRDS("data/references/pancreas_ref.rds") # seurat data, no need to log
pancreas_ref = SingleCellExperiment(pancreas_ref)

#skin
skin_ref = cancer_ref[,cancer_ref$disease == "melanoma"]

reference_vec = c(brain_ad_ref, brain_gl_ref, breat_cancer_ref, lung_ref, lung_high_res_ref, lymph_node_ref, ovarian_ref, pancreas_ref, skin_ref)
```

#### Calculating communication scores

```{r}
log_vec <- rep(TRUE, length(input_paths))
log_vec[8] <- FALSE
cellchat_vec <- rep(TRUE, length(input_paths))


results_list <- Map(
  process_st_data,
  input_filepath = input_paths,
  genes_file     = genes_file,
  ref            = reference_vec,
  log            = log_vec,
  cellchat       = cellchat_vec
)

# save
names = c("Brain_ad",
             "Brain_gl",
             "Breast_cancer",
             "Lung",
             "Lung_High_Res",
             "LymphNode",
             "Ovarian",
             "Pancreas",
             "Skin")
for (i in seq_along(results_list)) {
  file_name <- paste0("dataset_", names[i], ".rds")
  file_path <- file.path("data/communication_scores", file_name)
  
  # Save the list element
  saveRDS(results_list[[i]], file = file_path)
}
```

#### Creating the data for training the model

The data used for training contains the cell communication scores

```{r}

train_data_list = list()
for (i in 1:length(results_list)){
  comm_scores = files[i]
  training = Map(function(obj, region){
  net_arr <- obj@net[["prob"]]
  net_df <- as.data.frame(as.table(net_arr))
  names(net_df) <- c("source", "target", "interaction_name", paste0(region, "_score"))
  pval_df <- as.data.frame(as.table(obj@net[["pval"]]))
  names(pval_df) <- c("source", "target", "interaction_name", paste0(region, "_pval"))
  df_scores <- merge(net_df, pval_df, by = intersect(colnames(net_df), colnames(pval_df)))
  LRsig <- obj@LR$LRsig
  LRsig_sub <- LRsig[, c(
    "pathway_name", "annotation", "is_neurotransmitter","interaction_name", "ligand", "ligand.location","ligand.family", "ligand.keyword", "ligand.secreted_type","ligand.transmembrane","receptor", "receptor.family", "receptor.keyword", "receptor.surfaceome_main", "receptor.surfaceome_sub", "receptor.adhesome", "receptor.secreted_type", "receptor.location","receptor.transmembrane"
  )]
  df <- merge(df_scores, LRsig_sub, by = "interaction_name")
  d.spatial <- obj@images$distance * 0.01
  diag(d.spatial) <- NaN
  P.spatial <- 1/d.spatial
  P.spatial[is.na(d.spatial)] <- 0
  diag(P.spatial) <- max(P.spatial, na.rm = TRUE) 
  df[[paste0(region,"_pspatial")]] <- mapply(function(l, r) {
    if (!is.na(l) && !is.na(r) && l %in% rownames(P.spatial) && r %in% colnames(P.spatial)) {
      P.spatial[l, r]
    } else {
      NA
    }
  }, df$source, df$target)
  df[paste0(region, "_P1")] = df[paste0(region, "_score")]/ df[[paste0(region,"_pspatial")]]
  df[paste0(region, "_P1")][is.na( df[paste0(region, "_P1")])] <- 0
  df
}, comm_scores, names(comm_scores))
  training_dataframe = Reduce(function(x,y) merge(x,y, by = intersect(colnames(x), colnames(y)), all = FALSE), training)
  training_dataframe$sample = filename
  train_data_list[[i]] <- training_dataframe
}

train_data <- data.table::rbindlist(train_data_list, fill = TRUE)
# add information from HPA database downloaded from https://www.proteinatlas.org/humanproteome/subcellular/data#locations 
hpa_data = fread("data/subcellular_location.tsv")
train_data <- merge(train_data, hpa_data[, .(`Gene name`, `Main location`)], 
            by.x = "ligand", by.y = "Gene name", 
            all.x = TRUE)
setnames(df, old = "Main location", new = "ligand.location_hpa")
train_data <- merge(train_data, hpa_data[, .(`Gene name`, `Main location`)], 
            by.x = "receptor", by.y = "Gene name", 
            all.x = TRUE)
setnames(df, old = "Main location", new = "receptor.location_hpa")

# remove non significant interactions
train_data = train_data[cell_pval<0.05,]
fwrite(train_data, "data/raw_data.csv")

```

**Single Cell Data Processing**

```{r}

# processing 
sce <- fetchDataset("zilionis-lung-2019", "2023-12-20", path = "human")
# subset to include only patient 6
sce = sce[,sce$Patient %in% c("p6")]
sce = sce[,sce$disease == "lung cancer"]
sce = sce[,sce$donor_id == "QianLUAD_LC_5"]
sce = scuttle::logNormCounts(sce)
sce$cell_type = factor(sce$cell_type, levels = unique(sce$cell_type))

# perform cellchat
data.input <- SingleCellExperiment::logcounts(sce)
meta <- as.data.frame(SingleCellExperiment::colData(sce)) # extract a dataframe of the cell labels
cellchat <- createCellChat(object = data.input, meta = meta, group.by = "cell_type",
                             datatype = "RNA")
CellChatDB <- CellChatDB.human
cellChatDB.use <- CellChatDB
cellchat@DB <- cellChatDB.use
cellchat <- subsetData(cellchat)
cellchat <- identifyOverExpressedGenes(cellchat, do.DE = FALSE, min.cells = 10)
# options(future.globals.maxSize = 250 * 1024^3)
cellchat <- identifyOverExpressedInteractions(cellchat)
cellchat <- computeCommunProb(cellchat, type = "triMean", nboot = 20)
cellchat <- filterCommunication(cellchat, min.cells = 10)
# save the cell chat object 
saveRDS(cellchat, "data/processed/single_cell/cellchat_singlecell.rds")

# extract the ligand receptor pairs
net_arr <- cellchat@net[["prob"]]
df <- as.data.frame(as.table(net_arr))
names(df) <- c("source", "target", "interaction_name", 'cell_score')
pval_df <- as.data.frame(as.table(cellchat@net[["pval"]]))
names(pval_df) <- c("source", "target", "interaction_name", 'cell_pval')
df <- merge(df, pval_df, by = intersect(colnames(df), colnames(pval_df)))
LRsig <- cellchat@LR$LRsig
LRsig_sub <- LRsig[, c(
    "pathway_name", "annotation", "is_neurotransmitter","interaction_name", "ligand", "ligand.location","ligand.family", "ligand.keyword", "ligand.secreted_type","ligand.transmembrane","receptor", "receptor.family", "receptor.keyword", "receptor.surfaceome_main", "receptor.surfaceome_sub", "receptor.adhesome", "receptor.secreted_type", "receptor.location","receptor.transmembrane"
  )]
df <- merge(df, LRsig_sub, by = "interaction_name")
# add information from HPA database downloaded from https://www.proteinatlas.org/humanproteome/subcellular/data#locations 
hpa_data = fread("data/subcellular_location.tsv")
df <- merge(df, hpa_data[, .(`Gene name`, `Main location`)], 
            by.x = "ligand", by.y = "Gene name", 
            all.x = TRUE)
setnames(df, old = "Main location", new = "ligand.location_hpa")
df <- merge(df, hpa_data[, .(`Gene name`, `Main location`)], 
            by.x = "receptor", by.y = "Gene name", 
            all.x = TRUE)
setnames(df, old = "Main location", new = "receptor.location_hpa")
df = df[cell_pval<0.05,]
write_csv(df,'data/test_data/raw_single_cell.csv')

```

#### Obtaining the individual cellchat formula parts

```{r}
source(compute_formula_parts.R)
log_vec <- rep(TRUE, length(input_paths))
log_vec[8] <- FALSE
cellchat_vec <- rep(FALSE, length(input_paths))

results_list <- Map(
  run_cellchat_formula,
  input_filepath = input_paths,
  genes_file     = genes_file,
  ref            = reference_vec,
  log            = log_vec, 
  cellchat       = cellchat_vec
)

# save
names = c("Brain_ad",
             "Brain_gl",
             "Breast_cancer",
             "Lung",
             "Lung_High_Res",
             "LymphNode",
             "Ovarian",
             "Pancreas",
             "Skin")
for (i in seq_along(results_list)) {
  file_name <- paste0("dataset_", names[i], ".rds")
  file_path <- file.path("data/communication_scores", file_name)
  
  # Save the list element
  saveRDS(results_list[[i]], file = file_path)
}
```
