---
title: "Figure 2"
author: "rshreej"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-expand: true
    number-sections: true
    theme: united
    highlight: tango
    code-fold: true
    code-tools: true
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| message: false
#| include: false
#| warning: false
library(data.table)
library(circlize)
library(ggvenn)
library(cowplot)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(fgsea)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot) 
library(patchwork)
library(ggbiplot)
```

# Figure Caption

Figure2: Training and evaluation of a hierarchical model for predicting
subcellular cell communication scores.

A. Schematic of the hierarchical model architecture, combining a
classifier and a regressor to predict communication scores in
subcellular regions.

B. ROC curves showing the classification performance of the model on
nine datasets, with the area under the curve (AUC) for each dataset
indicated in the legend.

C. Performance metrics across the nine datasets: left, violin plots
showing the distribution of metrics; right, barplots representing the
mean across datasets with standard error bars.

D. Composite metrics comparing different base learners (left, Models
1–4) and feature engineering approaches (right, Models A–D).

E. Composite metrics for models trained with and without spatial
features across different dataset combinations. A total of 4,590 models
were trained (each of the 9 datasets trained across 255 combinations of
other datasets); plots show mean composite metrics with error bars.

```{r}
# adding the schematic here
```

```{r}
#| message: false
#| include: true
#| warning: false
theme_set(
  theme_classic(base_size = 14) +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
    )
)
```

### D. Composite metrics comparing different base learners (left, Models
1–4) and feature engineering approaches (right, Models A–D).

```{r}
#| message: false
#| include: true
#| warning: false
#| label: baselearners_feature_engineering

# load all the metrics obtained from training the base learners from analysis/baselearners.ipynb
ddir = "data/training_data/"
xgb = fread(file.path(ddir, "comparison_baselearners/xgboost_metrics.csv")) # Model1
rf = fread(file.path(ddir, "comparison_baselearners/rf_metrics.csv")) # Model2
xgb_rf = fread(file.path(ddir, "comparison_baselearners/xgboost_rf_metrics.csv")) # Model3
xgb_voting = fread(file.path(ddir, "comparison_baselearners/xgboost_voting_metrics.csv")) # Model4

xgb = xgb[, model:= "xgb"]
rf = rf[, model:= "rf"]
xgb_rf = xgb_rf[, model:= "xgb_rf"]
xgb_voting = xgb_voting[, model:= "xgb_voting"]

dt = rbind(xgb, rf, xgb_rf, xgb_voting)
dt[ , geometric_mean := (auc * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]

heatmap_df = melt(dt, id.vars= c('model', 'Dataset'), measure.vars= 23) 

# mapping
heatmap_df$Dataset <- factor(heatmap_df$Dataset,
  levels = unique(heatmap_df$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

heatmap_df$model <- factor(heatmap_df$model,
  levels = unique(heatmap_df$model),
  labels = c("Model 1", "Model 2", "Model 3", "Model 4")
)

colors = c("#eea3b5", "#c58fb0", "#9a7ea3", "#6f6d8f")

# plot the heatmap
p1 = ggplot(heatmap_df, aes(x = model, y = value, fill = model), color = ) +
  geom_boxplot() + 
  scale_fill_manual (values = colors) + 
  labs(x = "", y = "Composite metric") +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    legend.position = "",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  )

# load all the metrics obtained from training the base learners from analysis/feature_engineering.ipynb

ddir = "data/processed/plots_metrics"
a = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_ohe_not_separated_model2.csv")) # Modela
b = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_catboost_not_separated_model3.csv")) # Modelb
c = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_combined_model4.csv")) # Modelc
d = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_model1.csv")) #Modeld (same as model4)

a = a[, model:= "Model A"]
b = b[, model:= "Model B"]
c = c[, model:= "Model C"]
d = d[, model:= "Model D"]

dt = rbind(a, b, c, d)
dt[ , geometric_mean := (auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]


heatmap_df = melt(dt, id.vars= c('model', 'Dataset'), measure.vars= 23) 

# mapping
heatmap_df$Dataset <- factor(heatmap_df$Dataset,
  levels = unique(heatmap_df$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

# plot the heatmap
p2 = ggplot(heatmap_df, aes(x = model, y = value, fill = model), color = ) +
  geom_boxplot() + 
  scale_fill_manual (values = colors) + 
  labs(x = "", y = "Composite metric") +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    legend.position = "",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  )
ggarrange(p1,p2, align = "h")

ggsave("data/results/Final_Figures/Final_figure2/heatmap_featureengineering.pdf")
```

### B. ROC curves showing the classification performance of the model on nine datasets, with the area under the curve (AUC) for each dataset indicated in the legend. 

```{r}
#| message: false
#| include: true
#| warning: false
#| label: ROC-curve
# AUC curve 
# load the roc_tables output from experiments/run_logo.py with the spatial information 
roc_table = fread("data/processed/roc_tables.csv")
xgb_voting = fread(file.path(ddir, "comparison_baselearners/xgboost_voting_metrics.csv"))
roc_table = inner_join(roc_table, xgb_voting[,.(Dataset, auc)], by = 'Dataset')
roc_table$Dataset <- factor(roc_table$Dataset,
  levels = unique(roc_table$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

roc_table$dataset_label <- paste0(roc_table$Dataset, " (AUC=", round(roc_table$auc, 2), ")")

colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )
ggplot(roc_table, aes(x = `False Positive Rate`, y = `True Positive Rate`, color = dataset_label)) +
  geom_line(size = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "False Positive Rate",
       y = "True Positive Rate") +
  scale_color_manual(values = colors) + 
  theme_classic(base_size = 20) +
  theme(legend.position = "right") +
  guides(color = guide_legend(title = "Dataset"))

ggsave("data/results/Final_Figures/Final_figure2/roc_table.pdf", width =10)


```

### C. Performance metrics across the nine datasets: left, violin plots
showing the distribution of metrics; right, barplots representing the
mean across datasets with standard error bars.

```{r}
#| message: false
#| include: true
#| warning: false
#| label: ROC-curve

#  metrics plot
# the file with the metrics output from experiments/run_log.py is processed using extract_metrics function from utils.py and loaded
xgb_voting = fread(file.path(ddir, "comparison_baselearners/xgboost_voting_metrics.csv"))
xgb_voting = xgb_voting[,c(1,12,15,16,18,19,21)]
xgb_voting[ , geometric_mean := (auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]

dt = melt(xgb_voting, id.vars= c('Dataset')) 
dt$variable <- factor(dt$variable,
  levels = c("auc","macro_recall","cyt_R²","nuc_R²",
             # "cyt_RMSE","nuc_RMSE",
             "cyt_NRMSE","nuc_NRMSE","geometric_mean"),
  labels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "Composite metric")
)
#  metrics 
p1 = ggplot(dt, aes(x = variable, y = value, fill = variable, color = variable)) +
  geom_violin(trim = TRUE) +
  geom_boxplot(width = 0.15, color = "black", fill = "white", outlier.size = 0.8) +
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  labs(
    x = NULL,
    y = 'Value'
  ) +
  theme_classic(base_size = 20) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1,vjust = 1)
) + coord_cartesian(ylim = c(0,1))

p2 = ggplot(dt, aes(x = variable, y = value)) +
  stat_summary(fun = mean, geom = "bar", width = 0.7, aes(fill = variable)) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
               geom = "errorbar", width = 0.2, linewidth = 0.6) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.8, color = "black") +
  scale_fill_manual(values = colors)+ 
  labs(
    x = "", y = "Value"
  ) +
  theme_classic(base_size = 20) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) + coord_cartesian(ylim = c(0,1))

ggarrange(p1,p2, align = "h")
ggsave("data/results/Final_Figures/Final_figure2/combined_best_model.pdf", width =15, height = 6)
```

### E. Composite metrics for models trained with and without spatial
features across different dataset combinations. A total of 4,590 models
were trained (each of the 9 datasets trained across 255 combinations of
other datasets); plots show mean composite metrics with error bars.
```{r}
#| message: false
#| include: true
#| warning: false
#| label: composite_metric_combinations
# the file with the metrics output from experiments/run_combinationss.py is processed using extract_metrics function from utils.py and loaded
metrics = fread("data/processed/metrics_all_combinations_trainings_10032025.csv")
metrics[ , geometric_mean := (roc_auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]

df_long <- metrics[,c(9,28,12)]

df_summary <- df_long %>%
  group_by(combo_size, condition) %>%
  summarise(mean_val = mean(geometric_mean, na.rm = TRUE),
            sd_val = sd(geometric_mean, na.rm = TRUE)/sqrt(n()), .groups = "drop")

colors = c('with_spatial' = "#b715b7", "without_spatial" = "#14bf81")
ggplot(df_summary, aes(x = combo_size, y = mean_val, color = condition, group = condition)) +
  geom_line() +
  geom_point() +
  scale_color_manual(
  name   = "Model type",
  values = colors,
  labels = c(
    with_spatial    = "With spatial information",
    without_spatial = "Without spatial information"
  ),
  guide = guide_legend(nrow = 2, byrow = TRUE)
) + 
  geom_errorbar(aes(ymin = mean_val - sd_val, ymax = mean_val + sd_val), width = 0.2) +
  # facet_wrap(~metric, scales = "free_y", ncol = 4) +
  labs( x = "Number of training datasets", y = "Average of Composite metric") + 
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  )

ggsave("data/results/Final_Figures/Final_figure2/trainingcombinations.pdf", height = 4, width = 4.5)
```
