---
title: "Figure 2"
author: "rshreej"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-expand: true
    number-sections: true
    theme: united
    highlight: tango
    code-fold: true
    code-tools: true
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| message: false
#| include: false
#| warning: false
library(data.table)
library(tidyverse)
library(circlize)
library(ggvenn)
library(cowplot)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(fgsea)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot) 
library(patchwork)
library(ggbiplot)
library(reshape2)
library(ThemePark)
```

```{r}
theme_set(
  theme_classic(base_size = 22) +
    theme(
      axis.text = element_text(size = 20),
      axis.title = element_text(size = 22),
      legend.text = element_text(size = 22),
      legend.title = element_text(size = 22),
      plot.title = element_text(size = 22, hjust = 0.5)
    )
)
```

# Figure Caption

Figure 3: Assessing model stability by training each dataset with
varying combinations of the remaining datasets. Each dataset was trained
using 255 different combinations of the other eight datasets to evaluate
model stability.

A. Percentage of significant ligand–receptor (LR) pairs correctly
classified when each dataset was trained with 255 different combinations
of the other eight datasets.

B. Number of training combinations in which each significant LR pair
across the nine datasets was correctly classified: left, histogram of
correctly classified LR pairs; right, violin plots showing the
distribution of the number of combinations in which each LR pair was
correctly classified.

C. Counts of LR pairs consistently assigned to difficulty groups. Bars
represent the number of significant LR pairs that were classified as
Easy (≥75%), Intermediate (25–75%), or Hard (\<25%) based on the
proportion of training combinations in which they were correctly
classified across all datasets. Gray overlays indicate LR pairs that
were always assigned to the same difficulty group across all
combinations.

D. Violin plots of performance metrics across the 255 training
combinations for all nine datasets (outliers with cytoplasmic and
nuclear 𝑅2 \< 0 are not shown). E. Bar plot of LR pairs classified as
Hard; fractions shown to the right of each bar indicate the proportion
of combinations in which each LR pair was labeled as Hard.


### A. Percentage of significant ligand–receptor (LR) pairs correctly classified when each dataset was trained with 255 different combinations of the other eight datasets. 

```{r}
#| message: false
#| include: true
#| warning: false
#| label: classification_compartments
# process the outputs from run_compartments using the extract_metrics from utils.py and load
dir = "data/processed/scores_training/classification_raw"
files = list.files(dir)[1:9]
file_list = list()

for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(dir, i))
  dat$correct = ifelse(dat$pred == dat$true, 1, 0)
  x = nrow(dat)/255
  dat$id = rep(1:x, 255)
  data = dcast(dat, id ~ group, fun.aggregate = NULL, value.var = "correct")
  data$sample = sub("classification_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = data
}

prediction = setDT(do.call(rbind, file_list))
rownames(prediction) = 1:nrow(prediction)
col = prediction[, lapply(.SD, sum), by = sample, .SDcols = c(2:256)]
percent <- prediction[, lapply(.SD, function(x) sum(x) *100 / .N), 
                      by = sample, .SDcols = 2:256]

counts <- prediction[, .N, by = sample]
counts$sample <- factor(counts$sample,
  levels = unique(counts$sample),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)


colsums = melt(col, id.vars = "sample")
percent = melt(percent, id.vars = "sample")


percent$sample <- factor(percent$sample,
  levels = unique(percent$sample),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

p0 = ggplot(percent, aes(x = sample, y = value)) +
  # geom_violin(fill = "#ccebc5", width = 2) + 
  geom_boxplot(fill = "#ccebc5", width = 0.5, outlier.size = 0.5) + 
  # geom_jitter(width = 0.1, alpha = 0.3, size = 1) + 
  geom_text(
    data = counts,
    aes(x = sample, y = max(percent$value)-10, label = paste0("N=", N)),
    inherit.aes = FALSE,
    color = "#B43757",
    size = 5,
    angle = 90,
    hjust = 0,
    vjust =0
  ) + 
  labs(x = "", y = "% LR pairs classified correctly")  + 
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 18),
      legend.text = element_text(size = 18),
      legend.title = element_text(size = 18),
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.text.x = element_text(angle = 45,  hjust = 1, vjust = 1)
    )


# setDT(percent)
# percent <- percent[, .(mean_value = mean(value)), by = sample]

p1 <- ggplot(percent, aes(x = sample, y = value)) +
  stat_summary(fun = mean, geom = "bar", width = 0.7, fill = "#bebadb") +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
               geom = "errorbar", width = 0.2, linewidth = 0.6) +
  # geom_jitter(width = 0.001, size = 2, alpha = 0.8, color = "black")  +
  labs(x = "", y = "Mean % of correct classifications \n across the different combinations") +
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 18),
      legend.text = element_text(size = 18),
      legend.title = element_text(size = 18),
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.text.x = element_text(angle = 45,  hjust = 1, vjust = 1)
    )

ggarrange (p1, p0, common.legend = F)
ggsave("data/results/Final_Figures/Final_figure3/counts_plot.pdf", width = 8)
```

### B. Number of training combinations in which each significant LR pair
across the nine datasets was correctly classified: left, histogram of
correctly classified LR pairs; right, violin plots showing the
distribution of the number of combinations in which each LR pair was
correctly classified.

```{r}
#| message: false
#| include: true
#| warning: false
#| label: combos_correct

# id is not unique
row <- prediction[, .(value = rowSums(.SD)), .SDcols = 2:256]
row$percent <- row$value * 100 / 255
row$sample = prediction$sample  
p0 = ggplot(row, aes(x = value)) +
  geom_histogram(aes(y = after_stat(count / sum(count))),
                 binwidth = 5, fill = "#fb8073", color = "#fb8073") +
  scale_y_continuous(labels = scales::percent) +
  # scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  labs(x = "Number of Combinations", 
       y = "Proportion (%) of LR pairs classified correctly") +
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 14),
      axis.title = element_text(size = 18),
      legend.text = element_text(size = 18),
      legend.title = element_text(size = 18),
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.text.x = element_text(angle = 45,  hjust = 1, vjust = 1)
    )

p1 = ggplot(row, aes(x = "", y = value)) +
  geom_violin(fill = "#ffffb4", width = 0.6) + 
  geom_boxplot(color = "black", fill = "white", width = 0.05) + 
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )  + 
  labs(x = "Correctly classified LR pairs", y = "Number of combinations") +
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 18),
      legend.text = element_text(size = 18),
      legend.title = element_text(size = 18),
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.ticks.x = element_blank()
    )
ggarrange (p0, p1, align = "hv")
ggsave("data/results/Final_Figures/Final_figure3/lr_counts_plot.pdf", width = 8)

row$sample <- factor(row$sample,
  levels = unique(row$sample),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )

fwrite(row, "plot_supplementary7a")
```

### C. Counts of LR pairs consistently assigned to difficulty groups. Bars
represent the number of significant LR pairs that were classified as
Easy (≥75%), Intermediate (25–75%), or Hard (\<25%) based on the
proportion of training combinations in which they were correctly
classified across all datasets. Gray overlays indicate LR pairs that
were always assigned to the same difficulty group across all
combinations. E. Bar plot of LR pairs classified as
Hard; fractions shown to the right of each bar indicate the proportion
of combinations in which each LR pair was labeled as Hard.

```{r}
#| message: false
#| include: true
#| warning: false
#| label: easy_hard

# Easy is LR pairs with greater than 3rd quartile 
# Intermediate is LR pairs between 2nd and 3rd quartile
# Hard is LR pairs less than 1st quartile

prediction = setDT(do.call(rbind, file_list))
rownames(prediction) = 1:nrow(prediction)

# offset the rownames by the number of lr pairs
offsets <- prediction[, .N, by = sample][
  , offset := shift(cumsum(N), fill = 0)]
prediction <- merge(prediction, offsets[, .(sample, offset)], by = "sample")
prediction[, gid := id + offset]
# check 
x = as.integer(rownames(prediction))
y = prediction$gid
identical(x,y)


row <- prediction[, .(value = rowSums(.SD)), by = gid, .SDcols = 2:256]
row$percent <- row$value * 100 / 255
row[, group := fcase(
  percent > 75,  "Easy",
  percent < 25,  "Hard",
  default = "Intermediate"
)]

# join this with the dataframe of the features
training_data = read_csv("data/processed/training_data_071025.csv")
z = as.integer(rownames(training_data))
identical(y,z)
data = cbind(row, training_data[,1:38]) 

prop_df <- data[, .N, by = .(lr_pair, group)]
prop_df[, total := sum(N), by = lr_pair]
prop_df[, prop := N / total]

# Irrespective of cell type how many LR pairs are always Easy/ hard / intermediate
# There are 290 unique LR pairs and most of them are always easy to classify (it means that irrespective of celltype or dataset, that LR pairs have been classified 75% of times correctly)
# count all LR pairs per group
all_counts <- prop_df[, .N, by = group]

# count only "always" LR pairs per group
always = prop_df[prop == 1,]
always_counts <- always[, .N, by = group]
all_counts$group = factor(all_counts$group, levels = c("Easy", "Intermediate", "Hard"))

p1 <- ggplot(all_counts, aes(x = group, y = N, fill = group)) +
  geom_col(alpha = 0.6) +  # base bars: all pairs
  geom_col(data = always_counts, aes(x = group, y = N), 
           fill = "gray", width = 0.4) +  # overlay: always pairs
  labs(x = "Group", 
       y = "Number of LR pairs") + 
  guides(fill = "none") 
ggsave('data/results/Final_Figures/Final_figure3/easy_hard.pdf', height = 5)


hard_df <- prop_df[group == "Hard"]

p2 = ggplot(hard_df, aes(x = reorder(lr_pair, -prop), y = prop)) +
  geom_col(fill = "#85d692") +
  geom_text(aes(label = paste0(N, "/", total)), vjust = 0.5, hjust = 0, size = 6) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Hard LR pair", y = "Proportion of hard calls")

ggsave('data/results/Final_Figures/Final_figure3/hard_prop.pdf', width = 10, height = 12)
```

### D. Violin plots of performance metrics across the 255 training
combinations for all nine datasets (outliers with cytoplasmic and
nuclear 𝑅2 \< 0 are not shown). 


```{r}
#| message: false
#| include: true
#| warning: false
#| label: boxplot_metric_compartment

# process the outputs from run_compartments to extract all other metrics 
class_dir <- "data/processed/scores_training/classification_metrics/"

files = list.files(class_dir)
file_list = list()

for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(class_dir, i))
  dat$sample = sub("classification_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}

# join with the metrics with perclass metrics 
metrics = do.call(rbind, file_list)
perclass_metrics = fread("data/processed/scores_training/final_classification_metrics.csv")
colnames(perclass_metrics)[11] = "group"
perclass_metrics$sample = sub("classification_spe_list_(.*)_cellchat\\.csv$", "\\1", perclass_metrics$dataset)
perclass_metrics = perclass_metrics[,-c(9,10,12)]
metrics = metrics[perclass_metrics, on = .(sample, group)]
metrics_further = metrics


# Regression metrics 
cyt_reg_dir <- "data/processed/scores_training/cyt_metrics"
nuc_reg_dir <- "data/processed/scores_training/nuc_metrics"
files = list.files(cyt_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(cyt_reg_dir, i))
  dat$sample = sub("cyt_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
cyt_metrics = do.call(rbind, file_list)
colnames(cyt_metrics)[1:3] = paste0("cyt_",colnames(cyt_metrics)[1:3])

files = list.files(nuc_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(nuc_reg_dir, i))
  dat$sample = sub("nuc_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
nuc_metrics = do.call(rbind, file_list)
colnames(nuc_metrics)[1:3] = paste0("nuc_",colnames(nuc_metrics)[1:3])

# regression_metrics = cyt_metrics[nuc_metrics, on = .(group, sample)]
# regression_metrics$`cyt_R²` = ifelse(regression_metrics$`cyt_R²`<0, 0, regression_metrics$`cyt_R²`)
# regression_metrics$`nuc_R²` = ifelse(regression_metrics$`nuc_R²`<0, 0, regression_metrics$`nuc_R²`)
regression_metrics_further = regression_metrics

regression_metrics = melt(regression_metrics, id.vars = c('sample', "group"), measure.vars = c(1:3, 6:8))
regression_metrics$facets <- with(regression_metrics,
  ifelse(variable %in% levels(regression_metrics$variable)[c(1,4)], "R²",
  ifelse(variable %in% levels(regression_metrics$variable)[c(3,6)], "NRMSE", 'RMSE'))
)

regression_metrics$facets = factor(regression_metrics$facets, levels = c("R²", "NRMSE", 'RMSE'))



metrics = metrics_further[regression_metrics_further, on = .(group, sample)]
combo_metrics = metrics[, c(4,14,15,18,17,20)]
combo_metrics[ , geometric_mean := abs(roc_auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]


dt = melt(combo_metrics)

dt$variable <- factor(dt$variable,
  levels = c("roc_auc","macro_recall","cyt_R²","nuc_R²",
             # "cyt_RMSE","nuc_RMSE",
             "cyt_NRMSE","nuc_NRMSE","geometric_mean"),
  labels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "Composite metric")
)
colors = colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )
ggplot(dt, aes(x = variable, y = value)) +
  geom_violin(width = 1, aes(fill = variable)) + 
  scale_fill_manual(values = colors) + 
  geom_boxplot(color = "black", fill = "white", width = 0.2) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = ""
    ) +
  labs( x = "", y = "Value") + coord_cartesian(ylim=c(0,1)) 
ggsave("data/results/Final_Figures/Final_figure3/combo_metrics.pdf", height = 10, width = 8)

```

