---
title: "Figure 4"
author: "rshreej"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-expand: true
    number-sections: true
    theme: united
    highlight: tango
    code-fold: true
    code-tools: true
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| message: false
#| include: false
#| warning: false
library(data.table)
library(circlize)
library(ggvenn)
library(cowplot)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(fgsea)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot) 
library(patchwork)
library(ggbiplot)
library(gridGraphics)
library(gridExtra)
library(igraph)
library(RColorBrewer)
```

```{r}
#| message: false
#| include: true
#| warning: false
theme_set(
  theme_classic(base_size = 14) +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
    )
)
```

# Figure Caption

Figure 4: Training the CCI deconvolution model without including the
spatial component as a feature to test applicability to single cell
data.

A, B Classification and regression metrics on training the model without
the spatial feature. (A) Performance metrics (B) ROC curve ; left,
violin plots representing the distribution of the metrics across the
datasets; right, barplots representing the mean across datasets with
standard error bars.

C. Distribution of the pairwise difference between the model trainings
with and without the spatial information on different combinations of
datasets given a candidate dataset.

D. Heatmap showing the difference in the communication scores when the
model is applied on a single cell data.

### A, B Classification and regression metrics on training the model without the spatial feature. (A) Performance metrics (B) ROC curve ; left, violin plots representing the distribution of the metrics across the datasets; right, barplots representing the mean across datasets with standard error bars. 

```{r}
#| message: false
#| include: true
#| warning: false
#| label: roc_curve_metrics_non_spatial
# AUC curve 
ddir = "data/processed/single_cell/results_P1spatial_removed/"
roc_table = fread(file.path(ddir, "roc_tables_P1_trained.csv")) # get roc tables for the output from run_logo trained without spatial information
xgb_voting = fread(file.path(ddir, "xgboost_voting_metrics_P1_trained.csv"))# processed using extract_metrics in utils.py for the output from run_logo trained without spatial information
roc_table = inner_join(roc_table, xgb_voting[,.(Dataset, auc)], by = 'Dataset')
roc_table$Dataset <- factor(roc_table$Dataset,
  levels = unique(roc_table$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

roc_table$dataset_label <- paste0(roc_table$Dataset, " (AUC=", round(roc_table$auc, 2), ")")

colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )
ggplot(roc_table, aes(x = `False Positive Rate`, y = `True Positive Rate`, color = dataset_label)) +
  geom_line(size = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "False Positive Rate",
       y = "True Positive Rate") +
  scale_color_manual(values = colors) +
  theme_classic(base_size = 20) +
  theme(legend.position = "right") +
  guides(color = guide_legend(title = "Dataset"))

ggsave("data/results/Final_Figures/Final_figure4/roc_table.pdf", width =10)


xgb_voting = xgb_voting[,c(1,12,15,16,18,19,21)]
xgb_voting[ , geometric_mean := (auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]

dt = melt(xgb_voting, id.vars= c('Dataset')) 
dt$variable <- factor(dt$variable,
  levels = c("auc","macro_recall","cyt_R²","nuc_R²",
             # "cyt_RMSE","nuc_RMSE",
             "cyt_NRMSE","nuc_NRMSE","geometric_mean"),
  labels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "Composite metric")
)

#  metrics 
p1 = ggplot(dt, aes(x = variable, y = value, fill = variable, color = variable)) +
  geom_violin(trim = TRUE) +
  geom_boxplot(width = 0.15, color = "black", fill = "white", outlier.size = 0.8) +
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  labs(
    x = NULL,
    y = 'Value'
  ) +
  theme_classic(base_size = 20) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1,vjust = 1)
) + coord_cartesian(ylim = c(0,1))

p2 = ggplot(dt, aes(x = variable, y = value)) +
  stat_summary(fun = mean, geom = "bar", width = 0.7, aes(fill = variable)) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
               geom = "errorbar", width = 0.2, linewidth = 0.6) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.8, color = "black") +
  scale_fill_manual(values = colors)+ 
  labs(
    x = "", y = ""
  ) +
  theme_classic(base_size = 20) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) + coord_cartesian(ylim = c(0,1))

ggarrange(p1,p2, align = "h")

ggsave("data/results/Final_Figures/Final_figure4/combined_no_spatial.pdf", width = 12)
```

### C. Distribution of the pairwise difference between the model trainings with and without the spatial information on different combinations of datasets given a candidate dataset. 

```{r}
#| message: false
#| include: true
#| warning: false
#| label: diff_composite

metrics = fread("data/processed/metrics_all_combinations_trainings_10032025.csv") # run_combinations with and without spatial information and extract_metrics from utils.py
metrics[ , geometric_mean := (roc_auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]

df_long <- metrics[,c(9,28,10,11, 12)]

spatial = df_long[condition == "with_spatial",]
no_spatial = df_long[condition == "without_spatial",]
#check
identical(spatial[,3:5], no_spatial[,3:5])

df_long = spatial[,3:5]
df_long$difference = spatial$geometric_mean - no_spatial$geometric_mean
df_long = df_long[test_name != "spe_list_lung_cellchat",]
df_long = df_long[test_name != "spe_list_brainad_cellchat",]

df_long = na.omit(df_long)

df_long$test_name <- factor(df_long$test_name,
  levels = unique(df_long$test_name),
  labels = c(
             "Brain (GL)",
             "Breast Cancer",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

colors_box = c("#d4dda6", "#78c2e3", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )


ggplot(df_long, aes(x = test_name, y = difference, fill = test_name)) +
  geom_boxplot(outliers = TRUE) +
  scale_fill_manual(values = colors_box) + 
  # facet_wrap(~metric, scales = "free_y", ncol = 4) +
  labs( x = "", y = "Difference in Composite metric") + 
  theme_classic(base_size = 20) +
  theme(
    axis.text.x = element_text(angle = 45, size = 16, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "",
    strip.text = element_text(face = "bold", size = 16, hjust = 0.5),
axis.text.y = element_text(size = 16)
  )
ggsave("data/results/Final_Figures/Final_figure4/boxplot_difference.pdf", width =8)


```

### D. Heatmap showing the difference in the communication scores when the
model is applied on a single cell data.

```{r}
#| message: false
#| include: true
#| warning: false
#| label: single_cell
# load output from experiments/run_singlecell.py
cytoplasm = fread("data/results/predictions/scatter_data_cyt_single_cell.csv") 
nucleus = fread("data/results/predictions/scatter_data_nuc_single_cell.csv")
count_function = function(data){
  shift_val = min(0, min(data$pred))
  data$score = data$pred - shift_val
  # compute the network plot
  links = data.table(
    source= data$source,
    target= data$target,
    importance= data$score
    )
  edges_count <- links[, .(count = .N), by = .(source, target)]
  edges_sum   <- links[, .(sum = sum(importance, na.rm = TRUE)), 
                     by = .(source, target)]
  nodes <- unique(c(links$source, links$target))
  all_pairs <- CJ(source = nodes, target = nodes, unique=TRUE)
  matrix <- merge(all_pairs, edges_count, by=c("source","target"), all.x=TRUE)
  matrix <- merge(matrix, edges_sum,   by=c("source","target"), all.x=TRUE)
  matrix[is.na(count), count := 0]
  matrix[is.na(sum),   sum   := 0]
}

data = cytoplasm
counts_cyt = count_function(data)

data = nucleus
counts_nuc = count_function(data)

diff_counts <- merge(counts_cyt, counts_nuc, by = c("source","target"), suffixes = c("_cyto","_nuc")) %>%
  mutate(diff = sum_cyto - sum_nuc)

ggplot(diff_counts, aes(x = source, y = target, fill = diff)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "#366785", mid = "#cccccc", high = "#c85200", midpoint = 0,  name = "Interaction score \n difference \n (cytoplasm - nucleus)") +
  theme_minimal(base_size = 22) +
  theme(
    axis.text.x = element_text(angle = 90, size = 20, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    strip.text = element_text(face = "bold", size = 20, hjust = 0.5),
axis.text.y = element_text(size = 20)
  ) + labs(x = "Sender", y = "Receiver")

ggsave("data/results/Final_Figures/Final_figure4/scnetworks_heatmap.pdf", width = 12)
```
