---
title: "Figure 1"
author: "rshreej"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-expand: true
    number-sections: true
    theme: united
    highlight: tango
    code-fold: true
    code-tools: true
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| message: false
#| include: false
#| warning: false
library(data.table)
library(circlize)
library(scales)
library(ggvenn)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(fgsea)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot) 
library(patchwork)
library(ggbiplot)
library(tidyverse)
```

```{r}
themes = theme(
  text = element_text(family = "Arial", size = 10),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8),
  legend.text = element_text(size = 8),
  legend.title = element_text(size = 9),
  strip.text = element_text(size = 9)
)
```

# Figure Caption

Figure1: Differences in subcellular communication across 9 different
datasets.

A. Overview of subcellular cell cell communication and quantification of
communication scores across regions - schematic in powerpoint.

B. Heatmap of the communication score of significant interactions
occurring between the different regions of the cell.

C. Significant interactions detected using CellChat are different
between the cytoplasm and nucleus across the 9 different Xenium
datasets.

D. Log2 Fold Change differences in the LR communication score detected
between the various regions of the cell.

E. PCA of CellChat-derived communication score components.

### C. Significant interactions detected using CellChat are different between the cytoplasm and nucleus across the 9 different Xenium datasets

```{r}
#| message: false
#| include: true
#| warning: false
#| label: venn-diagram
#| fig-height: 15
#| fig-width: 15
# Load the processed training data with the OHE for the model 4/D (training data of the best selected model)
data <- fread("data/training_data/training_data.csv") 
data$unique_id = paste0(data$lr_pair, "_", data$source, "_", data$target)
venn_data = list()
plot_list = list()
datasets = unique(data$sample)
titles = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
for (i in 1:length(datasets)){
  subset = data[sample == datasets[i],]
  data_list = list(
    Cell = as.vector(subset[cell_pval <0.05 & cell_score >0,unique_id]),
    Cytoplasm = as.vector(subset[cyt_pval <0.05 & cyt_score > 0,unique_id]),
    Nucleus = as.vector(subset[nuc_pval <0.05& nuc_score > 0,unique_id])
  )
  venn_data[[i]] = data_list
  # Better theme
  p = ggvenn(
    data_list,
    fill_color = c("#c8e6a3", "#f37043", "#4965b0"),
    stroke_size = 0,
    fill_alpha = 0.7,
    stroke_color = "white",
    set_name_color = "#a41146",
    set_name_size = 9,
    text_size = 9,
    show_percentage = F,
  ) + 
    coord_fixed() + 
  ggtitle(titles[i]) +
  theme_minimal(base_size = 30) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 30),
    legend.position = "none",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()) +coord_cartesian(clip="off")

  plot_list[[i]] <- p
}
plot_grid(plotlist = plot_list, ncol = 3)

ggsave("data/results/Final_Figures/Final_figure1/venn_diagram.pdf", height = 20, width = 20)
```

### B. Heatmap of the communication score of significant interactions occurring between the different regions of the cell.

```{r}
#| message: false
#| warning: false
#| include: true
#| fig-width: 10
#| fig-height: 10 
#| label: heatmap
data$unique_id = paste0(data$lr_pair, "_", data$source, "_", data$target)
venn_data = list()
for (i in unique(data$sample)){
  subset = data[sample == i,]
  data_list = list(
    cel = subset[cell_pval <0.05 & cell_score >0,.(cell_score, unique_id)],
    cyt = subset[cyt_pval <0.05 & cyt_score > 0,.(cyt_score, unique_id)],
    nuc = subset[nuc_pval <0.05& nuc_score > 0,.(nuc_score,unique_id)]
  )
  venn_data[[i]] = data_list
}

# List all CellChat objects (.rds) for the 9 datasets used to calculate
# communication scores: cytoplasm → nucleus and nucleus → cytoplasm.
# Replace this path with your own directory if needed.
files = list.files('data/processed/cyt_to_nuc')
cyt_to_nuc = list()
nuc_to_cyt = list()
for (i in files){
  filename = gsub(".rds", "", i)
  obj = readRDS(file.path('data/processed/cyt_to_nuc', i))
  scores <- as.data.frame(as.table(obj[["Prob_P2_P3_P4"]]))
  pval <- as.data.frame(as.table(obj[["PvalP2_P3_P4"]]))
  colnames(scores) = c("source", "target", "lr_pair", "cyt_nuc_score")
  colnames(pval) = c("source", "target", "lr_pair", "pval")
  df = setDT(inner_join(scores, pval, by = intersect(colnames(scores), colnames(pval))))
  df$unique_id = paste0(df$lr_pair, "_", df$source, "_", df$target)
  cyt_to_nuc[[filename]] = df[pval<0.05,.(cyt_nuc_score, unique_id)]
  obj = readRDS(file.path('data/processed/nuc_to_cyt', i))
  scores <- as.data.frame(as.table(obj[["Prob_P2_P3_P4"]]))
  pval <- as.data.frame(as.table(obj[["PvalP2_P3_P4"]]))
  colnames(scores) = c("source", "target", "lr_pair", "nuc_cyt_score")
  colnames(pval) = c("source", "target", "lr_pair", "pval")
  df = setDT(inner_join(scores, pval, by = intersect(colnames(scores), colnames(pval))))
  df$unique_id = paste0(df$lr_pair, "_", df$source, "_", df$target)
  nuc_to_cyt[[filename]] = df[pval<0.05,.(nuc_cyt_score, unique_id)]
}

# add cyt-nuc to the venn data
for (i in names(venn_data)){
  venn_data[[i]][['cyt_to_nuc']] = cyt_to_nuc[[i]]
  venn_data[[i]][['nuc_to_cyt']] = nuc_to_cyt[[i]]
}

library(data.table)

# Store joined table per sample
sample_level_tables <- list()

for (sample_name in names(venn_data)) {
  sample_data <- venn_data[[sample_name]]
  dt_list <- list()

  for (group_name in names(sample_data)) {
    dt <- sample_data[[group_name]]
    
    if (is.data.table(dt)) {
      dt_copy <- copy(dt)
      
      # Rename the score column to group name (e.g., "cel", "cyt", ...)
      score_col <- setdiff(names(dt_copy), "unique_id")
      setnames(dt_copy, score_col, group_name)

      dt_list[[group_name]] <- dt_copy
    }
  }

  # Full outer join of all groups within this sample
  sample_dt <- Reduce(function(x, y) merge(x, y, by = "unique_id", all = TRUE), dt_list)
  sample_level_tables[[sample_name]] <- sample_dt
}


sample_level_tables <- list()

for (sample_name in names(venn_data)) {
  sample_data <- venn_data[[sample_name]]
  dt_list <- list()

  for (group_name in names(sample_data)) {
    dt <- sample_data[[group_name]]
    
    if (is.data.table(dt)) {
      dt_copy <- copy(dt)
      
      # Rename the score column to group name (e.g., "cel", "cyt", ...)
      score_col <- setdiff(names(dt_copy), "unique_id")
      setnames(dt_copy, score_col, group_name)

      dt_list[[group_name]] <- dt_copy
    }
  }

  # Full outer join of all groups within this sample
  sample_dt <- Reduce(function(x, y) merge(x, y, by = "unique_id", all = TRUE), dt_list)
  sample_level_tables[[sample_name]] <- sample_dt
}

heatmap_df <- Reduce(rbind, sample_level_tables)
heatmap_df[is.na(heatmap_df)] <- 0

# conditions for annotation
heatmap_df <- heatmap_df %>%
  mutate(compartment = case_when(
    cel > 0 & cyt > 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "common",

    cel > 0 & cyt > 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cell; cytoplasm; nucleus; cytoplasm to nucleus",
    cel > 0 & cyt > 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cell; cytoplasm; nucleus; nucleus to cytoplasm",
    cel > 0 & cyt > 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cell; cytoplasm; nucleus",

    cel > 0 & cyt > 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cell; cytoplasm; cytoplasm to nucleus; nucleus to cytoplasm",
    cel > 0 & cyt > 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cell; cytoplasm; cytoplasm to nucleus",
    cel > 0 & cyt > 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cell; cytoplasm; nucleus to cytoplasm",
    cel > 0 & cyt > 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cell; cytoplasm",

    cel > 0 & cyt == 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cell; nucleus; cytoplasm to nucleus; nucleus to cytoplasm",
    cel > 0 & cyt == 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cell; nucleus; cytoplasm to nucleus",
    cel > 0 & cyt == 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cell; nucleus; nucleus to cytoplasm",
    cel > 0 & cyt == 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cell; nucleus",

    cel > 0 & cyt == 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cell; cytoplasm to nucleus; nucleus to cytoplasm",
    cel > 0 & cyt == 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cell; cytoplasm to nucleus",
    cel > 0 & cyt == 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cell; nucleus to cytoplasm",
    cel > 0 & cyt == 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cell",

    cel == 0 & cyt > 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cytoplasm; nucleus;cytoplasm to nucleus; nucleus to cytoplasm",
    cel == 0 & cyt > 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cytoplasm; nucleus ; cytoplasm to nucleus",
    cel == 0 & cyt > 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cytoplasm; nucleus; nucleus to cytoplasm",
    cel == 0 & cyt > 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cytoplasm; nucleus",

    cel == 0 & cyt > 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cytoplasm; cytoplasm to nucleus; nucleus to cytoplasm",
    cel == 0 & cyt > 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cytoplasm; cytoplasm to nucleus",
    cel == 0 & cyt > 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cytoplasm; nucleus to cytoplasm",
    cel == 0 & cyt > 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cytoplasm",

    cel == 0 & cyt == 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "nucleus; cytoplasm to nucleus; nucleus to cytoplasm",
    cel == 0 & cyt == 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "nucleus; cytoplasm to nucleus",
    cel == 0 & cyt == 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "nucleus; nucleus to cytoplasm",
    cel == 0 & cyt == 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "nucleus",

    cel == 0 & cyt == 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cytoplasm to nucleus;nucleus to cytoplasm",
    cel == 0 & cyt == 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cytoplasm to nucleus",
    cel == 0 & cyt == 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "nucleus to cytoplasm",

    cel == 0 & cyt == 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "none"
  ))

heatmap_matrix <- as.matrix(heatmap_df[,c(-1,-7)])
annotation_row <- data.frame(Compartment = heatmap_df$compartment)
log_mat <- log10(heatmap_matrix+1e-10)

annotation_row <- data.frame(Compartment = heatmap_df$compartment)
rownames(log_mat) = rownames(annotation_row)
color <- brewer.pal(9, "Reds")
library(RColorBrewer)
palette1 <- brewer.pal(12, "Paired")
palette2 <- brewer.pal(8, "Dark2")
palette3 <- brewer.pal(8, "Set2")
palette4 <- brewer.pal(8, "Accent")
all_colors <- c(palette1, palette2, palette3, palette4)
all_colors <- unique(all_colors)[1:32]
compartments <- unique(annotation_row$Compartment)
compartment_colors <- setNames(all_colors[1:length(compartments)], compartments)
ann_colors <- list(Compartment = compartment_colors)
colnames(log_mat) = c("cell", "Cytoplasm", "Nucleus", "Cytoplasm to Nucleus", "Nucleus to Cytoplasm")
pdf("data/results/Final_Figures/Final_figure1/heatmap.pdf", width=10, height=10)
pheatmap(
  log_mat,
  annotation_row = annotation_row,
  annotation_colors = ann_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  annotation_legend = FALSE,
  legend = FALSE,
  fontsize = 20,
  annotation_names_row = FALSE,
  cellheight = 0.1,
  # title = "Communication score (log 10)",
  color = colorRampPalette(color)(100)[14:100]
)
dev.off()

# get the annotation legend separately 
pdf("data/results/Final_Figures/Final_figure1/heatmap_annotation.pdf", width=5, height=5)
lgd = Legend(labels = names(ann_colors$Compartment), legend_gp = gpar(fill = ann_colors$Compartment, fontsize = 14),
       title = "Region", nrow = 23)
grid.newpage()
draw(lgd)
dev.off()

pdf("data/results/Final_Figures/Final_figure1/heatmap_legend.pdf", width=5, height=3)
col_fun = colorRamp2(
  c(-2, -6, -10),                      
  c("#67000D", "#F5553C", "#FDDECF")     
)

lgd = Legend(
  at = c(-2, -6, -10),
  labels = c(expression(10^-2), expression(10^-6), expression(10^-10)),
  col_fun = col_fun,
  title = "Communication score",
  direction = "horizontal"
)
grid.newpage()
draw(lgd)
dev.off()

```



### D. Log2 Fold Change differences in the LR communication score detected
between the various regions of the cell.

```{r}
#| message: false
#| warning: false
#| include: true
#| label: log2foldchange difference
data$color = case_when(
  data$cell_score != 0 & data$nuc_score == 0 & data$cyt_score == 0 ~ "cell",
  data$cell_score != 0 & data$nuc_score == 0 & data$cyt_score != 0 ~ "cell; cytoplasm",
  data$cell_score != 0 & data$nuc_score != 0 & data$cyt_score == 0 ~ "cell; nucleus",
  data$cell_score != 0 & data$nuc_score != 0 & data$cyt_score != 0 ~ "cell; cytoplasm; nucleus",
  .default = "NS"
)
  
data$color = factor(data$color, levels = c(
  "cell",
  "cell; cytoplasm",
  "cell; nucleus",
  "cell; cytoplasm; nucleus"
))
# cyt_score - nuc_score
data$diff =  log2(data$cyt_score + 1e-6) -  log2(data$nuc_score + 1e-6)


p = ggplot(data, aes(x = cell_score, y = diff, color = color)) +
  geom_point(size = 1.8) +
  scale_x_log10(
    breaks = c(1e-6, 1e-4, 1e-2),
    labels = scales::trans_format("log10", math_format(10^.x))
  ) +
  labs(
    x = "Interaction score in the cell (log scale)",
    y = "Log2FC of interaction score \n (Cytoplasm - Nucleus)",
    color = "Localisation"
  ) +
  scale_color_manual(values = c("#a31044", "#f69d5d", "#c8e7a5", "#4a66ae"),
                     guide = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_classic(base_size = 18) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 30),
    legend.title = element_text(size = 30),
    axis.text.x = element_text(size = 22, angle = 0, hjust = 1),
    axis.text.y = element_text(size = 22),
    axis.title = element_text(size = 30),
    axis.line = element_line(size = 0.8)
  ) + coord_cartesian(xlim = c(1e-6, 1e-2))

ggsave('data/results/Final_Figures/Final_figure1/ma_plot.pdf', width = 12, height = 8)
  
```

### E. PCA of CellChat-derived communication score components.

```{r}
#| message: false
#| warning: false
#| include: true
#| label: PCA plot of lung cancer
# calculate the PCA formula components of cellchat to identify if the components influence separation
# Load the training data with the interaction LR pairs and the processed cell chat of cell cytoplasm and nucleus eg: here is Dataset E in paper
lr_df <- read_csv("data/training_data/training_data.csv")
results <- readRDS("data/processed/communication_score/Lung_high_res.rds")

# add the formula parts to the training_data df to perform PCA 
formula_parts = lapply(results, function (data) {
  P1_vector = c()
  P2_vector = c()
  P3_vector = c()
  P4_vector = c()
  P.spatial_vector = c()
  for (i in 1:nrow(lr_df)){
  P1_vector = append(P1_vector, data@net[["P1"]][lr_df$source[i], lr_df$target[i], lr_df$interaction_name[i]])
  P2_vector = append(P2_vector, data@net[["P2"]][lr_df$source[i], lr_df$target[i], lr_df$interaction_name[i]])
  P3_vector = append(P3_vector, data@net[["P3"]][lr_df$source[i], lr_df$target[i], lr_df$interaction_name[i]])
  P4_vector = append(P4_vector, data@net[["P4"]][lr_df$source[i], lr_df$target[i], lr_df$interaction_name[i]])
  P.spatial_vector = append(P.spatial_vector, data@net[["P.spatial"]][lr_df$source[i], lr_df$target[i], lr_df$interaction_name[i]])
  }
  formula_list = list(P1_vector, P2_vector, P3_vector, P4_vector, P.spatial_vector)
  return(formula_list)
})

names(formula_parts) = c("cell", "cytoplasm", "nucleus")

formula_df = as.data.frame(cbind(do.call(cbind, formula_parts[[1]]), do.call(cbind, formula_parts[[2]]), do.call(cbind, formula_parts[[3]])))
colnames(formula_df) = unlist(lapply(list("cell", "cytoplasm", "nucleus"), function(i){c(paste0(i,"_",'P1_formula'),paste0(i,"_",'P2_formula'),                                                                            paste0(i,"_",'P3_formula'),                                                                            paste0(i,"_",'P4_formula'),                                                                            paste0(i,"_",'P.spatial_formula')
)}))
# remove columns with constant values and other unnecessary columns. The columns retained are
#  "cell_P1_formula"             "cell_P3_formula"             "cell_P.spatial_formula"      "cytoplasm_P1_formula" "cytoplasm_P2_formula"       "cytoplasm_P3_formula"        "cytoplasm_P4_formula"        "cytoplasm_P.spatial_formula" "nucleus_P1_formula"          "nucleus_P2_formula"         "nucleus_P3_formula"          "nucleus_P4_formula"          "nucleus_P.spatial_formula"    "LR-pair"  

formula_df = formula_df[,-c(2,4)]
# retain unique interactions
formula_df$venn_relation = lr_df$venn_relation
formula_df$interaction = lr_df$interaction
formula_df = formula_df[!duplicated(formula_df$interaction),]
pc = prcomp(formula_df[,-c(14,15)],
                   center = TRUE,
                  scale = TRUE)

 formula_df$venn_relation = case_when(
  formula_df$venn_relation == 'cyt' ~ "cytoplasm",
  formula_df$venn_relation == 'nuc' ~ "nucleus",
  formula_df$venn_relation == 'cell + cyt' ~ "cell; cytoplasm",
  formula_df$venn_relation == 'cyt+ nuc' ~ "cytoplasm; nucleus",
  formula_df$venn_relation == 'cell + nuc' ~ "cell; nucleus",
  formula_df$venn_relation == 'cell + cyt + nuc' ~ "cell; cytoplasm; nucleus",
  .default = "cell"
)

# Are the different venn relations clustering? 
pc_df = as.data.frame(pc[["x"]])
pc_df$Compartment = formula_df$venn_relation
pc_df$cell_expression = ifelse(formula_df$cell_P1_formula == 0, 1e-10, formula_df$cell_P1_formula)
pc_df$cyt_expression = ifelse(formula_df$cytoplasm_P1_formula == 0, 1e-10, formula_df$cytoplasm_P1_formula)
pc_df$nuc_expression = ifelse(formula_df$nucleus_P1_formula == 0, 1e-10, formula_df$nucleus_P1_formula)
write_csv(pc_df, "data/processed/pcs_df.csv")
ggplot(pc_df, aes(PC1, PC2, color = Compartment)) + 
  geom_point(size = 0.5) +
  theme_classic() +
  geom_polygon(stat = "ellipse", aes(fill = Compartment), alpha = 0.05) +
  ylim(y = c(-3, 10)) + 
  xlim(-10,10) +
  scale_color_manual(values = c("#a31044", "#f69d5d", "#c8e7a5", "#4a66ae","#db5142","#70aca8","#ebf29f"),
                     guide = guide_legend(nrow = 3, byrow = TRUE)) + 
  scale_fill_manual(values = c("#a31044", "#f69d5d", "#c8e7a5", "#4a66ae","#db5142","#70aca8","#ebf29f")) +
  labs(x = "PC1 (Explained variance: 43.7%)",
        y = "PC2 (Explained variance: 20%)") +
  theme_classic(base_size = 18) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 30),
    legend.title = element_text(size = 30),
    axis.text.x = element_text(size = 22, angle = 0, hjust = 1),
    axis.text.y = element_text(size = 22),
    axis.title = element_text(size = 30)
  )
ggsave("data/results/Final_Figures/Final_figure1/pca_plot.pdf", width = 12, height = 8)
```
