---
title: "Figure 2"
author: "rshreej"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-expand: true
    number-sections: true
    theme: united
    highlight: tango
    code-fold: true
    code-tools: true
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| message: false
#| include: false
#| warning: false
library(data.table)
library(circlize)
library(ggvenn)
library(cowplot)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(fgsea)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot) 
library(patchwork)
library(ggbiplot)
```
# Training a hierarchical model involving a classifier and regressor to predict the communication scores of the subcellular regions of the cell


## Overview of the hierarchical model involving a classifier and regressor for predicting the subcellular cell communication scores 
```{r}
# adding the schematic here
```

```{r}
#| message: false
#| include: true
#| warning: false
theme_set(
  theme_classic(base_size = 14) +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
    )
)
```

## Performance metrics of different base learners involving tree models. Model 1: XgBoost classifier and regressor; Model 2: Random Forest Classifier and Regressor; Model 3: Random Forest Classifier and XgBoost Regressor; Model 4: Voting (Random Forest + XgBoost) Classifier and XgBoost Regressor
```{r}
#| message: false
#| include: true
#| warning: false
ddir = "data/processed/plots_metrics"
xgb = fread(file.path(ddir, "comparison_baselearners/xgboost_metrics.csv"))
rf = fread(file.path(ddir, "comparison_baselearners/rf_metrics.csv"))
xgb_rf = fread(file.path(ddir, "comparison_baselearners/xgboost_rf_metrics.csv"))
xgb_voting = fread(file.path(ddir, "comparison_baselearners/xgboost_voting_metrics.csv"))

xgb = xgb[, model:= "xgb"]
rf = rf[, model:= "rf"]
xgb_rf = xgb_rf[, model:= "xgb_rf"]
xgb_voting = xgb_voting[, model:= "xgb_voting"]

dt = rbind(xgb, rf, xgb_rf, xgb_voting)
dt[ , geometric_mean := (auc * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]

heatmap_df = melt(dt, id.vars= c('model', 'Dataset'), measure.vars= 23) 

# mapping
heatmap_df$Dataset <- factor(heatmap_df$Dataset,
  levels = unique(heatmap_df$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

heatmap_df$model <- factor(heatmap_df$model,
  levels = unique(heatmap_df$model),
  labels = c("Model 1", "Model 2", "Model 3", "Model 4")
)

colors = c("#eea3b5", "#c58fb0", "#9a7ea3", "#6f6d8f")

# plot the heatmap
p1 = ggplot(heatmap_df, aes(x = model, y = value, fill = model), color = ) +
  geom_boxplot() + 
  scale_fill_manual (values = colors) + 
  labs(x = "", y = "Composite metric") +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    legend.position = "",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  )

# ggplot(heatmap_df, aes(x = model, y = value)) +
#   stat_summary(fun = mean, geom = "bar", width = 0.7, aes(fill = model)) +
#   stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
#                geom = "errorbar", width = 0.2, linewidth = 0.6) +
#   geom_jitter(width = 0.15, size = 2, alpha = 0.8, color = "black") +
#   scale_fill_manual(values = colors)+ 
#   labs(
#     x = "", y = ""
#   ) +
#   theme_classic(base_size = 20) +
#   theme(legend.position = "none",
#         axis.text.x = element_text(angle = 90, hjust = 1))

```

## Performance metrics of feature engineering; Each models had different methods of encoding the location features
```{r}
#| message: false
#| include: true
#| warning: false
ddir = "data/processed/plots_metrics"
a = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_ohe_not_separated_model2.csv"))
b = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_catboost_not_separated_model3.csv"))
c = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_combined_model4.csv"))
d = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_model1.csv"))

a = a[, model:= "Model A"]
b = b[, model:= "Model B"]
c = c[, model:= "Model C"]
d = d[, model:= "Model D"]

dt = rbind(a, b, c, d)
dt[ , geometric_mean := (auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]


heatmap_df = melt(dt, id.vars= c('model', 'Dataset'), measure.vars= 23) 

# mapping
heatmap_df$Dataset <- factor(heatmap_df$Dataset,
  levels = unique(heatmap_df$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

# plot the heatmap
p2 = ggplot(heatmap_df, aes(x = model, y = value, fill = model), color = ) +
  geom_boxplot() + 
  scale_fill_manual (values = colors) + 
  labs(x = "", y = "Composite metric") +
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    legend.position = "",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  )
ggarrange(p1,p2, align = "h")

ggsave("data/results/Final_Figures/Final_figure2/heatmap_featureengineering.pdf")
```

## Performance metrics of the model. Training was done with LOGO CV on 9 different 10X Xenium datasets

```{r}
#| message: false
#| include: true
#| warning: false
# AUC curve 
roc_table = fread("data/processed/roc_tables.csv")
xgb_voting = fread(file.path(ddir, "comparison_baselearners/xgboost_voting_metrics.csv"))
roc_table = inner_join(roc_table, xgb_voting[,.(Dataset, auc)], by = 'Dataset')
roc_table$Dataset <- factor(roc_table$Dataset,
  levels = unique(roc_table$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

roc_table$dataset_label <- paste0(roc_table$Dataset, " (AUC=", round(roc_table$auc, 2), ")")

colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )
ggplot(roc_table, aes(x = `False Positive Rate`, y = `True Positive Rate`, color = dataset_label)) +
  geom_line(size = 0.3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "False Positive Rate",
       y = "True Positive Rate") +
  scale_color_manual(values = colors) + 
  theme_classic(base_size = 20) +
  theme(legend.position = "right") +
  guides(color = guide_legend(title = "Dataset"))

ggsave("data/results/Final_Figures/Final_figure2/roc_table.pdf", width =10)
# Regression metrics 
xgb_voting = fread(file.path(ddir, "comparison_baselearners/xgboost_voting_metrics.csv"))
xgb_voting = xgb_voting[,c(1,12,15,16,18,19,21)]
xgb_voting[ , geometric_mean := (auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]

dt = melt(xgb_voting, id.vars= c('Dataset')) 
dt$variable <- factor(dt$variable,
  levels = c("auc","macro_recall","cyt_R²","nuc_R²",
             # "cyt_RMSE","nuc_RMSE",
             "cyt_NRMSE","nuc_NRMSE","geometric_mean"),
  labels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "Composite metric")
)

#  metrics 
p1 = ggplot(dt, aes(x = variable, y = value, fill = variable, color = variable)) +
  geom_violin(trim = TRUE) +
  geom_boxplot(width = 0.15, color = "black", fill = "white", outlier.size = 0.8) +
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  labs(
    x = NULL,
    y = 'Value'
  ) +
  theme_classic(base_size = 20) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1,vjust = 1)
) + coord_cartesian(ylim = c(0,1))

p2 = ggplot(dt, aes(x = variable, y = value)) +
  stat_summary(fun = mean, geom = "bar", width = 0.7, aes(fill = variable)) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
               geom = "errorbar", width = 0.2, linewidth = 0.6) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.8, color = "black") +
  scale_fill_manual(values = colors)+ 
  labs(
    x = "", y = "Value"
  ) +
  theme_classic(base_size = 20) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) + coord_cartesian(ylim = c(0,1))

ggarrange(p1,p2, align = "h")
ggsave("data/results/Final_Figures/Final_figure2/combined_best_model.pdf", width =15, height = 6)


# xgb_voting[ , NRMSE_weighted_auc := sqrt(auc * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) ) ]
# xgb_voting[ , R2_weighted_auc    := sqrt(auc * ((`cyt_R²` + `nuc_R²`) / 2)) ]
# xgb_voting[ , NRMSE_weighted_recall := sqrt(`macro_recall` * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) ) ]
# xgb_voting[ , R2_weighted_recall    := sqrt(`macro_recall` * ((`cyt_R²` + `nuc_R²`) / 2)) ]
# df = melt(xgb_voting, id.vars= c('Dataset'), measure.vars=c("R2_weighted_auc","R2_weighted_recall", "NRMSE_weighted_auc", "NRMSE_weighted_recall")) 
# 
# df$variable <- factor(
#   df$variable,
#   levels = c("R2_weighted_auc", "R2_weighted_recall",
#              "NRMSE_weighted_auc", "NRMSE_weighted_recall"),
#   labels = c("GM(AUC, R²)", 
#              "GM(Recall, R²)",
#              "GM(AUC, 1−NRMSE)", 
#              "GM(Recall, 1−NRMSE)")
# )
# 
# colors = c("#f57e6f", "#d2dba4", "#78c3e7", "#f2a7db")
# p1 = ggplot(df, aes(x = variable, y = value)) +
#   stat_summary(fun = mean, geom = "bar", width = 0.7, aes(fill = variable)) +
#   stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
#                geom = "errorbar", width = 0.2, linewidth = 0.6) +
#   geom_jitter(width = 0.15, size = 2, alpha = 0.8, color = "black") +
#   scale_fill_manual(values = colors)+ 
#   labs(
#     x = "Metric", y = "Score"
#   ) +
#   theme_classic(base_size = 20) +
#   theme(legend.position = "none",
#         axis.text.x = element_text(angle = 90, hjust = 1))
# 
# p2 = ggplot(df, aes(x = variable, y = value)) +
#   geom_violin(aes(alpha = 0.7, fill = variable)) +
#   geom_boxplot(width = 0.3,  color = "black", fill = "white") + #, aes(fill = variable)) +
#   geom_jitter(width = 0.15, size = 2, alpha = 0.8, color = "black") +
#   labs(
#     x = "Metric", y = "Score"
#   ) +
#   scale_fill_manual(values = colors) + 
#   scale_color_manual(values = colors) + 
#   theme_classic(base_size = 20) +
#   theme(legend.position = "none",
#         axis.text.x = element_text(angle = 90, hjust = 1))
# ggarrange(p1, p2, nrow = 1)
# ggsave("data/results/Final_Figures/Final_figure2/combined_best_model.pdf")
```

## geometric mean of the metrics across different combinations with and without the spatial feature in the model
```{r}
metrics = fread("data/processed/metrics_all_combinations_trainings_10032025.csv")
metrics[ , geometric_mean := (roc_auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]

df_long <- metrics[,c(9,28,12)]

df_summary <- df_long %>%
  group_by(combo_size, condition) %>%
  summarise(mean_val = mean(geometric_mean, na.rm = TRUE),
            sd_val = sd(geometric_mean, na.rm = TRUE)/sqrt(n()), .groups = "drop")

colors = c('with_spatial' = "#b715b7", "without_spatial" = "#14bf81")
ggplot(df_summary, aes(x = combo_size, y = mean_val, color = condition, group = condition)) +
  geom_line() +
  geom_point() +
  scale_color_manual(
  name   = "Model type",
  values = colors,
  labels = c(
    with_spatial    = "With spatial information",
    without_spatial = "Without spatial information"
  ),
  guide = guide_legend(nrow = 2, byrow = TRUE)
) + 
  geom_errorbar(aes(ymin = mean_val - sd_val, ymax = mean_val + sd_val), width = 0.2) +
  # facet_wrap(~metric, scales = "free_y", ncol = 4) +
  labs( x = "Number of training datasets", y = "Average of Composite metric") + 
  theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  )

ggsave("data/results/Final_Figures/Final_figure2/trainingcombinations.pdf", height = 4, width = 4.5)
```


