---
title: "Figure 2"
author: "rshreej"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-expand: true
    number-sections: true
    theme: united
    highlight: tango
    code-fold: true
    code-tools: true
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| message: false
#| include: false
#| warning: false
library(data.table)
library(tidyverse)
library(circlize)
library(ggvenn)
library(cowplot)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(fgsea)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot) 
library(patchwork)
library(ggbiplot)
library(reshape2)
library(ThemePark)
```


```{r}
theme_set(
  theme_classic(base_size = 22) +
    theme(
      axis.text = element_text(size = 20),
      axis.title = element_text(size = 22),
      legend.text = element_text(size = 22),
      legend.title = element_text(size = 22),
      plot.title = element_text(size = 22, hjust = 0.5)
    )
)
```

# Training the lung cancer dataset with 255 varying combinations of the 8 other datasets to check if the dataset choice affects the model stability

## Number of LR pairs correctly classified when training the lung cancer dataset on 255 combinations of the 8 training datasets (raw counts on the left, percentages on the right)

```{r}
#| message: false
#| include: true
#| warning: false
dir = "data/processed/scores_training/classification_raw"
files = list.files(dir)[1:9]
file_list = list()

for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(dir, i))
  dat$correct = ifelse(dat$pred == dat$true, 1, 0)
  x = nrow(dat)/255
  dat$id = rep(1:x, 255)
  data = dcast(dat, id ~ group, fun.aggregate = NULL, value.var = "correct")
  data$sample = sub("classification_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = data
}

prediction = setDT(do.call(rbind, file_list))
rownames(prediction) = 1:nrow(prediction)
col = prediction[, lapply(.SD, sum), by = sample, .SDcols = c(2:256)]
percent <- prediction[, lapply(.SD, function(x) sum(x) *100 / .N), 
                      by = sample, .SDcols = 2:256]

counts <- prediction[, .N, by = sample]
counts$sample <- factor(counts$sample,
  levels = unique(counts$sample),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)


colsums = melt(col, id.vars = "sample")
percent = melt(percent, id.vars = "sample")


percent$sample <- factor(percent$sample,
  levels = unique(percent$sample),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

p0 = ggplot(percent, aes(x = sample, y = value)) +
  # geom_violin(fill = "#ccebc5", width = 2) + 
  geom_boxplot(fill = "#ccebc5", width = 0.5, outlier.size = 0.5) + 
  # geom_jitter(width = 0.1, alpha = 0.3, size = 1) + 
  geom_text(
    data = counts,
    aes(x = sample, y = max(percent$value)-10, label = paste0("N=", N)),
    inherit.aes = FALSE,
    color = "#B43757",
    size = 5,
    angle = 90,
    hjust = 0,
    vjust =0
  ) + 
  labs(x = "", y = "% LR pairs classified correctly")  + 
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 18),
      legend.text = element_text(size = 18),
      legend.title = element_text(size = 18),
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.text.x = element_text(angle = 45,  hjust = 1, vjust = 1)
    )


# setDT(percent)
# percent <- percent[, .(mean_value = mean(value)), by = sample]

p1 <- ggplot(percent, aes(x = sample, y = value)) +
  stat_summary(fun = mean, geom = "bar", width = 0.7, fill = "#bebadb") +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
               geom = "errorbar", width = 0.2, linewidth = 0.6) +
  # geom_jitter(width = 0.001, size = 2, alpha = 0.8, color = "black")  +
  labs(x = "", y = "Mean % of correct classifications \n across the different combinations") +
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 18),
      legend.text = element_text(size = 18),
      legend.title = element_text(size = 18),
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.text.x = element_text(angle = 45,  hjust = 1, vjust = 1)
    )

ggarrange (p1, p0, common.legend = F)
ggsave("data/results/Final_Figures/Final_figure3/counts_plot.pdf", width = 8)
```

Note: 
Barplot = “How good is each dataset on average?” Stability 
Histogram = “How are individual LR pairs spread out in terms of classification performance?”

## Number of times each LR pair was classified correctly 
```{r}
#| message: false
#| include: true
#| warning: false

# id is not unique
row <- prediction[, .(value = rowSums(.SD)), .SDcols = 2:256]
row$percent <- row$value * 100 / 255
row$sample = prediction$sample  
p0 = ggplot(row, aes(x = value)) +
  geom_histogram(aes(y = after_stat(count / sum(count))),
                 binwidth = 5, fill = "#fb8073", color = "#fb8073") +
  scale_y_continuous(labels = scales::percent) +
  # scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  labs(x = "Number of Combinations", 
       y = "Proportion (%) of LR pairs classified correctly") +
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 14),
      axis.title = element_text(size = 18),
      legend.text = element_text(size = 18),
      legend.title = element_text(size = 18),
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.text.x = element_text(angle = 45,  hjust = 1, vjust = 1)
    )

p1 = ggplot(row, aes(x = "", y = value)) +
  geom_violin(fill = "#ffffb4", width = 0.6) + 
  geom_boxplot(color = "black", fill = "white", width = 0.05) + 
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )  + 
  labs(x = "Correctly classified LR pairs", y = "Number of combinations") +
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 18),
      legend.text = element_text(size = 18),
      legend.title = element_text(size = 18),
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.ticks.x = element_blank()
    )
ggarrange (p0, p1, align = "hv")
ggsave("data/results/Final_Figures/Final_figure3/lr_counts_plot.pdf", width = 8)

row$sample <- factor(row$sample,
  levels = unique(row$sample),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )

ggplot(row, aes(x = sample, y = percent, fill = sample)) +
  geom_violin(width = 0.6) + 
  geom_boxplot(color = "black", fill = "white", width = 0.05) + 
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )  + 
  labs(x = "", y = "Number of correctly \n classified LR pairs") +
  scale_fill_manual(values = colors) + 
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 18),
      legend.text = element_text(size = 18),
      legend.title = element_text(size = 18),
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.text.x = element_text(angle = 45,  hjust = 1, vjust = 1),
      legend.position = "None"
    )
ggsave("data/results/Final_Figures/Supplementary_figures/counts_plots/lr_box_plot.pdf", width = 8)
```

# Easy vs hard analysis
```{r}
#| message: false
#| include: true
#| warning: false

# Easy is LR pairs with greater than 3rd quartile 
# Intermediate is LR pairs between 2nd and 3rd quartile
# Hard is LR pairs less than 1st quartile

prediction = setDT(do.call(rbind, file_list))
rownames(prediction) = 1:nrow(prediction)

# offset the rownames by the number of lr pairs
offsets <- prediction[, .N, by = sample][
  , offset := shift(cumsum(N), fill = 0)]
prediction <- merge(prediction, offsets[, .(sample, offset)], by = "sample")
prediction[, gid := id + offset]
# check 
x = as.integer(rownames(prediction))
y = prediction$gid
identical(x,y)


row <- prediction[, .(value = rowSums(.SD)), by = gid, .SDcols = 2:256]
row$percent <- row$value * 100 / 255
row[, group := fcase(
  percent > 75,  "Easy",
  percent < 25,  "Hard",
  default = "Intermediate"
)]

# join this with the dataframe of the features
training_data = read_csv("data/processed/training_data_071025.csv")
z = as.integer(rownames(training_data))
identical(y,z)
data = cbind(row, training_data[,1:38]) 

prop_df <- data[, .N, by = .(lr_pair, group)]
prop_df[, total := sum(N), by = lr_pair]
prop_df[, prop := N / total]

# Irrespective of cell type how many LR pairs are always Easy/ hard / intermediate
# There are 290 unique LR pairs and most of them are always easy to classify (it means that irrespective of celltype or dataset, that LR pairs have been classified 75% of times correctly)
# count all LR pairs per group
all_counts <- prop_df[, .N, by = group]

# count only "always" LR pairs per group
always = prop_df[prop == 1,]
always_counts <- always[, .N, by = group]
all_counts$group = factor(all_counts$group, levels = c("Easy", "Intermediate", "Hard"))

p1 <- ggplot(all_counts, aes(x = group, y = N, fill = group)) +
  geom_col(alpha = 0.6) +  # base bars: all pairs
  geom_col(data = always_counts, aes(x = group, y = N), 
           fill = "gray", width = 0.4) +  # overlay: always pairs
  labs(x = "Group", 
       y = "Number of LR pairs") + 
  guides(fill = "none") 
ggsave('data/results/Final_Figures/Final_figure3/easy_hard.pdf', height = 5)


hard_df <- prop_df[group == "Hard"]

p2 = ggplot(hard_df, aes(x = reorder(lr_pair, -prop), y = prop)) +
  geom_col(fill = "#85d692") +
  geom_text(aes(label = paste0(N, "/", total)), vjust = 0.5, hjust = 0, size = 6) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Hard LR pair", y = "Proportion of hard calls")

ggsave('data/results/Final_Figures/Final_figure3/hard_prop.pdf', width = 10, height = 12)
```
## Question is how do I check what is APP_CD74. I think the reason for misclassifiction is because there is no prior information and the nucleus score is zero but was there in all the celltpyes in one particular tissue

## Regression and classification metric across the 255 different trainings
```{r}
#| message: false
#| include: true
#| warning: false
class_dir <- "data/processed/scores_training/classification_metrics/"

files = list.files(class_dir)
file_list = list()

for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(class_dir, i))
  dat$sample = sub("classification_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}

# join with the metrics with perclass metrics 
metrics = do.call(rbind, file_list)
perclass_metrics = fread("data/processed/scores_training/final_classification_metrics.csv")
colnames(perclass_metrics)[11] = "group"
perclass_metrics$sample = sub("classification_spe_list_(.*)_cellchat\\.csv$", "\\1", perclass_metrics$dataset)
perclass_metrics = perclass_metrics[,-c(9,10,12)]
metrics = metrics[perclass_metrics, on = .(sample, group)]
metrics_further = metrics


# Regression metrics 
cyt_reg_dir <- "data/processed/scores_training/cyt_metrics"
nuc_reg_dir <- "data/processed/scores_training/nuc_metrics"
files = list.files(cyt_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(cyt_reg_dir, i))
  dat$sample = sub("cyt_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
cyt_metrics = do.call(rbind, file_list)
colnames(cyt_metrics)[1:3] = paste0("cyt_",colnames(cyt_metrics)[1:3])

files = list.files(nuc_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(nuc_reg_dir, i))
  dat$sample = sub("nuc_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
nuc_metrics = do.call(rbind, file_list)
colnames(nuc_metrics)[1:3] = paste0("nuc_",colnames(nuc_metrics)[1:3])

# regression_metrics = cyt_metrics[nuc_metrics, on = .(group, sample)]
# regression_metrics$`cyt_R²` = ifelse(regression_metrics$`cyt_R²`<0, 0, regression_metrics$`cyt_R²`)
# regression_metrics$`nuc_R²` = ifelse(regression_metrics$`nuc_R²`<0, 0, regression_metrics$`nuc_R²`)
regression_metrics_further = regression_metrics

regression_metrics = melt(regression_metrics, id.vars = c('sample', "group"), measure.vars = c(1:3, 6:8))
regression_metrics$facets <- with(regression_metrics,
  ifelse(variable %in% levels(regression_metrics$variable)[c(1,4)], "R²",
  ifelse(variable %in% levels(regression_metrics$variable)[c(3,6)], "NRMSE", 'RMSE'))
)

regression_metrics$facets = factor(regression_metrics$facets, levels = c("R²", "NRMSE", 'RMSE'))



metrics = metrics_further[regression_metrics_further, on = .(group, sample)]
combo_metrics = metrics[, c(4,14,15,18,17,20)]
combo_metrics[ , geometric_mean := abs(roc_auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]


dt = melt(combo_metrics)

dt$variable <- factor(dt$variable,
  levels = c("roc_auc","macro_recall","cyt_R²","nuc_R²",
             # "cyt_RMSE","nuc_RMSE",
             "cyt_NRMSE","nuc_NRMSE","geometric_mean"),
  labels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "Composite metric")
)
colors = colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )
ggplot(dt, aes(x = variable, y = value)) +
  geom_violin(width = 1, aes(fill = variable)) + 
  scale_fill_manual(values = colors) + 
  geom_boxplot(color = "black", fill = "white", width = 0.2) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = ""
    ) +
  labs( x = "", y = "Value") + coord_cartesian(ylim=c(0,1)) 
ggsave("data/results/Final_Figures/Final_figure3/combo_metrics.pdf", height = 10, width = 8)


# AUC split by datasets and the composite metric split by dataset 
metrics_auc = metrics[, .(roc_auc, sample)]
ggplot(metrics_auc, aes(x = sample, y = roc_auc)) +
  geom_violin(width = 1, aes(fill = sample)) + 
  scale_fill_manual(values = colors) + 
  geom_boxplot(color = "black", fill = "white", width = 0.2) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = ""
    ) +
  labs( x = "", y = "Value") + coord_cartesian(ylim=c(0,1))

metrics_auc = metrics_auc %>% group_by(sample) %>% mutate(median = median(roc_auc))

```

## Confusion matrix of the 256 trainings
```{r}
#| message: false
#| include: true
#| warning: false
dir = "data/processed/scores_training/classification_raw/correct"
files = list.files(dir)
file_list = list()

for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(dir, i))
  file_list[[i]]= dat
}

confusion = setDT(do.call(rbind, file_list))

cm <- table(Predicted = confusion$pred, Actual = confusion$true)

# from chatgpt
# confusion[, c("TN","FP","FN","TP") := {
#   # remove [ and ], then trim leading/trailing spaces
#   parts <- trimws(gsub("\\[|\\]", "", cm))
#   
#   # split into numeric vectors
#   vals <- strsplit(parts, "\\s+")
#   
#   # turn into matrix (rows = datasets, cols = TN FP FN TP)
#   m <- t(sapply(vals, function(x) as.numeric(x[x != ""])))
#   
#   # return as list of columns so := works
#   split(m, col(m))
# }]
# 
# confusion_matrix = colSums(confusion[, c("TN","FP","FN","TP")]) 
# 
# m <- matrix(c(confusion_matrix[1], confusion_matrix[2],
#                 confusion_matrix[3], confusion_matrix[4]), nrow = 2, byrow = TRUE)
# colnames(m) <- c("cell", "cytoplasm/nucleus")
# rownames(m) <- c("cell", "cytoplasm/nucleus")
  
dfm <- melt(cm)
ggplot(dfm, aes(x = Predicted, y = Actual, fill = value)) +
  geom_tile(color = "black") +
  geom_text(aes(label = value), size = 5) +
  scale_fill_gradient(low = "white", high = "#bc80bd", guide = "none") +
  labs(x = "Predicted", y = "Truth") +
  theme_minimal(base_size = 14) + scale_y_discrete(limits = rev) +
  theme(panel.grid.major = element_blank())
ggsave("data/results/Final_Figures/Final_figure3/confusion.pdf")
```



