---
title: "Figure 1"
author: "rshreej"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-expand: true
    number-sections: true
    theme: united
    highlight: tango
    code-fold: true
    code-tools: true
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| message: false
#| include: false
#| warning: false
library(data.table)
library(circlize)
library(ggvenn)
library(cowplot)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(fgsea)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot) 
library(patchwork)
library(ggbiplot)
library(tidyverse)
```

```{r}
themes = theme(
  text = element_text(family = "Arial", size = 10),
  axis.title = element_text(size = 9),
  axis.text = element_text(size = 8),
  legend.text = element_text(size = 8),
  legend.title = element_text(size = 9),
  strip.text = element_text(size = 9)
)
```

# Figure 1: Differences in subcellular communication across 9 different datasets 

## Overview of subcellular cell cell communication and quantification of communication scores across regions

```{r}
#| message: false
#| include: true
#| warning: false
#| label: panelA
```

## Significant interactions detected using CellChat are different between the cytoplasm and nucleus across the 9 tested data sets

```{r}
#| message: false
#| include: true
#| warning: false
#| label: panelB 
#| fig-height: 15
#| fig-width: 15
data <- fread("data/processed/training_data_071025.csv")
data$unique_id = paste0(data$lr_pair, "_", data$source, "_", data$target)
venn_data = list()
plot_list = list()
datasets = unique(data$sample)
titles = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
for (i in 1:length(datasets)){
  subset = data[sample == datasets[i],]
  data_list = list(
    Cell = as.vector(subset[cell_pval <0.05 & cell_score >0,unique_id]),
    Cytoplasm = as.vector(subset[cyt_pval <0.05 & cyt_score > 0,unique_id]),
    Nucleus = as.vector(subset[nuc_pval <0.05& nuc_score > 0,unique_id])
  )
  venn_data[[i]] = data_list
  # Better theme
  p = ggvenn(
    data_list,
    fill_color = c("#c8e6a3", "#f37043", "#4965b0"),
    stroke_size = 0,
    fill_alpha = 0.7,
    stroke_color = "white",
    set_name_color = "#a41146",
    set_name_size = 9,
    text_size = 9,
    show_percentage = F,
  ) + 
    coord_fixed() + 
  ggtitle(titles[i]) +
  theme_minimal(base_size = 30) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 30),
    legend.position = "none",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()) +coord_cartesian(clip="off")

  plot_list[[i]] <- p
}
plot_grid(plotlist = plot_list, ncol = 3)

ggsave("data/results/Final_Figures/Final_figure1/venn_diagram.pdf", height = 20, width = 20)
```

##Heatmap of the communication score of significant interactions occurring between the different regions of the cell

```{r}
#| message: false
#| warning: false
#| include: true
#| fig-width: 10
#| fig-height: 10 
data$unique_id = paste0(data$lr_pair, "_", data$source, "_", data$target)
venn_data = list()
for (i in unique(data$sample)){
  subset = data[sample == i,]
  data_list = list(
    cel = subset[cell_pval <0.05 & cell_score >0,.(cell_score, unique_id)],
    cyt = subset[cyt_pval <0.05 & cyt_score > 0,.(cyt_score, unique_id)],
    nuc = subset[nuc_pval <0.05& nuc_score > 0,.(nuc_score,unique_id)]
  )
  venn_data[[i]] = data_list
}

files = list.files('data/processed/cyt_to_nuc')
cyt_to_nuc = list()
nuc_to_cyt = list()
for (i in files){
  filename = gsub(".rds", "", i)
  obj = readRDS(file.path('data/processed/cyt_to_nuc', i))
  scores <- as.data.frame(as.table(obj[["Prob_P2_P3_P4"]]))
  pval <- as.data.frame(as.table(obj[["PvalP2_P3_P4"]]))
  colnames(scores) = c("source", "target", "lr_pair", "cyt_nuc_score")
  colnames(pval) = c("source", "target", "lr_pair", "pval")
  df = setDT(inner_join(scores, pval, by = intersect(colnames(scores), colnames(pval))))
  df$unique_id = paste0(df$lr_pair, "_", df$source, "_", df$target)
  cyt_to_nuc[[filename]] = df[pval<0.05,.(cyt_nuc_score, unique_id)]
  obj = readRDS(file.path('data/processed/nuc_to_cyt', i))
  scores <- as.data.frame(as.table(obj[["Prob_P2_P3_P4"]]))
  pval <- as.data.frame(as.table(obj[["PvalP2_P3_P4"]]))
  colnames(scores) = c("source", "target", "lr_pair", "nuc_cyt_score")
  colnames(pval) = c("source", "target", "lr_pair", "pval")
  df = setDT(inner_join(scores, pval, by = intersect(colnames(scores), colnames(pval))))
  df$unique_id = paste0(df$lr_pair, "_", df$source, "_", df$target)
  nuc_to_cyt[[filename]] = df[pval<0.05,.(nuc_cyt_score, unique_id)]
}

# add cyt-nuc to the venn data
for (i in names(venn_data)){
  venn_data[[i]][['cyt_to_nuc']] = cyt_to_nuc[[i]]
  venn_data[[i]][['nuc_to_cyt']] = nuc_to_cyt[[i]]
}

library(data.table)

# Store joined table per sample
sample_level_tables <- list()

for (sample_name in names(venn_data)) {
  sample_data <- venn_data[[sample_name]]
  dt_list <- list()

  for (group_name in names(sample_data)) {
    dt <- sample_data[[group_name]]
    
    if (is.data.table(dt)) {
      dt_copy <- copy(dt)
      
      # Rename the score column to group name (e.g., "cel", "cyt", ...)
      score_col <- setdiff(names(dt_copy), "unique_id")
      setnames(dt_copy, score_col, group_name)

      dt_list[[group_name]] <- dt_copy
    }
  }

  # Full outer join of all groups within this sample
  sample_dt <- Reduce(function(x, y) merge(x, y, by = "unique_id", all = TRUE), dt_list)
  sample_level_tables[[sample_name]] <- sample_dt
}


sample_level_tables <- list()

for (sample_name in names(venn_data)) {
  sample_data <- venn_data[[sample_name]]
  dt_list <- list()

  for (group_name in names(sample_data)) {
    dt <- sample_data[[group_name]]
    
    if (is.data.table(dt)) {
      dt_copy <- copy(dt)
      
      # Rename the score column to group name (e.g., "cel", "cyt", ...)
      score_col <- setdiff(names(dt_copy), "unique_id")
      setnames(dt_copy, score_col, group_name)

      dt_list[[group_name]] <- dt_copy
    }
  }

  # Full outer join of all groups within this sample
  sample_dt <- Reduce(function(x, y) merge(x, y, by = "unique_id", all = TRUE), dt_list)
  sample_level_tables[[sample_name]] <- sample_dt
}

heatmap_df <- Reduce(rbind, sample_level_tables)
heatmap_df[is.na(heatmap_df)] <- 0

heatmap_df <- heatmap_df %>%
  mutate(compartment = case_when(
    cel > 0 & cyt > 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "common",

    cel > 0 & cyt > 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cell; cytoplasm; nucleus; cytoplasm to nucleus",
    cel > 0 & cyt > 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cell; cytoplasm; nucleus; nucleus to cytoplasm",
    cel > 0 & cyt > 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cell; cytoplasm; nucleus",

    cel > 0 & cyt > 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cell; cytoplasm; cytoplasm to nucleus; nucleus to cytoplasm",
    cel > 0 & cyt > 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cell; cytoplasm; cytoplasm to nucleus",
    cel > 0 & cyt > 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cell; cytoplasm; nucleus to cytoplasm",
    cel > 0 & cyt > 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cell; cytoplasm",

    cel > 0 & cyt == 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cell; nucleus; cytoplasm to nucleus; nucleus to cytoplasm",
    cel > 0 & cyt == 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cell; nucleus; cytoplasm to nucleus",
    cel > 0 & cyt == 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cell; nucleus; nucleus to cytoplasm",
    cel > 0 & cyt == 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cell; nucleus",

    cel > 0 & cyt == 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cell; cytoplasm to nucleus; nucleus to cytoplasm",
    cel > 0 & cyt == 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cell; cytoplasm to nucleus",
    cel > 0 & cyt == 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cell; nucleus to cytoplasm",
    cel > 0 & cyt == 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cell",

    cel == 0 & cyt > 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cytoplasm; nucleus;cytoplasm to nucleus; nucleus to cytoplasm",
    cel == 0 & cyt > 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cytoplasm; nucleus ; cytoplasm to nucleus",
    cel == 0 & cyt > 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cytoplasm; nucleus; nucleus to cytoplasm",
    cel == 0 & cyt > 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cytoplasm; nucleus",

    cel == 0 & cyt > 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cytoplasm; cytoplasm to nucleus; nucleus to cytoplasm",
    cel == 0 & cyt > 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cytoplasm; cytoplasm to nucleus",
    cel == 0 & cyt > 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "cytoplasm; nucleus to cytoplasm",
    cel == 0 & cyt > 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "cytoplasm",

    cel == 0 & cyt == 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "nucleus; cytoplasm to nucleus; nucleus to cytoplasm",
    cel == 0 & cyt == 0 & nuc > 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "nucleus; cytoplasm to nucleus",
    cel == 0 & cyt == 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "nucleus; nucleus to cytoplasm",
    cel == 0 & cyt == 0 & nuc > 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "nucleus",

    cel == 0 & cyt == 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt > 0 ~ "cytoplasm to nucleus;nucleus to cytoplasm",
    cel == 0 & cyt == 0 & nuc == 0 & cyt_to_nuc > 0 & nuc_to_cyt == 0 ~ "cytoplasm to nucleus",
    cel == 0 & cyt == 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt > 0 ~ "nucleus to cytoplasm",

    cel == 0 & cyt == 0 & nuc == 0 & cyt_to_nuc == 0 & nuc_to_cyt == 0 ~ "none"
  ))

heatmap_matrix <- as.matrix(heatmap_df[,c(-1,-7)])
annotation_row <- data.frame(Compartment = heatmap_df$compartment)
log_mat <- log10(heatmap_matrix+1e-10)

annotation_row <- data.frame(Compartment = heatmap_df$compartment)
rownames(log_mat) = rownames(annotation_row)
color <- brewer.pal(9, "Reds")
library(RColorBrewer)
palette1 <- brewer.pal(12, "Paired")
palette2 <- brewer.pal(8, "Dark2")
palette3 <- brewer.pal(8, "Set2")
palette4 <- brewer.pal(8, "Accent")
all_colors <- c(palette1, palette2, palette3, palette4)
all_colors <- unique(all_colors)[1:32]
compartments <- unique(annotation_row$Compartment)
compartment_colors <- setNames(all_colors[1:length(compartments)], compartments)
ann_colors <- list(Compartment = compartment_colors)
colnames(log_mat) = c("cell", "Cytoplasm", "Nucleus", "Cytoplasm to Nucleus", "Nucleus to Cytoplasm")
pdf("data/results/Final_Figures/Final_figure1/heatmap.pdf", width=10, height=10)
pheatmap(
  log_mat,
  annotation_row = annotation_row,
  annotation_colors = ann_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  annotation_legend = FALSE,
  legend = FALSE,
  fontsize = 20,
  annotation_names_row = FALSE,
  cellheight = 0.1,
  # title = "Communication score (log 10)",
  color = colorRampPalette(color)(100)[14:100]
)
dev.off()

# get the annotation legend separately 
pdf("data/results/Final_Figures/Final_figure1/heatmap_annotation.pdf", width=5, height=5)
lgd = Legend(labels = names(ann_colors$Compartment), legend_gp = gpar(fill = ann_colors$Compartment, fontsize = 14),
       title = "Region", nrow = 23)
grid.newpage()
draw(lgd)
dev.off()

pdf("data/results/Final_Figures/Final_figure1/heatmap_legend.pdf", width=5, height=3)
col_fun = colorRamp2(
  c(-2, -6, -10),                      
  c("#67000D", "#F5553C", "#FDDECF")     
)

lgd = Legend(
  at = c(-2, -6, -10),
  labels = c(expression(10^-2), expression(10^-6), expression(10^-10)),
  col_fun = col_fun,
  title = "Communication score",
  direction = "horizontal"
)
grid.newpage()
draw(lgd)
dev.off()

```

## Gene set enrichment analysis done on the ligands and receptors of significant interactions by region showed communication pathways approximate relation to that region

```{r}
#| message: false
#| warning: false
#| include: true
#| fig-width: 30
#| fig-height: 30 
# Gene set enrichment analysis ********************************************
data = fread("data/processed/training_data_071025.csv")
theme_set(theme_classic(base_size = 14))
# # Download th gmt files
# gs_cc <- msigdbr(species = "Homo sapiens", category="C5", subcategory="GO:CC")
# # gs_bp <- msigdbr(species = "Homo sapiens", category="C5", subcategory="GO:BP")
# # gs_reactome <- msigdbr(species = "Homo sapiens", category="C2", subcategory="CP:REACTOME")
# 
# # convert each to list
# gs_cc_list <- split(gs_cc$gene_symbol, gs_cc$gs_name)
# #gs_bp_list <- split(gs_bp$gene_symbol, gs_bp$gs_name)
# # gs_reactome_list <- split(gs_reactome$gene_symbol, gs_reactome$gs_name)
# 
# # combine
# gene_sets <- c(gs_cc_list)
# 
# ##################################### Ligand - GO enrichment in the cytoplasm ##################################### 
# # cyt_data = data[cyt_score >0,.(cyt_score, ligand)]
# cyt_data = data[,.(cyt_score, ligand)]
# # rank the genes
# ligand_scores = cyt_data[order(cyt_data$cyt_score, decreasing = TRUE), ]
# # ligand_scores = ligand_scores[, .(cyt_score = max(cyt_score, na.rm = TRUE)), by = ligand]
# rankings = ligand_scores$cyt_score
# names(rankings) = make.unique(ligand_scores$ligand)
# ## Run GSEA 
# GSEAres <- fgsea(pathways = gene_sets, 
#                  stats = rankings,
#                  scoreType = 'std', 
#                  minSize = 10,
#                  maxSize = 50,
#                  nproc = 1) 
# # plot
# fgsea_df <- GSEAres %>%
#   arrange(NES) %>%
#   mutate(
#     pathway = factor(pathway, levels = pathway),  # Keep order
#     significant = padj < 0.25                     # TRUE/FALSE significance
#   )
# 
# 
# p1 = ggplot(fgsea_df, aes(x = NES, y = pathway, fill = significant)) +
#   geom_col() +
#   scale_fill_manual(values = c("TRUE" = "springgreen3", "FALSE" = "deeppink")) +
#   geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
#   labs(title = "Ligands in cytoplasm - GSEA",
#        x = "Normalized Enrichment Score (NES)",
#        y = "Pathway",
#        fill = "P padj < 0.25") +
#   theme_minimal(base_size = 14)
# 
# ##################################### Receptor - GO enrichment in the cytoplasm ##################################### 
# cyt_data = data[,.(cyt_score, receptor)]
# # rank the genes
# receptor_scores = cyt_data[order(cyt_data$cyt_score, decreasing = TRUE), ]
# # receptor_scores = receptor_scores[, .(score = max(cyt_score, na.rm = TRUE)), by = receptor]
# rankings = receptor_scores$cyt_score
# names(rankings) = make.unique(receptor_scores$receptor)
# ## Run GSEA 
# GSEAres <- fgsea(pathways = gene_sets, # List of gene sets to check
#                  stats = rankings,
#                  scoreType = 'std', # in this case we have both pos and neg rankings. if only pos or neg, set to 'pos', 'neg'
#                  minSize = 10,
#                  maxSize = 500,
#                  nproc = 1) 
# # plot
# fgsea_df <- GSEAres %>%
#   arrange(NES) %>%
#   mutate(
#     pathway = factor(pathway, levels = pathway),  # Keep order
#     significant = padj < 0.25                    # TRUE/FALSE significance
#   )
# p2 = ggplot(fgsea_df, aes(x = NES, y = pathway, fill = significant)) +
#   geom_col() +
#   scale_fill_manual(values = c("TRUE" = "springgreen3", "FALSE" = "deeppink")) +
#   geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
#   labs(title = "Receptors in cytoplasm - GSEA",
#        x = "Normalized Enrichment Score (NES)",
#        y = "Pathway",
#        fill = "P padj < 0.25") +
#   theme_minimal(base_size = 14)
# 
# ##################################### Ligand - GO enrichment in the nucleus ##################################### 
# nuc_data = data[,.(nuc_score, ligand)]
# # rank the genes
# ligand_scores = nuc_data[order(nuc_data$nuc_score, decreasing = TRUE), ]
# # ligand_scores = ligand_scores[, .(score = max(nuc_score, na.rm = TRUE)), by = ligand]
# rankings = ligand_scores$nuc_score
# names(rankings) = make.unique(ligand_scores$ligand)
# ## Run GSEA 
# GSEAres <- fgsea(pathways = gene_sets, # List of gene sets to check
#                  stats = rankings,
#                  scoreType = 'std', # in this case we have both pos and neg rankings. if only pos or neg, set to 'pos', 'neg'
#                  minSize = 10,
#                  maxSize = 500,
#                  nproc = 1) 
# # plot
# fgsea_df <- GSEAres %>%
#   arrange(NES) %>%
#   mutate(
#     pathway = factor(pathway, levels = pathway),  # Keep order
#     significant = padj < 0.25                    # TRUE/FALSE significance
#   )
# p3 = ggplot(fgsea_df, aes(x = NES, y = pathway, fill = significant)) +
#   geom_col() +
#   scale_fill_manual(values = c("TRUE" = "springgreen3", "FALSE" = "deeppink")) +
#   geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
#   labs(title = "Ligands in nucleus - GSEA",
#        x = "Normalized Enrichment Score (NES)",
#        y = "Pathway",
#        fill = "P padj < 0.25") +
#   theme_minimal(base_size = 14)
# 
# ##################################### Receptor - GO enrichment in the nucleus ##################################### 
# nuc_data = data[,.(nuc_score, receptor)]
# # rank the genes
# receptor_scores = nuc_data[order(nuc_data$nuc_score, decreasing = TRUE), ]
# # receptor_scores = receptor_scores[, .(score = max(nuc_score, na.rm = TRUE)), by = receptor]
# rankings = receptor_scores$nuc_score
# names(rankings) = make.unique(receptor_scores$receptor)
# ## Run GSEA 
# GSEAres <- fgsea(pathways = gene_sets, # List of gene sets to check
#                  stats = rankings,
#                  scoreType = 'std', # in this case we have both pos and neg rankings. if only pos or neg, set to 'pos', 'neg'
#                  minSize = 10,
#                  maxSize = 500,
#                  nproc = 1) 
# # plot
# fgsea_df <- GSEAres %>%
#   arrange(NES) %>%
#   mutate(
#     pathway = factor(pathway, levels = pathway),  # Keep order
#     significant = padj < 0.25                     # TRUE/FALSE significance
#   )
# p4 = ggplot(fgsea_df, aes(x = NES, y = pathway, fill = significant)) +
#   geom_col() +
#   scale_fill_manual(values = c("TRUE" = "springgreen3", "FALSE" = "deeppink")) +
#   geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
#   labs(title = "Receptors in nucleus - GSEA",
#        x = "Normalized Enrichment Score (NES)",
#        y = "Pathway",
#        fill = "P padj < 0.25") +
#   theme_minimal(base_size = 14)
# 
# ggarrange(p1,p2,p3,p4)


# Clusterprofiler
# choose location-relevant sets: GO:CC (+ optional localization/trafficking GO:BP)
gs_cc <- setDT(msigdbr(species = "Homo sapiens", category="C5", subcategory="GO:CC"))
TERM2GENE  <- gs_cc[,.(term = gs_name, gene = gene_symbol)]
unique_genes_df <- read_csv("data/processed/training_data_062325.csv")
universe_genes <- c(unique_genes_df$ligand, unique_genes_df$receptor)%>%unique()

run_ora <- function(detected_genes, universe_genes, TERM2GENE,
                    minGS = 10, maxGS = 500, p_adj = "BH") {
  detected_genes <- intersect(unique(detected_genes), universe_genes)
  # if (length(detected_genes) < 5) {
  #   return(data.table())  
  # }
  ora <- enricher(gene          = detected_genes,
                  TERM2GENE     = TERM2GENE,
                  universe      = universe_genes,
                  pAdjustMethod = p_adj,
                  minGSSize     = minGS,
                  maxGSSize     = maxGS)
  as.data.table(ora@result)[order(p.adjust)]
}

options(enrichplot.colours = c("#1B9E77","#D81B60"))

plot_ora <- function(ora_dt, top = 15, title = "") {
  if (!nrow(ora_dt)) return(invisible(NULL))
  ora_obj <- new("enrichResult", result = as.data.frame(ora_dt))
  barplot(ora_obj, showCategory = top) + ggtitle(title) 
}

pretty_term <- function(x) {
  x <- sub("^(GO(CC|BP|MF)|HALLMARK|REACTOME|KEGG|WP|PID)_", "", x)
  x <- gsub("_", " ", x)
  tools::toTitleCase(tolower(x))
}

res_lig_cyt <- run_ora(
  detected_genes = as.vector(data[cyt_score>0,ligand]),
  universe_genes = universe_genes,   
  TERM2GENE      = TERM2GENE
)

res_lig_cyt$Description = pretty_term(res_lig_cyt$Description)

p1 = plot_ora(res_lig_cyt, top = 15, title = "Ligands - Cytoplasm") + theme_classic() + theme(
      panel.grid.major.y = element_blank(),
      axis.text.y = element_text(
        margin = margin(r = 8),     
        lineheight = 0.7,
        size = 14
      ))

res_rec_cyt <- run_ora(
  detected_genes = as.vector(data[cyt_score>0,receptor]),
  universe_genes = universe_genes,   
  TERM2GENE      = TERM2GENE
)

res_rec_cyt$Description = pretty_term(res_rec_cyt$Description)

p2 = plot_ora(res_rec_cyt, top = 15, title = "Receptors - Cytoplasm") + theme_classic() + theme(
      panel.grid.major.y = element_blank(),
      axis.text.y = element_text(
        margin = margin(r = 8),     
        lineheight = 0.7,
        size = 14
      )) 

res_lig_nuc <- run_ora(
  detected_genes = as.vector(data[nuc_score>0,ligand]),
  universe_genes = universe_genes,   
  TERM2GENE      = TERM2GENE
)

res_lig_nuc$Description = pretty_term(res_lig_nuc$Description)

p3 = plot_ora(res_lig_nuc, top = 15, title = "Ligands - Nucleus")+ theme_classic() + theme(
      panel.grid.major.y = element_blank(),
      axis.text.y = element_text(
        margin = margin(r = 8),    
        lineheight = 0.7,
        size = 14
      ))

res_rec_nuc <- run_ora(
  detected_genes = as.vector(data[nuc_score>0,receptor]),
  universe_genes = universe_genes,   
  TERM2GENE      = TERM2GENE
)

res_rec_nuc$Description = pretty_term(res_rec_nuc$Description)

p4 = plot_ora(res_rec_nuc, top = 15, title = "Receptors - Nucleus") + theme_classic() + theme(
      panel.grid.major.y = element_blank(),
      axis.text.y = element_text(
        margin = margin(r = 8),    
        lineheight = 0.7,
       size = 14
      ))

ggarrange(p1,p2,p3,p4)

# helper to normalize text sizes on any plot
fix_fonts <- function(p, base = 11, y = 9, x = 9, title = 12){
  p + theme(
    text = element_text(size = base),
    axis.text.y = element_text(size = y),
    axis.text.x = element_text(size = x),
    plot.title  = element_text(size = title, face = "bold")
  )
}

p1 <- fix_fonts(p1)
p2 <- fix_fonts(p2)
p3 <- fix_fonts(p3)
p4 <- fix_fonts(p4) 

# if you use patchwork:
combined <- (p1 | p2) / (p3 | p4) &
  theme(                       # applies to ALL subplots uniformly
    text = element_text(size = 12),
    axis.text.y = element_text(size = 12)
  )
combined

ggsave('data/results/Final_Figures/Final_figure1/overlap_test.pdf', height = 15, width = 15)

plot_ora_tree <- function(ora_dt, top = 15, title = "") {
  if (!nrow(ora_dt)) return(invisible(NULL))

  # build enrichResult object
  ora_obj <- new("enrichResult", result = as.data.frame(ora_dt))

  # keep top categories (by adjusted p)
  if (!is.null(ora_obj@result) && nrow(ora_obj@result) > top) {
    keep_ids <- head(ora_obj@result[order(ora_obj@result$p.adjust), "ID"], top)
    ora_obj@result <- ora_obj@result[ora_obj@result$ID %in% keep_ids, , drop = FALSE]
  }

  # compute pairwise term similarity then treeplot
  # requires enrichplot >= 1.20 and DOSE loaded as a dep
  if (requireNamespace("enrichplot", quietly = TRUE)) {
    ora_sim <- enrichplot::pairwise_termsim(ora_obj)
    if ("treeplot" %in% getNamespaceExports("enrichplot")) {
      p <- enrichplot::treeplot(ora_sim, showCategory = top, fontsize = 10,
                                extend = 1,
                                label_format_tiplab = 30,
                                label_format_cladelab = 20,
                                leave_fontsize = 4,
                                clade_fontsize = 4,
                                offset =8,
                                legend_n = 0,
                                geneClusterPanel = "dotplot",
                                group_color = c("#f27042",
                                                "#fcda83", "#b62e48", 
                                                "#7ecba4", "#4965b0"
                                                )) + ggtitle(title)
      return(p)
    } else {
      # fallback: enrichment map if treeplot not available
      p <- enrichplot::emapplot(ora_sim, showCategory = top) + ggtitle(paste0(title, " (emap)"))
      return(p)
    }
  } else {
    stop("Package 'enrichplot' is required for tree/emap plots. Install with BiocManager::install('enrichplot').")
  }
}

# Suppose you have four enrichment results
padj_min <- min(c(res_lig_cyt$p.adjust,
                  res_lig_nuc$p.adjust,
                  res_rec_cyt$p.adjust,
                  res_rec_nuc$p.adjust))

padj_max <- max(c(res_lig_cyt$p.adjust,
                  res_lig_nuc$p.adjust,
                  res_rec_cyt$p.adjust,
                  res_rec_nuc$p.adjust))

gene_ratio_min <- 0
gene_ratio_max <- max(c(res_lig_cyt$Count,
                        res_lig_nuc$Count,
                        res_rec_cyt$Count,
                        res_rec_nuc$Count))


p1 = plot_ora_tree(res_lig_cyt, top = 17, title = "A. Ligands - Cytoplasm") +
  scale_color_gradientn(
    colors = c("#1B9E77", "#1B9E77","#D81B60", "#D81B60"),  # repeated colors
    values = scales::rescale(c(0, 0.05, 0.05, 1)),           # cutoff at 0.05
    limits = c(0, 1),                                        # full range
    name = "FDR-adjusted p-value"
  )+ 
      scale_size_continuous(limits = c(gene_ratio_min, gene_ratio_max), name = "Number of genes") +
      theme(legend.position = "none")
p2 = plot_ora_tree(res_lig_nuc, top = 17, title = "B. Ligands - Nucleus") + 
  scale_color_gradientn(
    colors = c("#1B9E77", "#1B9E77","#D81B60", "#D81B60"),  # repeated colors
    values = scales::rescale(c(0, 0.05, 0.05, 1)),           # cutoff at 0.05
    limits = c(0, 1),                                        # full range
    name = "FDR-adjusted p-value"
  ) + 
  scale_size_continuous(limits = c(gene_ratio_min, gene_ratio_max), name = "Number of genes") +
  theme(legend.position = "none")
p3 = plot_ora_tree(res_rec_cyt, top = 17, title = "C. Receptors - Cytoplasm") + 
  scale_color_gradientn(
    colors = c("#1B9E77", "#1B9E77","#D81B60", "#D81B60"),  # repeated colors
    values = scales::rescale(c(0, 0.05, 0.05, 1)),           # cutoff at 0.05
    limits = c(0, 1),                                        # full range
    name = "FDR-adjusted p-value"
  ) + 
  scale_size_continuous(limits = c(gene_ratio_min, gene_ratio_max), name = "Number of genes") +
  theme(legend.position = "none")
p4 = plot_ora_tree(res_rec_nuc, top = 17, title = "D. Receptors - Nucleus") + 
  scale_color_gradientn(
    colors = c("#1B9E77", "#1B9E77","#D81B60", "#D81B60"),  # repeated colors
    values = scales::rescale(c(0, 0.05, 0.05, 1)),           # cutoff at 0.05
    limits = c(0, 1),                                        # full range
    name = "FDR-adjusted p-value"
  ) + 
  scale_size_continuous(limits = c(gene_ratio_min, gene_ratio_max), name = "Number of genes") +
  theme(legend.position = "none")

combined <- (p1 | p2) / (p3 | p4) &
  theme(                       # applies to ALL subplots uniformly
    text = element_text(size = 20),
    axis.text.y = element_text(size = 0),
    axis.text.x = element_text(size = 0)
  )
combined
ggsave('data/results/Final_Figures/Final_figure1/overlap_test_tree.pdf', height = 15, width = 20)

p1 + theme(legend.position = "bottom")
ggsave('data/results/Final_Figures/Final_figure1/overlap_test_tree_legend.pdf', height = 15, width = 20)


```

## Log2FoldChange differences in the LR communication score of detected between the various regions of the cell

```{r}
#| message: false
#| warning: false
#| include: true
data$color = case_when(
  data$cell_score != 0 & data$nuc_score == 0 & data$cyt_score == 0 ~ "cell",
  data$cell_score != 0 & data$nuc_score == 0 & data$cyt_score != 0 ~ "cell; cytoplasm",
  data$cell_score != 0 & data$nuc_score != 0 & data$cyt_score == 0 ~ "cell; nucleus",
  data$cell_score != 0 & data$nuc_score != 0 & data$cyt_score != 0 ~ "cell; cytoplasm; nucleus",
  .default = "NS"
)
  
data$color = factor(data$color, levels = c(
  "cell",
  "cell; cytoplasm",
  "cell; nucleus",
  "cell; cytoplasm; nucleus"
))
# cyt_score - nuc_score
data$diff =  log2(data$cyt_score + 1e-6) -  log2(data$nuc_score + 1e-6)

library(scales)

p = ggplot(data, aes(x = cell_score, y = diff, color = color)) +
  geom_point(size = 1.8) +
  scale_x_log10(
    breaks = c(1e-6, 1e-4, 1e-2),
    labels = scales::trans_format("log10", math_format(10^.x))
  ) +
  labs(
    x = "Interaction score in the cell (log scale)",
    y = "Log2FC of interaction score \n (Cytoplasm - Nucleus)",
    color = "Localisation"
  ) +
  scale_color_manual(values = c("#a31044", "#f69d5d", "#c8e7a5", "#4a66ae"),
                     guide = guide_legend(nrow = 2, byrow = TRUE)) +
  theme_classic(base_size = 18) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 30),
    legend.title = element_text(size = 30),
    axis.text.x = element_text(size = 22, angle = 0, hjust = 1),
    axis.text.y = element_text(size = 22),
    axis.title = element_text(size = 30),
    axis.line = element_line(size = 0.8)
  ) + coord_cartesian(xlim = c(1e-6, 1e-2))

ggsave('data/results/Final_Figures/Final_figure1/ma_plot.pdf', width = 12, height = 8)
  
```

## Linear projection of CellChat-derived communication score components using PCA. Bottom: PCA plot colored by the hill function of expression (P1 component of CellChat) across the three regions.

```{r}
#| message: false
#| warning: false
#| include: true
lr_df <- read_csv("data/processed/archive/lung_cancer/lr_df.csv")
results <- readRDS("data/processed/archive/cellchat/lung_hig_res.rds")

formula_parts = lapply(results, function (data) {
  P1_vector = c()
  P2_vector = c()
  P3_vector = c()
  P4_vector = c()
  P.spatial_vector = c()
  for (i in 1:nrow(lr_df)){
  P1_vector = append(P1_vector, data@net[["P1"]][lr_df$source[i], lr_df$target[i], lr_df$interaction_name[i]])
  P2_vector = append(P2_vector, data@net[["P2"]][lr_df$source[i], lr_df$target[i], lr_df$interaction_name[i]])
  P3_vector = append(P3_vector, data@net[["P3"]][lr_df$source[i], lr_df$target[i], lr_df$interaction_name[i]])
  P4_vector = append(P4_vector, data@net[["P4"]][lr_df$source[i], lr_df$target[i], lr_df$interaction_name[i]])
  P.spatial_vector = append(P.spatial_vector, data@net[["P.spatial"]][lr_df$source[i], lr_df$target[i], lr_df$interaction_name[i]])
  }
  formula_list = list(P1_vector, P2_vector, P3_vector, P4_vector, P.spatial_vector)
  return(formula_list)
})

names(formula_parts) = c("cell", "cytoplasm", "nucleus")

formula_df = as.data.frame(cbind(do.call(cbind, formula_parts[[1]]), do.call(cbind, formula_parts[[2]]), do.call(cbind, formula_parts[[3]])))
colnames(formula_df) = unlist(lapply(list("cell", "cytoplasm", "nucleus"), function(i){c(paste0(i,"_",'P1_formula'),paste0(i,"_",'P2_formula'),                                                                            paste0(i,"_",'P3_formula'),                                                                            paste0(i,"_",'P4_formula'),                                                                            paste0(i,"_",'P.spatial_formula')
)}))
# remove columns with constant values. They do nothing
formula_df = formula_df[,-c(2,4)]
# retain unique interactions
formula_df$venn_relation = lr_df$venn_relation
formula_df$interaction = lr_df$interaction
formula_df = formula_df[!duplicated(formula_df$interaction),]
pc = prcomp(formula_df[,-c(14,15)],
                   center = TRUE,
                  scale = TRUE)

 formula_df$venn_relation = case_when(
  formula_df$venn_relation == 'cyt' ~ "cytoplasm",
  formula_df$venn_relation == 'nuc' ~ "nucleus",
  formula_df$venn_relation == 'cell + cyt' ~ "cell; cytoplasm",
  formula_df$venn_relation == 'cyt+ nuc' ~ "cytoplasm; nucleus",
  formula_df$venn_relation == 'cell + nuc' ~ "cell; nucleus",
  formula_df$venn_relation == 'cell + cyt + nuc' ~ "cell; cytoplasm; nucleus",
  .default = "cell"
)

# Are the different venn relations clustering? 
pc_df = as.data.frame(pc[["x"]])
pc_df$Compartment = formula_df$venn_relation

ggplot(pc_df, aes(PC1, PC2, color = Compartment)) + 
  geom_point(size = 0.5) +
  theme_classic() +
  geom_polygon(stat = "ellipse", aes(fill = Compartment), alpha = 0.05) +
  ylim(y = c(-3, 10)) + 
  xlim(-10,10) +
  scale_color_manual(values = c("#a31044", "#f69d5d", "#c8e7a5", "#4a66ae","#db5142","#70aca8","#ebf29f"),
                     guide = guide_legend(nrow = 3, byrow = TRUE)) + 
  scale_fill_manual(values = c("#a31044", "#f69d5d", "#c8e7a5", "#4a66ae","#db5142","#70aca8","#ebf29f")) +
  labs(x = "PC1 (Explained variance: 43.7%)",
        y = "PC2 (Explained variance: 20%)") +
  theme_classic(base_size = 18) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 30),
    legend.title = element_text(size = 30),
    axis.text.x = element_text(size = 22, angle = 0, hjust = 1),
    axis.text.y = element_text(size = 22),
    axis.title = element_text(size = 30)
  )
ggsave("data/results/Final_Figures/Final_figure1/pca_plot.pdf", width = 12, height = 8)


pc_df$cell_expression = ifelse(formula_df$cell_P1_formula == 0, 1e-10, formula_df$cell_P1_formula)
# color by product of expression 
p1= ggplot(pc_df, aes(PC1, PC2, color = cell_expression)) + 
  geom_point(size = 1, alpha = 0.7) +
  theme_classic() +
  ylim(y = c(-3, 10)) + 
  xlim(-10,10) +
  scale_color_gradient(low="#fadc85",
                     high="#a31044", space ="Lab", trans = "log10",
    breaks = c(1e-7, 1e-5, 1e-3),
  labels = c(expression(10^-7), expression(10^-5), expression(10^-3))
)+
  labs(x = "PC1 (Explained variance: 43.7%)",
        y = "PC2 (Explained variance: 20%)"
       ) +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 20),
    axis.text = element_text(size = 20),
    axis.title = element_text(size = 20),
    panel.grid.minor = element_blank(),
    legend.title = element_text(size = 20)
  ) +
  guides(color=guide_colorbar(title="LR Expression(log10)"))

pc_df$cyt_expression = ifelse(formula_df$cytoplasm_P1_formula == 0, 1e-10, formula_df$cytoplasm_P1_formula)
p2 = ggplot(pc_df, aes(PC1, PC2, color = cyt_expression)) + 
  geom_point(size = 1, alpha = 0.7) +
  theme_classic() +
  ylim(y = c(-3, 10)) + 
  xlim(-10,10) +
  scale_color_gradient(low="#fadc85",
                     high="#a31044", space ="Lab", trans = "log10") +
  labs(x = "PC1 (Explained variance: 43.7%)",
        y = "PC2 (Explained variance: 20%)"
       ) +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 20),
    axis.text = element_text(size = 20),
    axis.title = element_text(size = 20),
    panel.grid.minor = element_blank(),
    legend.title = element_text(size = 20)
  ) +
  guides(color=guide_colorbar(title="LR Expression(log10)"))

pc_df$nuc_expression = ifelse(formula_df$nucleus_P1_formula == 0, 1e-10, formula_df$nucleus_P1_formula)
p3 = ggplot(pc_df, aes(PC1, PC2, color = nuc_expression)) + 
  geom_point(size = 1, alpha = 0.7) +
  theme_classic() +
  ylim(y = c(-3, 10)) + 
  xlim(-10,10) +
  scale_color_gradient(low="#fadc85",
                     high="#a31044", space ="Lab", trans = "log10") +
  labs(x = "PC1 (Explained variance: 43.7%)",
        y = "PC2 (Explained variance: 20%)"
       ) +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 20),
    axis.text = element_text(size = 20),
    axis.title = element_text(size = 20),
    panel.grid.minor = element_blank(),
    legend.title = element_text(size = 20)
  ) +
  guides(color=guide_colorbar(title="LR Expression(log10)"))
ggarrange(p1,p2,p3, ncol = 3, common.legend = TRUE, legend = "none")

ggsave("data/results/Final_Figures/Final_figure1/pca_plot_expr.pdf", width = 20)

legend <- get_legend(
  p1 +
    guides(color = guide_colorbar(
      title = "LR Expression (log10)",
      title.position = "bottom",
      title.hjust = 0.5,
      barwidth = 15,
      barheight = 1,
      direction = "horizontal"
    ))
)


ggsave("data/results/Final_Figures/Final_figure1/pca_legend.pdf", legend, width=4, height=1)

```
