---
title: "Figure 4"
author: "rshreej"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-expand: true
    number-sections: true
    theme: united
    highlight: tango
    code-fold: true
    code-tools: true
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| message: false
#| include: false
#| warning: false
library(data.table)
library(circlize)
library(ggvenn)
library(cowplot)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(fgsea)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot) 
library(patchwork)
library(ggbiplot)
library(gridGraphics)
library(gridExtra)
library(igraph)
library(RColorBrewer)
```
# Training a hierarchical model involving a classifier and regressor to predict the communication scores of the subcellular regions of the cell

```{r}
#| message: false
#| include: true
#| warning: false
theme_set(
  theme_classic(base_size = 14) +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
    )
)
```

## 4A: Classification and regression metrics on training the model without the spatial feature 
```{r}
#| message: false
#| include: true
#| warning: false
# AUC curve 
ddir = "data/processed/single_cell/results_P1spatial_removed/"
roc_table = fread(file.path(ddir, "roc_tables_P1_trained.csv"))
xgb_voting = fread(file.path(ddir, "xgboost_voting_metrics_P1_trained.csv"))
roc_table = inner_join(roc_table, xgb_voting[,.(Dataset, auc)], by = 'Dataset')
roc_table$Dataset <- factor(roc_table$Dataset,
  levels = unique(roc_table$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

roc_table$dataset_label <- paste0(roc_table$Dataset, " (AUC=", round(roc_table$auc, 2), ")")

colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )
ggplot(roc_table, aes(x = `False Positive Rate`, y = `True Positive Rate`, color = dataset_label)) +
  geom_line(size = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "False Positive Rate",
       y = "True Positive Rate") +
  scale_color_manual(values = colors) +
  theme_classic(base_size = 20) +
  theme(legend.position = "right") +
  guides(color = guide_legend(title = "Dataset"))

ggsave("data/results/Final_Figures/Final_figure4/roc_table.pdf", width =10)
# 
# # Regression metrics 
# r2 = melt(dt, id.vars= c('Dataset'), measure.vars=c("cyt_R²","nuc_R²")) 
# r2$variable <- factor(r2$variable,
#   levels = c("cyt_R²", "nuc_R²"),
#   labels = c("R² – cytoplasm", "R² – nucleus")
# )
# 
# r2_plot = ggplot(r2, aes(x = variable, y = value, fill = variable)) +
#   geom_violin(trim = FALSE, color = "black") +
#   geom_boxplot(width = 0.15, color = "black", fill = "white", outlier.size = 0.8) +
#   scale_fill_manual(values = c("#b8aeea", "#f4b974")) +
#   labs(
#     x = NULL,
#     y = expression(R^2)
#   ) +
#   theme_classic(base_size = 20) +
#   theme(
#     legend.position = "none",
#     plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
#     axis.text.y = element_text(face = "bold", size = 20),
#     axis.text.x = element_text(angle = 90, hjust = 1, size = 20, vjust = 1)
#     
#   ) 
# 
# nrmse = melt(dt, id.vars= c('Dataset'), measure.vars=c("cyt_NRMSE","nuc_NRMSE")) 
# nrmse$variable <- factor(nrmse$variable,
#   levels = c("cyt_NRMSE", "nuc_NRMSE"),
#   labels = c("NRMSE – cytoplasm", "NRMSE – nucleus")
# )
# 
# nrmse_plot = ggplot(nrmse, aes(x = variable, y = value, fill = variable)) +
#   geom_violin(trim = FALSE, color = "black") +
#   geom_boxplot(width = 0.15, color = "black", fill = "white", outlier.size = 0.8) +
#   scale_fill_manual(values = c("#b8aeea", "#f4b974")) +
#   labs(
#     x = NULL,
#     y = "Normalized RMSE"
#   ) +
#   theme_classic(base_size = 20) +
#   theme(
#     legend.position = "none",
#     plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
#     axis.text.y = element_text(face = "bold"),
#     axis.text.x = element_text(angle = 90, size = 20, hjust = 1, vjust = 1)
#   )
# 
# r2_plot + nrmse_plot
# ggsave("data/results/Final_Figures/Final_figure4/distribution_regression_best_model.pdf")
# 
# # combined metrics 
# xgb_voting[ , NRMSE_weighted_auc := auc * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) ]
# xgb_voting[, R2_weighted_auc := auc * ((`cyt_R²` + `nuc_R²`) / 2)]
# xgb_voting[ , NRMSE_weighted_recall := `Class_1 Recall` * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) ]
# xgb_voting[ , R2_weighted_recall := `Class_1 Recall` * ((`cyt_R²` + `nuc_R²`)/2) ]
# df = melt(xgb_voting, id.vars= c('Dataset'), measure.vars=c("R2_weighted_auc","R2_weighted_recall", "NRMSE_weighted_auc", "NRMSE_weighted_recall")) 
# 
# df$variable <- factor(df$variable,
#   levels = c("R2_weighted_auc", "R2_weighted_recall",
#              "NRMSE_weighted_auc", "NRMSE_weighted_recall"),
#   labels = c("AUC × R²", "Recall × R²",
#              "AUC × (1−NRMSE)", "Recall × (1−NRMSE)")
# )
# 
# colors = c("#a0bfc2", "#083162", "#d38659", "#71a82f")
# p1 = ggplot(df, aes(x = variable, y = value)) +
#   stat_summary(fun = mean, geom = "bar", width = 0.7, aes(fill = variable)) +
#   stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
#                geom = "errorbar", width = 0.2, linewidth = 0.6) +
#   geom_jitter(width = 0.15, size = 2, alpha = 0.8, color = "black") +
#   scale_fill_manual(values = colors)+ 
#   labs(
#     x = "Metric", y = "Score"
#   ) +
#   theme_classic(base_size = 20) +
#   theme(legend.position = "none",
#         axis.text.x = element_text(angle = 90, hjust = 1))
# 
# p2 = ggplot(df, aes(x = variable, y = value)) +
#   geom_violin(aes(alpha = 0.7, fill = variable)) +
#   geom_boxplot(width = 0.3,  color = "black", fill = "white" ) + #aes(fill = variable)) +
#   geom_jitter(width = 0.15, size = 2, alpha = 0.8, color = "black") +
#   labs(
#     x = "Metric", y = "Score"
#   ) +
#   scale_fill_manual(values = colors) + 
#   scale_color_manual(values = colors) + 
#   theme_classic(base_size = 20) +
#   theme(legend.position = "none",
#         axis.text.x = element_text(angle = 90, hjust = 1))
# ggarrange(p1, p2, nrow = 1)


xgb_voting = xgb_voting[,c(1,12,15,16,18,19,21)]
xgb_voting[ , geometric_mean := (auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]

dt = melt(xgb_voting, id.vars= c('Dataset')) 
dt$variable <- factor(dt$variable,
  levels = c("auc","macro_recall","cyt_R²","nuc_R²",
             # "cyt_RMSE","nuc_RMSE",
             "cyt_NRMSE","nuc_NRMSE","geometric_mean"),
  labels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "Composite metric")
)

#  metrics 
p1 = ggplot(dt, aes(x = variable, y = value, fill = variable, color = variable)) +
  geom_violin(trim = TRUE) +
  geom_boxplot(width = 0.15, color = "black", fill = "white", outlier.size = 0.8) +
  scale_fill_manual(values = colors) +
  scale_color_manual(values = colors) +
  labs(
    x = NULL,
    y = 'Value'
  ) +
  theme_classic(base_size = 20) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1,vjust = 1)
) + coord_cartesian(ylim = c(0,1))

p2 = ggplot(dt, aes(x = variable, y = value)) +
  stat_summary(fun = mean, geom = "bar", width = 0.7, aes(fill = variable)) +
  stat_summary(fun.data = mean_sdl, fun.args = list(mult = 1),
               geom = "errorbar", width = 0.2, linewidth = 0.6) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.8, color = "black") +
  scale_fill_manual(values = colors)+ 
  labs(
    x = "", y = ""
  ) +
  theme_classic(base_size = 20) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1)) + coord_cartesian(ylim = c(0,1))

ggarrange(p1,p2, align = "h")

ggsave("data/results/Final_Figures/Final_figure4/combined_no_spatial.pdf", width = 12)
```

## 4B: Difference in the the metrics with and without the spatial component during training
```{r}
#| message: false
#| include: true
#| warning: false

metrics = fread("data/processed/metrics_all_combinations_trainings_10032025.csv")
metrics[ , geometric_mean := (roc_auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]

df_long <- metrics[,c(9,28,10,11, 12)]

spatial = df_long[condition == "with_spatial",]
no_spatial = df_long[condition == "without_spatial",]
#check
identical(spatial[,3:5], no_spatial[,3:5])

df_long = spatial[,3:5]
df_long$difference = spatial$geometric_mean - no_spatial$geometric_mean
df_long = df_long[test_name != "spe_list_lung_cellchat",]
df_long = df_long[test_name != "spe_list_brainad_cellchat",]

df_long = na.omit(df_long)

df_long$test_name <- factor(df_long$test_name,
  levels = unique(df_long$test_name),
  labels = c(
             "Brain (GL)",
             "Breast Cancer",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

colors_box = c("#d4dda6", "#78c2e3", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )


ggplot(df_long, aes(x = test_name, y = difference, fill = test_name)) +
  geom_boxplot(outliers = TRUE) +
  scale_fill_manual(values = colors_box) + 
  # facet_wrap(~metric, scales = "free_y", ncol = 4) +
  labs( x = "", y = "Difference in Composite metric") + 
  theme_classic(base_size = 20) +
  theme(
    axis.text.x = element_text(angle = 45, size = 16, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "",
    strip.text = element_text(face = "bold", size = 16, hjust = 0.5),
axis.text.y = element_text(size = 16)
  )
ggsave("data/results/Final_Figures/Final_figure4/boxplot_difference.pdf", width =8)


# # Heatmap of combined metrics 
# dt = fread(file.path(ddir,"xgboost_voting_metrics_P1_trained.csv"))
# dt[ , NRMSE_weighted_auc := sqrt(auc * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) ) ]
# dt[ , R2_weighted_auc    := sqrt(auc * ((`cyt_R²` + `nuc_R²`) / 2)) ]
# dt[ , NRMSE_weighted_recall := sqrt(`macro_recall` * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) ) ]
# dt[ , R2_weighted_recall    := sqrt(`macro_recall` * ((`cyt_R²` + `nuc_R²`) / 2)) ]
# no_spatial = dt[, .(NRMSE_weighted_recall, R2_weighted_recall, NRMSE_weighted_auc, R2_weighted_auc)]
# 
# dt = fread("data/processed/plots_metrics/comparison_baselearners/xgboost_voting_metrics.csv")
# dt[ , NRMSE_weighted_auc := sqrt(auc * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) ) ]
# dt[ , R2_weighted_auc    := sqrt(auc * ((`cyt_R²` + `nuc_R²`) / 2)) ]
# dt[ , NRMSE_weighted_recall := sqrt(`macro_recall` * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) ) ]
# dt[ , R2_weighted_recall    := sqrt(`macro_recall` * ((`cyt_R²` + `nuc_R²`) / 2)) ]
# spatial = dt[, .(NRMSE_weighted_recall, R2_weighted_recall, NRMSE_weighted_auc, R2_weighted_auc)]
# 
# heatmap_df = setDT(spatial - no_spatial)
# heatmap_df$Dataset = dt$Dataset
#   
# heatmap_df = melt(heatmap_df, id.vars= c('Dataset'), measure.vars=c("R2_weighted_auc","R2_weighted_recall", "NRMSE_weighted_auc", "NRMSE_weighted_recall"),
#                   value.name = "Difference") 
# 
# # mapping
# heatmap_df$variable <- factor(
#   heatmap_df$variable,
#   levels = c("R2_weighted_auc", "R2_weighted_recall",
#              "NRMSE_weighted_auc", "NRMSE_weighted_recall"),
#   labels = c("GM(AUC, R²)", 
#              "GM(Recall, R²)",
#              "GM(AUC, 1−NRMSE)", 
#              "GM(Recall, 1−NRMSE)")
# )
# 
# heatmap_df$Dataset <- factor(heatmap_df$Dataset,
#   levels = unique(heatmap_df$Dataset),
#   labels = c("Brain (AD)",
#              "Brain (GL)",
#              "Breast Cancer",
#              "Lung",
#              "Lung (High-Res)",
#              "Lymph Node",
#              "Ovarian",
#              "Pancreas",
#              "Skin")
# )
# 
# 
# # plot the heatmap
# ggplot(heatmap_df, aes(x = variable, y = Dataset, fill = Difference)) +
#   geom_tile(color = "black", linewidth = 0.1) + 
#   scale_fill_gradientn (colours = c("white", "#78b4d9", "#073266")) + 
#   labs(x = "Evaluation Metric", y = "Dataset") +
#   theme_minimal(base_size = 13) +
#   theme(
#     axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
#     panel.grid = element_blank(),
#     legend.position = "right",
#     strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
# axis.text.y = element_text(size = 12)
#   ) +
#   coord_fixed() 
# ggsave("data/results/Final_Figures/Final_figure4/heatmap_difference.pdf")
```

## 4C. Heatmap of the LR pair scores (sum and counts) in the model with and without P.spatial for the lung cancer dataset. 
```{r}
#| message: false
#| include: true
#| warning: false


cytoplasm = fread("data/results/predictions/scatter_data_cyt_P1.csv")
nucleus = fread("data/results/predictions/scatter_data_nuc_P1.csv")
cytoplasm_spatial = fread("data/results/predictions/scatter_data_cyt_fullmodel.csv")
nucleus_spatial = fread("data/results/predictions/scatter_data_nuc_fullmodel.csv")

# sum and count of interactions between each cell type 
count_function = function(data){
  shift_val = min(0, min(data$pred))
  data$score = data$pred - shift_val
  # compute the network plot
  links = data.table(
    source= data$source,
    target= data$target,
    importance= data$score
    )
  edges_count <- links[, .(count = .N), by = .(source, target)]
  edges_sum   <- links[, .(sum = sum(importance, na.rm = TRUE)), 
                     by = .(source, target)]
  nodes <- unique(c(links$source, links$target))
  all_pairs <- CJ(source = nodes, target = nodes, unique=TRUE)
  matrix <- merge(all_pairs, edges_count, by=c("source","target"), all.x=TRUE)
  matrix <- merge(matrix, edges_sum,   by=c("source","target"), all.x=TRUE)
  matrix[is.na(count), count := 0]
  matrix[is.na(sum),   sum   := 0]
}

data = cytoplasm[group == 'spe_list_lung_cellchat',]
counts = count_function(data)

p1 = ggplot(counts, aes(x=source, y=target, fill=sum)) +
  geom_tile(color = "black", linewidth = 0.1) +
  geom_text(aes(label = round(sum, 2)), color="black", size=3) +
  scale_fill_gradientn (colours = c("white", "#edda98", "#d38558"), name = "Interaction \n score") +
  labs(x = "Sender", y = "Receiver") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  ) +
  coord_fixed() + ggtitle ("Cytoplasmic  Interaction - Lung (without spatial)")

data = nucleus[group == 'spe_list_lung_cellchat',]
counts = count_function(data)

p2 = ggplot(counts, aes(x=source, y=target, fill=sum)) +
  geom_tile(color = "black", linewidth = 0.1) +
  geom_text(aes(label = round(sum, 2)), color="black", size=3) +
  scale_fill_gradientn (colours = c("white", "#edda98", "#d38558"), name = "Interaction \n score") +
  labs(x = "Sending cell", y = "Receiving cell") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  ) +
  coord_fixed() + ggtitle ("Nuclear  Interaction - Lung (without spatial)")

data = cytoplasm_spatial[group == 'spe_list_lung_cellchat',]
counts = count_function(data)

p3 = ggplot(counts, aes(x=source, y=target, fill=sum)) +
  geom_tile(color = "black", linewidth = 0.1) +
  geom_text(aes(label = round(sum, 2)), color="black", size=3) +
  scale_fill_gradientn (colours = c("white", "#edda98", "#d38558"), name = "Interaction \n score") +
  labs(x = "Sending cell", y = "Receiving cell") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  ) +
  coord_fixed() + ggtitle ("Cytoplasmic  Interaction - Lung (with spatial)")

data = nucleus_spatial[group == 'spe_list_lung_cellchat',]
counts = count_function(data)

p4 = ggplot(counts, aes(x=source, y=target, fill=sum)) +
  geom_tile(color = "black", linewidth = 0.1) +
  geom_text(aes(label = round(sum, 2)), color="black", size=3) +
  scale_fill_gradientn (colours = c("white", "#edda98", "#d38558"), name = "Interaction \n score") +
  labs(x = "Sending cell", y = "Receiving cell") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  ) +
  coord_fixed() + ggtitle ("Nuclear Interaction - Lung (with spatial)")

ggarrange (p1, p2, p3, p4, common.legend = T)

ggsave("data/results/Final_Figures/Final_figure4/heatmap_connections.pdf", height = 20, width = 20)
```


## 4D. Is there a variation between the scores if there is lesser number of training datasets? Does including more training dataset counter the effect of not training without spatial features?
```{r}
# plot 1, shows the classification metrics faceted by training dataset
class_dir <- "data/processed/P1_training/classification_metrics/"
files = list.files(class_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(class_dir, i))
  dat$sample = sub("classification_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
metrics = do.call(rbind, file_list)
perclass_metrics = fread("data/processed/P1_training/final_classification_metrics.csv")
colnames(perclass_metrics)[11] = "group"
perclass_metrics$sample = sub("classification_spe_list_(.*)_cellchat\\.csv$", "\\1", perclass_metrics$dataset)
perclass_metrics = perclass_metrics[,-c(9,10,12)]
metrics = metrics[perclass_metrics, on = .(sample, group)]
metrics_p1 = metrics

# same for scores
class_dir <- "data/processed/scores_training/classification_metrics/"
files = list.files(class_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(class_dir, i))
  dat$sample = sub("classification_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
# join with the metrics with perclass metrics 
metrics = do.call(rbind, file_list)
perclass_metrics = fread("data/processed/scores_training/final_classification_metrics.csv")
colnames(perclass_metrics)[11] = "group"
perclass_metrics$sample = sub("classification_spe_list_(.*)_cellchat\\.csv$", "\\1", perclass_metrics$dataset)
perclass_metrics = perclass_metrics[,-c(9,10,12)]
metrics = metrics[perclass_metrics, on = .(sample, group)]
metrics_sc = metrics

metrics_sc$condition <- "with_spatial"
metrics_p1$condition <- "without_spatial"

# read and bind with combo_number
dict = fread("data/processed/combinations_iters.csv")
dict$sample = sub("spe_list_(.*)_cellchat", "\\1", dict$test_name)
colnames(dict)[2] = "group"
metrics_all <- rbind(metrics_sc, metrics_p1)
metrics_all = metrics_all[dict, on = .(group, sample)]
metrics_all_class = metrics_all

# df_long <- metrics_all %>%
#   pivot_longer(cols = c(accuracy, f1, roc_auc, macro_precision, macro_recall),
#                names_to = "metric", values_to = "value")

df_long <- metrics_all %>%
   pivot_longer(cols = colnames(metrics)[c(1:4,7:14)],
                names_to = "metric", values_to = "value")

df_summary <- df_long %>%
  group_by(combo_size, condition, metric) %>%
  summarise(mean_val = mean(value, na.rm = TRUE),
            sd_val = sd(value, na.rm = TRUE), .groups = "drop")

ggplot(df_summary, aes(x = combo_size, y = mean_val, color = condition, group = condition)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean_val - sd_val, ymax = mean_val + sd_val), width = 0.2) +
  facet_wrap(~metric, scales = "free_y", ncol = 4) +
  theme_classic()

ggplot(df_long, aes(x = factor(combo_size),
                     y = value,
                     fill = condition)) +
     geom_boxplot(outlier.size = 0.5, position = position_dodge(width = 0.8)) +
     facet_wrap(~metric, scales = "free_y") +
     theme_classic() +
     labs(x = "Combo size", y = "Metric value")

# plot 2 shows regression faceted by training dataset

# Regression metrics 

# with spatial 
cyt_reg_dir <- "data/processed/scores_training/cyt_metrics"
nuc_reg_dir <- "data/processed/scores_training/nuc_metrics"
files = list.files(cyt_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(cyt_reg_dir, i))
  dat$sample = sub("cyt_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
cyt_metrics = do.call(rbind, file_list)
colnames(cyt_metrics)[1:3] = paste0("cyt_",colnames(cyt_metrics)[1:3])

files = list.files(nuc_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(nuc_reg_dir, i))
  dat$sample = sub("nuc_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
nuc_metrics = do.call(rbind, file_list)
colnames(nuc_metrics)[1:3] = paste0("nuc_",colnames(nuc_metrics)[1:3])

regression_metrics = cyt_metrics[nuc_metrics, on = .(group, sample)]
regression_metrics$`cyt_R²` = ifelse(regression_metrics$`cyt_R²`<0, 0, regression_metrics$`cyt_R²`)
regression_metrics$`nuc_R²` = ifelse(regression_metrics$`nuc_R²`<0, 0, regression_metrics$`nuc_R²`)

regression_metrics$condition = "with_spatial"
regression_metrics_sc = regression_metrics


cyt_reg_dir <- "data/processed/P1_training//cyt_metrics"
nuc_reg_dir <- "data/processed/P1_training/nuc_metrics"
files = list.files(cyt_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(cyt_reg_dir, i))
  dat$sample = sub("cyt_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
cyt_metrics = do.call(rbind, file_list)
colnames(cyt_metrics)[1:3] = paste0("cyt_",colnames(cyt_metrics)[1:3])

files = list.files(nuc_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(nuc_reg_dir, i))
  dat$sample = sub("nuc_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
nuc_metrics = do.call(rbind, file_list)
colnames(nuc_metrics)[1:3] = paste0("nuc_",colnames(nuc_metrics)[1:3])

regression_metrics = cyt_metrics[nuc_metrics, on = .(group, sample)]
regression_metrics$`cyt_R²` = ifelse(regression_metrics$`cyt_R²`<0, 0, regression_metrics$`cyt_R²`)
regression_metrics$`nuc_R²` = ifelse(regression_metrics$`nuc_R²`<0, 0, regression_metrics$`nuc_R²`)

regression_metrics$condition = "without_spatial"
regression_metrics_p1 = regression_metrics

# read and bind with combo_number
dict = fread("data/processed/combinations_iters.csv")
dict$sample = sub("spe_list_(.*)_cellchat", "\\1", dict$test_name)
colnames(dict)[2] = "group"
metrics_all_reg <- rbind(regression_metrics_sc, regression_metrics_p1)
metrics_all_reg = metrics_all_reg[dict, on = .(group, sample)]

# df_long <- metrics_all %>%
#   pivot_longer(cols = c(accuracy, f1, roc_auc, macro_precision, macro_recall),
#                names_to = "metric", values_to = "value")

df_long <- metrics_all_reg %>%
   pivot_longer(cols = colnames(metrics_all_reg)[c(1,3,6,8)],
                names_to = "metric", values_to = "value")

df_summary <- df_long %>%
  group_by(combo_size, condition, metric) %>%
  summarise(mean_val = mean(value, na.rm = TRUE),
            sd_val = sd(value, na.rm = TRUE), .groups = "drop")

ggplot(df_summary, aes(x = combo_size, y = mean_val, color = condition, group = condition)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean_val - sd_val, ymax = mean_val + sd_val), width = 0.2) +
  facet_wrap(~metric, scales = "free_y", ncol = 4) +
  theme_classic()

ggplot(df_long, aes(x = factor(combo_size),
                     y = value,
                     fill = condition)) +
     geom_boxplot(outlier.size = 0.5, position = position_dodge(width = 0.8)) +
     facet_wrap(~metric, scales = "free_y") +
     theme_classic() +
     labs(x = "Combo size", y = "Metric value")

metrics = metrics_all_reg[metrics_all_class, on = .(group, sample, condition)]

# plot 3 combined metrics faceted by training dataset

combo_metrics = data.table(
  `GM(AUC, R²)`          = sqrt(metrics$roc_auc * ((metrics$`cyt_R²` + metrics$`nuc_R²`) / 2)),
  `GM(Recall, R²)`       =  sqrt(metrics$macro_recall * ((metrics$`cyt_R²` + metrics$`nuc_R²`) / 2)),
  `GM(AUC, 1−NRMSE)`     = sqrt(metrics$roc_auc * (1 - ((metrics$cyt_NRMSE + metrics$nuc_NRMSE) / 2))),
  `GM(Recall, 1−NRMSE)`  = sqrt(metrics$macro_recall * (1 - ((metrics$cyt_NRMSE + metrics$nuc_NRMSE) / 2))),
  combo_size             = metrics$combo_size, 
  condition              = metrics$condition,
  sample                 = metrics$sample,
  group                  = metrics$group
)


df_long <- combo_metrics %>%
   pivot_longer(cols = colnames(combo_metrics)[c(1:4)],
                names_to = "metric", values_to = "value")

df_summary <- df_long %>%
  group_by(combo_size, condition, metric) %>%
  summarise(mean_val = mean(value, na.rm = TRUE),
            sd_val = sd(value, na.rm = TRUE)/sqrt(n()), .groups = "drop")

ggplot(df_summary, aes(x = combo_size, y = mean_val, color = condition, group = condition)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean_val - sd_val, ymax = mean_val + sd_val), width = 0.2) +
  facet_wrap(~metric, scales = "free_y", ncol = 4) +
  theme_classic()

ggplot(df_long, aes(x = metric,
                     y = value,
                     fill = condition)) +
     geom_boxplot(outlier.size = 0.5, position = position_dodge(width = 0.8)) +
     theme_classic() +
     labs(x = "Combo size", y = "Metric value")

# colors = c("#f57e6f", "#d2dba4", "#78c3e7", "#f2a7db")
p = ggplot(df_long, aes(x = metric, y = value, fill = condition)) +
  geom_violin(width = 1, trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.2, color = "black", outlier.size = 0.5, position = position_dodge(width = 0.9)) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "none"
  ) +
  labs(
    x = "Evaluation metrics (GM = Geometric Mean)",
    y = "Value"
  )

```





## 4E. Network plots on the single cell data of the cell, predicted cytoplasm and nucleus 
```{r}
cytoplasm = fread("data/results/predictions/scatter_data_cyt_single_cell.csv")
nucleus = fread("data/results/predictions/scatter_data_nuc_single_cell.csv")
count_function = function(data){
  shift_val = min(0, min(data$pred))
  data$score = data$pred - shift_val
  # compute the network plot
  links = data.table(
    source= data$source,
    target= data$target,
    importance= data$score
    )
  edges_count <- links[, .(count = .N), by = .(source, target)]
  edges_sum   <- links[, .(sum = sum(importance, na.rm = TRUE)), 
                     by = .(source, target)]
  nodes <- unique(c(links$source, links$target))
  all_pairs <- CJ(source = nodes, target = nodes, unique=TRUE)
  matrix <- merge(all_pairs, edges_count, by=c("source","target"), all.x=TRUE)
  matrix <- merge(matrix, edges_sum,   by=c("source","target"), all.x=TRUE)
  matrix[is.na(count), count := 0]
  matrix[is.na(sum),   sum   := 0]
}
# for each of the group, shift the prediction and then construct adjacency and from the network plot
data = cytoplasm
# shift the scale 
shift_val = min(0, min(data$pred))
data$score = data$pred - shift_val
# compute the network plot
links = data.table(
  source= data$source,
  target= data$target,
  importance= data$score
  )
edges_count <- links[, .(weight = .N), by = .(source, target)]
edges_sum   <- links[, .(weight = sum(importance, na.rm = TRUE)), 
                   by = .(source, target)]
nodes = unique(c(data$source, data$target))
network <- graph_from_data_frame(d=edges_count, vertices=nodes, directed=F) 
coul  <- brewer.pal(9, "Set1") 
my_color = colorRampPalette(coul)(20)
pdf("data/results/Final_Figures/Final_figure4/scnetworks.pdf", width = 15)
par(mfrow = c(1,2))
# plot(network, vertex.color=coul, edge.width=0.2*(E(network)$weight), main = "Cyt Network (counts)", layout=layout.circle)
# plots_count = recordPlot()
network <- graph_from_data_frame(d=edges_sum, vertices=nodes, directed=F) 
plot(network, vertex.color=coul, edge.width=0.2*(E(network)$weight), main = "Cyt Network (sum)", layout=layout.circle)
plots_sum = recordPlot()

data = nucleus
# shift the scale 
shift_val = min(0, min(data$pred))
data$score = data$pred - shift_val
# compute the network plot
links = data.table(
  source= data$source,
  target= data$target,
  importance= data$score
  )
edges_count <- links[, .(weight = .N), by = .(source, target)]
edges_sum   <- links[, .(weight = sum(importance, na.rm = TRUE)), 
                   by = .(source, target)]
nodes = unique(c(data$source, data$target))
network <- graph_from_data_frame(d=edges_count, vertices=nodes, directed=F) 
coul  <- brewer.pal(9, "Set1") 
my_color = colorRampPalette(coul)(20)
# plot(network, vertex.color=coul, edge.width=0.2*(E(network)$weight), main = "Nuc Network (counts)", layout=layout.circle)
# plots_count = recordPlot()
network <- graph_from_data_frame(d=edges_sum, vertices=nodes, directed=F) 
plot(network, vertex.color=coul, edge.width=0.2*(E(network)$weight), main = "Nuc Network (sum):", layout=layout.circle)
plots_sum = recordPlot()
dev.off()



data = cytoplasm
counts_cyt = count_function(data)

data = nucleus
counts_nuc = count_function(data)

diff_counts <- merge(counts_cyt, counts_nuc, by = c("source","target"), suffixes = c("_cyto","_nuc")) %>%
  mutate(diff = sum_cyto - sum_nuc)

ggplot(diff_counts, aes(x = source, y = target, fill = diff)) +
  geom_tile(color = "black") +
  scale_fill_gradient2(low = "#366785", mid = "#cccccc", high = "#c85200", midpoint = 0,  name = "Interaction score \n difference \n (cytoplasm - nucleus)") +
  theme_minimal(base_size = 22) +
  theme(
    axis.text.x = element_text(angle = 90, size = 20, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    strip.text = element_text(face = "bold", size = 20, hjust = 0.5),
axis.text.y = element_text(size = 20)
  ) + labs(x = "Sender", y = "Receiver")

ggsave("data/results/Final_Figures/Final_figure4/scnetworks_heatmap.pdf", width = 12)
```

