---
title: "Figure 4"
author: "rshreej"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-expand: true
    number-sections: true
    theme: united
    highlight: tango
    code-fold: true
    code-tools: true
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| message: false
#| include: false
#| warning: false
library(data.table)
library(circlize)
library(ggvenn)
library(ggridges)
library(cowplot)
library(pheatmap)
library(cowplot)
library(ComplexHeatmap)
library(fgsea)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot) 
library(patchwork)
library(ggbiplot)
library(gridGraphics)
library(gridExtra)
library(igraph)
library(RColorBrewer)
library(pheatmap)
library(ggvenn)
library(cowplot)
library(ggVennDiagram)
library(magick)
library(ragg)
library(tidyr)
library(dplyr)
```
# Training a hierarchical model involving a classifier and regressor to predict the communication scores of the subcellular regions of the cell

```{r}
#| message: false
#| include: true
#| warning: false
theme_set(
  theme_classic(base_size = 14) +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
    )
)
```

## S2: Upset plots of the LR communicating between each of the regions
```{r}
#| message: false
#| include: true
#| warning: false
data <- fread("data/processed/training_data_071025.csv")
data$unique_id = paste0(data$lr_pair, "_", data$source, "_", data$target)
venn_data = list()
plot_list = list()
for (i in unique(data$sample)){
  subset = data[sample == i,]
  data_list = list(
    `WC-->WC` = as.vector(subset[cell_pval <0.05 & cell_score >0,unique_id]),
    `C-->C` = as.vector(subset[cyt_pval <0.05 & cyt_score > 0,unique_id]),
    `N-->N` = as.vector(subset[nuc_pval <0.05& nuc_score > 0,unique_id])
  )
  venn_data[[i]] = data_list
}


files = list.files('data/processed/cyt_to_nuc')
cyt_to_nuc = list()
nuc_to_cyt = list()
for (i in files){
  filename = gsub(".rds", "", i)
  obj = readRDS(file.path('data/processed/cyt_to_nuc', i))
  scores <- as.data.frame(as.table(obj[["Prob_P2_P3_P4"]]))
  pval <- as.data.frame(as.table(obj[["PvalP2_P3_P4"]]))
  colnames(scores) = c("source", "target", "lr_pair", "score")
  colnames(pval) = c("source", "target", "lr_pair", "pval")
  df = setDT(inner_join(scores, pval, by = intersect(colnames(scores), colnames(pval))))
  df$unique_id = paste0(df$lr_pair, "_", df$source, "_", df$target)
  cyt_to_nuc[[filename]] = df[pval<0.05,unique_id]
  obj = readRDS(file.path('data/processed/nuc_to_cyt', i))
  scores <- as.data.frame(as.table(obj[["Prob_P2_P3_P4"]]))
  pval <- as.data.frame(as.table(obj[["PvalP2_P3_P4"]]))
  colnames(scores) = c("source", "target", "lr_pair", "score")
  colnames(pval) = c("source", "target", "lr_pair", "pval")
  df = setDT(inner_join(scores, pval, by = intersect(colnames(scores), colnames(pval))))
  df$unique_id = paste0(df$lr_pair, "_", df$source, "_", df$target)
  nuc_to_cyt[[filename]] = df[pval<0.05,unique_id]
}

# add cyt-nuc to the venn data
for (i in names(venn_data)){
  venn_data[[i]][['C-->N']] = cyt_to_nuc[[i]]
  venn_data[[i]][['N-->C']] = nuc_to_cyt[[i]]
}

names(venn_data) = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
upset_plots = list()
interactions = c(10, 11, 4, 17, 21, 15, 8, 7, 16)
save_dir = "data/results/Final_Figures/Supplementary_figures/upset_plots"
for (i in 1:length(names(venn_data))){
  filename = file.path(save_dir, paste0(names(venn_data)[[i]], ".pdf"))
  pdf(filename)
  list = venn_data[[names(venn_data)[i]]]
  venn = Venn(list)
  p = plot_upset(venn, min_size = 1, nintersects = interactions[i], 
                 top.bar.color = "#3C3B6E", 
                 sets.bar.color = "#B22234",
                 intersection.matrix.color = "darkgreen",
                 top.bar.numbers.size = 6,
                 relative_height = 4,
                 relative_width = 0.4
                 )
  
  p[["plotlist"]][[2]] = p[["plotlist"]][[2]] + ggtitle(names(venn_data)[i]) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 16),
      strip.text = element_text(size = 16)
    )
  p[["plotlist"]][[3]] = p[["plotlist"]][[3]] + theme(
      plot.title = element_text(size = 16, face = "bold"),
      axis.text = element_text(size = 16, angle = 45),
      axis.title = element_text(size = 16),
      strip.text = element_text(size = 16)
    )
  
  p[["plotlist"]][[1]] = p[["plotlist"]][[1]] + theme(
      plot.title = element_text(size = 16, face = "bold"),
      axis.text = element_text(size = 20),
      axis.title = element_text(size = 16),
      strip.text = element_text(size = 16)
    )
  
  print(p)
  dev.off()
}

```

## S3. Distribution of communication scores in each dataset across the regions 
```{r}
#| message: false
#| include: true
#| warning: false
data = fread("data/processed/training_data_071025.csv")
box_data <- data[(nuc_pval < 0.05) | (cyt_pval < 0.05) | (cell_pval < 0.05), .(cell_score, cell_P1, cyt_score, cyt_P1, nuc_score, nuc_P1, sample)]
box_data <- box_data %>%
  mutate(compartment = case_when(
    cyt_score > 0 & nuc_score > 0 & cell_score > 0 ~ "cell+cyt+nuc",
    cyt_score == 0 & nuc_score == 0 & cell_score > 0 ~ "cell",
    cyt_score > 0 & nuc_score > 0 & cell_score == 0 ~ "cyt+nuc",
    cyt_score == 0 & nuc_score > 0 & cell_score > 0 ~ "cell+nuc",
    cyt_score == 0 & nuc_score > 0 & cell_score == 0 ~ "nuc",
    cyt_score > 0 & nuc_score == 0 & cell_score == 0 ~ "cyt",
    cyt_score > 0 & nuc_score == 0 & cell_score > 0 ~ "cell+cyt",
    TRUE ~ "none"   # fallback in case none of the conditions match
  ))
box_data$sample = factor(box_data$sample, unique(box_data$sample), 
                         labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin"))
colnames(box_data) = c('cell', 'cell_P1', 'cytoplasm', 'cyt_P1', 'nucleus', 'nuc_P1', 'sample', 'compartment')
cell_cyt_data <- box_data[compartment == "cell+cyt"]
cell_nuc_data <- box_data[compartment == "cell+nuc"]
cell_cyt_nuc_data <- box_data[compartment == "cell+cyt+nuc"]

cell_cyt_long <- melt(cell_cyt_data[,.(cell, cytoplasm, sample)],measure.vars = c("cell", "cytoplasm"),variable.name = "Compartment",value.name = "Score")
cell_nuc_long <- melt(cell_nuc_data[,.(cell, nucleus, sample)],measure.vars = c("cell", "nucleus"),variable.name = "Compartment",value.name = "Score")
cell_cyt_nuc_long <- melt(cell_cyt_nuc_data[,.(cell, cytoplasm, nucleus, sample)],measure.vars = c("cell", "cytoplasm", "nucleus"),variable.name = "Compartment",value.name = "Score")
setDT(cell_cyt_long)
setDT(cell_nuc_long)
setDT(cell_cyt_nuc_long)
cell_cyt_long[, log_Score := log10(Score + 1e-10)] 
cell_nuc_long[, log_Score := log10(Score + 1e-10)] 
cell_cyt_nuc_long[, log_Score := log10(Score + 1e-10)] 

plotdir = "data/results/Final_Figures/Supplementary_figures/compare_ditributions/"
ggboxplot(
  cell_cyt_long, 
  x = "Compartment", 
  y = "log_Score",
  color = "Compartment", 
  palette = c("#b089c0", "#97c9c8"),  # colorblind-friendly
  add = "jitter", 
  add.params = list(alpha = 1, size = 0.5)
) +
  facet_wrap(~ sample, ncol = 3) +
  stat_compare_means(
    comparisons = list(
      c("cell", "cytoplasm")
    ),
    method = "wilcox.test",
    size = 4
  ) +
  labs(
    y = expression(log[10]~"Communication scores"),
    x = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1), # box around each facet
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 30, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.position = "none"
  )
ggsave(file.path(plotdir, 'cellvscyt.pdf'), height = 11)

ggboxplot(
  cell_nuc_long, 
  x = "Compartment", 
  y = "log_Score",
  color = "Compartment", 
  palette = c("#b089c0", "#ea9b8e"),  # colorblind-friendly
  add = "jitter", 
  add.params = list(alpha = 1, size = 0.5)
) +
  facet_wrap(~ sample, ncol = 3) +
  stat_compare_means(
    comparisons = list(
      c("cell", "nucleus")
    ),
    method = "wilcox.test",
    size = 4
  ) +
  labs(
    y = expression(log[10]~"Communication scores"),
    x = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1), # box around each facet
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 30, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.position = "none"
  )

ggsave(file.path(plotdir, 'cellvsnuc.pdf'), height = 11)


ggboxplot(
  cell_cyt_nuc_long, 
  x = "Compartment", 
  y = "log_Score",
  color = "black", 
  fill = "Compartment", 
  palette = c("#b089c0", "#97c9c8", "#ea9b8e"),  # colorblind-friendly
  add = "jitter", 
  add.params = list(alpha = 0.2, size = 0.3, width = 0.1)
) +
  facet_wrap(~ sample,ncol = 3) +
  stat_compare_means(
    comparisons = list(
      c("cell", "cytoplasm"),
      c("cell", "nucleus"),
      c("cytoplasm", "nucleus")
    ),
    method = "wilcox.test",
    size = 4
  ) +
  labs(
    y = expression(log[10]~"Communication scores"),
    x = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "lightgray", fill = NA, size = 1), # box around each facet
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.position = "none"
  )
ggsave(file.path(plotdir, 'cellvscytvsnuc.pdf'), height = 15)
```

## S4. Performance metrics of the model.

```{r}
#| message: false
#| include: true
#| warning: false
ddir = "data/processed/plots_metrics"
xgb_voting = fread(file.path(ddir, "comparison_baselearners/xgboost_voting_metrics.csv"))
xgb_voting = xgb_voting[, c(1,2:7,11,13,14,17,20)]
class_metrics = melt(xgb_voting, id.vars = c('Dataset'), measure.vars = 2:12)

class_metrics$variable <- factor(class_metrics$variable,
  levels = unique(class_metrics$variable),
  labels = c("F1 (Class 0)", "Precision (Class 0)", "Recall (Class 0)",
  "F1 (Class 1)", "Precision (Class 1)", "Recall (Class 1)",
  "Accuracy", "Macro F1", "Macro Precision",
  "Cytoplasm RMSE", "Nucleus RMSE")
)
colors = c("#1f77b4","#3a99d9","#5cb0e0","#7fc7e6","#a3d5eb","#c7e3f0","#fdb462","#fec978","#fee0a3","#9467bd","#b087d0")
ggplot(class_metrics, aes(x = variable, y = value)) +
  geom_violin(width = 1, aes(fill = variable)) +
  geom_jitter(width = 0.1) + 
  scale_fill_manual(values = colors) + 
  geom_boxplot(color = "black", fill = "white", width = 0.05) +
  theme_classic(base_size = 20) + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = ""
    ) +
  labs( x = "", y = "Value")
ggsave("data/results/Final_Figures/Supplementary_figures/best_model_metrics/class_metrics.pdf", width = 12)

```


## S5. Classification and regression  metrics of the different base learners.
```{r}
#| message: false
#| include: true
#| warning: false
ddir = "data/processed/plots_metrics"
xgb = fread(file.path(ddir, "comparison_baselearners/xgboost_metrics.csv"))
rf = fread(file.path(ddir, "comparison_baselearners/rf_metrics.csv"))
xgb_rf = fread(file.path(ddir, "comparison_baselearners/xgboost_rf_metrics.csv"))
xgb_voting = fread(file.path(ddir, "comparison_baselearners/xgboost_voting_metrics.csv"))

xgb = xgb[, model:= "xgb"]
rf = rf[, model:= "rf"]
xgb_rf = xgb_rf[, model:= "xgb_rf"]
xgb_voting = xgb_voting[, model:= "xgb_voting"]

dt = rbind(xgb, rf, xgb_rf, xgb_voting)

heatmap_df = melt(dt, id.vars= c('model', 'Dataset'), measure.vars= c(12,15,16,18,19,21)) 
heatmap_df$variable <- factor(heatmap_df$variable,
  levels = c("auc","macro_recall","cyt_R²","nuc_R²",
             # "cyt_RMSE","nuc_RMSE",
             "cyt_NRMSE","nuc_NRMSE","geometric_mean"),
  labels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "Geometric Mean")
)

heatmap_df$Dataset <- factor(heatmap_df$Dataset,
  levels = unique(heatmap_df$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

heatmap_df$model <- factor(heatmap_df$model,
  levels = unique(heatmap_df$model),
  labels = c("Model 1", "Model 2", "Model 3", "Model 4")
)

# plot the heatmap
ggplot(heatmap_df, aes(x = variable, y = Dataset, fill = value)) +
  geom_tile(color = "black", linewidth = 0.1) + 
  scale_fill_gradientn (colours = c("white", "#cdedea", "#20b3a5")) + 
  labs(x = "", y = "Dataset") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  ) +
  coord_fixed() + 
  facet_wrap(~model, nrow = 1)

ggsave("data/results/Final_Figures/Supplementary_figures/heatmap_comparisons/heatmap_baselearners.pdf", width = 10)
```


## S6. Classification and regression metrics of models with different feature engineering approaches
```{r}
#| message: false
#| include: true
#| warning: false
ddir = "data/processed/plots_metrics"
a = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_ohe_not_separated_model2.csv"))
b = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_catboost_not_separated_model3.csv"))
c = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_combined_model4.csv"))
d = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_model1.csv"))

a = a[, model:= "Model A"]
b = b[, model:= "Model B"]
c = c[, model:= "Model C"]
d = d[, model:= "Model D"]

dt = rbind(a, b, c, d)

heatmap_df = melt(dt, id.vars= c('model', 'Dataset'), measure.vars= c(12,15,16,18,19,21)) 
heatmap_df$variable <- factor(heatmap_df$variable,
  levels = c("auc","macro_recall","cyt_R²","nuc_R²",
             # "cyt_RMSE","nuc_RMSE",
             "cyt_NRMSE","nuc_NRMSE","geometric_mean"),
  labels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "Geometric Mean")
)

heatmap_df$Dataset <- factor(heatmap_df$Dataset,
  levels = unique(heatmap_df$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

heatmap_df$model <- factor(heatmap_df$model,
  levels = unique(heatmap_df$model),
  labels = c("Model A", "Model B", "Model C", "Model D")
)

# plot the heatmap
ggplot(heatmap_df, aes(x = variable, y = Dataset, fill = value)) +
  geom_tile(color = "black", linewidth = 0.1) + 
  scale_fill_gradientn (colours = c("white", "#cdc2ea", "#b585d1")) + 
  labs(x = "", y = "Dataset") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  ) +
  coord_fixed() + 
  facet_wrap(~model, nrow = 1)
ggsave("data/results/Final_Figures/Supplementary_figures/heatmap_comparisons/heatmap_fe_regression.pdf", width = 10)

```

## S7. Number of LR pairs correctly classified when training the lung cancer dataset on 255 combinations of the 8 training datasets (raw counts on the left, percentages on the right)

```{r}
#| message: false
#| include: true
#| warning: false
dir = "data/processed/training_256/correct"
files = list.files(dir)
file_list = list()

for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(dir, i))
  x = nrow(dat)/255
  dat$id = rep(1:x, 255)
  data = dcast(dat, id ~ group, fun.aggregate = NULL, value.var = "correct")
  data$sample = sub("classification_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = data
}

prediction = setDT(do.call(rbind, file_list))
rownames(prediction) = 1:nrow(prediction)
col = prediction[, lapply(.SD, sum), by = sample, .SDcols = c(2:256)]
percent <- prediction[, lapply(.SD, function(x) sum(x) *100 / .N), 
                      by = sample, .SDcols = 2:256]

colsums = melt(col, id.vars = "sample")
percent = melt(percent, id.vars = "sample")
# p1 = ggplot(colsums, aes(x = value)) +
#   geom_histogram(binwidth = 1, fill = "#bebadb", color = "#bebadb") +
#   theme_classic(base_size = 12) +
#   labs(x = "Number of Correct classifications", y = "Number of combinations") +
#   facet_wrap(~sample, scales = 'free')
# 
# p2 = ggplot(colsums, aes(x = "", y = value)) +
#   geom_violin(fill = "#fcdeed", width = 0.4) + 
#   geom_boxplot(color = "gray50", fill = "white", width = 0.05) + 
#   geom_jitter(width = 0.3, alpha = 0.1, size = 0.5) + 
#   theme_classic(base_size = 12) +
#   theme(
#     axis.text.x = element_blank(),
#     axis.ticks.x = element_blank()
#   ) + 
#   labs(x = "", y = "LR pairs - classfied correct") +
#   facet_wrap(~sample, scales = 'free')

# p3 <- ggplot(percent, aes(x = value)) +
#   geom_histogram(binwidth = 1, fill = "#8ed3c7", color = "black") +
#   theme_classic(base_size = 12) +
#   labs(x = "% of Correct classifications", y = "Number of combinations") +
#   facet_wrap(~sample, scales = 'free')
# 
# p4 = ggplot(percent, aes(x = "", y = value)) +
#   geom_violin(fill = "#ccebc5", width = 0.4) + 
#   geom_boxplot(color = "black", fill = "white", width = 0.05) + 
#   geom_jitter(width = 0.3, alpha = 0.1, size = 1) + 
#   theme_classic(base_size = 12) +
#   theme(
#     axis.text.x = element_blank(),
#     axis.ticks.x = element_blank()
#   )  + 
#   labs(x = "", y = "LR pairs - classfied correct (%)") + 
#   facet_wrap(~sample, scales = 'free')

# ggarrange (p1, p2, common.legend = T)
# ggsave("data/results/Final_Figures/Supplementary_figures/counts_plots/counts_plots_combinations.pdf", height = 10, width = 10)
```

## S8.  Number of times each LR pair was classified correctly 
```{r}
#| message: false
#| include: true
#| warning: false

row <- prediction[, .(value = rowSums(.SD)), by = sample, .SDcols = 2:256]
percent <- prediction[, .(value = rowSums(.SD) * 100/ 256), 
                      by = sample, .SDcols = 2:256]

# p1 = ggplot(row, aes(x = value)) +
#   geom_histogram(binwidth = 5, fill = "#fb8073", color = "black") +
#   theme_classic(base_size = 12) +
#   labs(x = "Number of combinations", y = "count of LR") +
#   facet_wrap(~sample, scales = 'free')

row$sample <- factor(row$sample,
  levels = unique(row$sample),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)


colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )
ggplot(row, aes(x = value, y = sample, fill = sample)) +
  geom_density_ridges(alpha = 0.7, color = "black", scale = 1) +
  theme_classic(base_size = 16) + xlim(0,255) + 
  labs(x = "Number of correctly \n classified LR pairs", y = "Datasets") +
  scale_fill_manual(values = colors) + # customize colors
  theme(legend.position = "none") 

# 
# ggplot(percent, aes(x = value, y = sample, fill = sample)) +
#   geom_density_ridges(alpha = 0.7, color = "black", scale = 1) +
#   theme_classic(base_size = 12) +
#   labs(x = "Number of combinations", y = "Sample")
# p2 = ggplot(row, aes(x = "", y = value)) +
#   geom_violin(fill = "#efaba3", width = 0.6) + 
#   geom_boxplot(color = "black", fill = "white", width = 0.05) + 
#   geom_jitter(width = 0.3, alpha = 0.1, size = 1) + 
#   theme_classic(base_size = 12) +
#   theme(
#     axis.text.x = element_blank(),
#     axis.ticks.x = element_blank()
#   )  + 
#   labs(x = "", y = "Number of combinations") +
#   facet_wrap(~sample, scales = 'free')

# p3 <- ggplot(percent, aes(x = value)) +
#   geom_histogram(binwidth = 5, fill = "#fdb461", color = "black") +
#   theme_classic(base_size = 12) +
#   labs(x = "% of combinations", y = "count of LR") +
#   facet_wrap(~sample, scales = 'free')
# 
# p4 = ggplot(percent, aes(x = "", y = value)) +
#   geom_violin(fill = "#ffffb4", width = 0.6) + 
#   geom_boxplot(color = "black", fill = "white", width = 0.05) + 
#   geom_jitter(width = 0.3, alpha = 0.1, size = 1) + 
#   theme_classic(base_size = 12) +
#   theme(
#     axis.text.x = element_blank(),
#     axis.ticks.x = element_blank()
#   )  + 
#   labs(x = "", y = "% of combinations")+
#   facet_wrap(~sample, scales = 'free')

# ggarrange (p1, p2, p3, p4)
ggsave("data/results/Final_Figures/Supplementary_figures/counts_plots/lr_counts_plot.pdf")



# other metrics 

class_dir <- "data/processed/scores_training/classification_metrics/"

files = list.files(class_dir)
file_list = list()

for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(class_dir, i))
  dat$sample = sub("classification_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}

# join with the metrics with perclass metrics 
metrics = do.call(rbind, file_list)
perclass_metrics = fread("data/processed/scores_training/final_classification_metrics.csv")
colnames(perclass_metrics)[11] = "group"
perclass_metrics$sample = sub("classification_spe_list_(.*)_cellchat\\.csv$", "\\1", perclass_metrics$dataset)
perclass_metrics = perclass_metrics[,-c(9,10,12)]
metrics = metrics[perclass_metrics, on = .(sample, group)]
metrics_further = metrics


# Regression metrics 
cyt_reg_dir <- "data/processed/scores_training/cyt_metrics"
nuc_reg_dir <- "data/processed/scores_training/nuc_metrics"
files = list.files(cyt_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(cyt_reg_dir, i))
  dat$sample = sub("cyt_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
cyt_metrics = do.call(rbind, file_list)
colnames(cyt_metrics)[1:3] = paste0("cyt_",colnames(cyt_metrics)[1:3])

files = list.files(nuc_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(nuc_reg_dir, i))
  dat$sample = sub("nuc_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
nuc_metrics = do.call(rbind, file_list)
colnames(nuc_metrics)[1:3] = paste0("nuc_",colnames(nuc_metrics)[1:3])

regression_metrics = cyt_metrics[nuc_metrics, on = .(group, sample)]
regression_metrics$`cyt_R²` = ifelse(regression_metrics$`cyt_R²`<0, 0, regression_metrics$`cyt_R²`)
regression_metrics$`nuc_R²` = ifelse(regression_metrics$`nuc_R²`<0, 0, regression_metrics$`nuc_R²`)
regression_metrics_further = regression_metrics

regression_metrics = melt(regression_metrics, id.vars = c('sample', "group"), measure.vars = c(1:3, 6:8))
regression_metrics$facets <- with(regression_metrics,
  ifelse(variable %in% levels(regression_metrics$variable)[c(1,4)], "R²",
  ifelse(variable %in% levels(regression_metrics$variable)[c(3,6)], "NRMSE", 'RMSE'))
)

regression_metrics$facets = factor(regression_metrics$facets, levels = c("R²", "NRMSE", 'RMSE'))



metrics = metrics_further[regression_metrics_further, on = .(group, sample)]
combo_metrics = metrics[, c(9,7,8,12,10,11,1,3,13,17,19)]
dt = melt(combo_metrics)
dt$variable <- factor(dt$variable,
  levels = unique(dt$variable),
  labels = c("F1 (Class 0)", "Precision (Class 0)", "Recall (Class 0)",
  "F1 (Class 1)", "Precision (Class 1)", "Recall (Class 1)",
  "Accuracy", "Macro F1", "Macro Precision",
  "Cytoplasm RMSE", "Nucleus RMSE")
)

colors = c("#1f77b4","#3a99d9","#5cb0e0","#7fc7e6","#a3d5eb","#c7e3f0","#fdb462","#fec978","#fee0a3","#9467bd","#b087d0")
ggplot(dt, aes(x = variable, y = value)) +
  geom_violin(width = 1, aes(fill = variable)) + 
  scale_fill_manual(values = colors) + 
  geom_boxplot(color = "black", fill = "white", width = 0.2) +
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 18),
      legend.position = "",
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.text.x = element_text(angle = 45,  hjust = 1, vjust = 1)
    ) +
  labs( x = "", y = "Value") + coord_cartesian(ylim=c(0,1)) 
ggsave("data/results/Final_Figures/Supplementary_figures/counts_plots/combo_metrics.pdf", width = 8)

# split the composite metric alone
combo_metrics = metrics[, c(4,5,6,14,15,18,17,20)]
combo_metrics[ , geometric_mean := abs(roc_auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]


combo_metrics = combo_metrics[,c(3,9)]


combo_metrics$sample <- factor(combo_metrics$sample,
  levels = unique(combo_metrics$sample),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)


colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )
ggplot(combo_metrics, aes(x = geometric_mean, y = sample, fill = sample)) +
  geom_density_ridges(alpha = 0.7, color = "black", scale = 1) +
  theme_classic(base_size = 20) + xlim(0,1) + 
  labs(x = "Composite metric", y = "Datasets") +
  scale_fill_manual(values = colors) + # customize colors
  theme(legend.position = "none") 


ggsave("data/results/Final_Figures/Supplementary_figures/counts_plots/composite_metric.pdf")
```

## S9. Are spatial features important?
```{r}
metrics = fread("data/processed/metrics_all_combinations_trainings_10032025.csv")
df_long <- metrics[,c(16,24,1,6,3,8,9,12)]
colnames(df_long) = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "condition",
             "combo_size")
df_long <- df_long %>%
   pivot_longer(cols = colnames(df_long)[c(1:6)],
                names_to = "metric", values_to = "value")

df_summary <- df_long %>%
  group_by(combo_size, condition, metric) %>%
  summarise(mean_val = mean(value, na.rm = TRUE),
            sd_val = sd(value, na.rm = TRUE)/sqrt(n()), .groups = "drop")

df_summary$metric = factor(df_summary$metric,
                           levels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "condition",
             "combo_size")
                             )

colors = c('with_spatial' = "#b715b7", "without_spatial" = "#14bf81")
ggplot(df_summary, aes(x = combo_size, y = mean_val, color = condition, group = condition)) +
  geom_line() +
  geom_point() +
  scale_color_manual(
  name   = "Model type",
  values = colors,
  labels = c(
    with_spatial    = "With spatial information",
    without_spatial = "Without spatial information"
  )
) + 
  geom_errorbar(aes(ymin = mean_val - sd_val, ymax = mean_val + sd_val), width = 0.2) +
  # facet_wrap(~metric, scales = "free_y", ncol = 4) +
  labs( x = "Number of training datasets", y = "Value") + 
  facet_wrap(~metric, scales = "free",ncol =3) + theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text.y = element_text(size = 12),
    strip.background = element_blank(), strip.placement = "outside"
  ) 

ggsave("data/results/Final_Figures/Supplementary_figures/best_model_metrics/spatial_features.pdf", height = 10)
```
 
## S9. Correlation of region specific scores with the cell score with and without spatial feature, B.  Distribution of the pairwise difference between the various model trainings on different combination of datasets given a candidate dataset (lung cancer dataset included) C. Network plot (interaction sum) of the LR scores on the single cell data in the  predicted cytoplasm and nucleus

```{r}
df =  fread("data/processed/training_data_071025.csv")

# compute correlation per sample
corr_df <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_P1, cyt_P1, use="complete.obs", method="pearson"),
            .groups="drop")

corr_df$sample <- c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
# plot
p1 = ggplot(corr_df, aes(x=sample, y=corr, fill=sample)) +
  geom_col() +
  coord_flip() +
  labs(title="Correlation between cell_P1 and cyt_P1 \n(without spatial)",
       y="Pearson correlation", x="Sample") +
  theme(legend.position = "")


corr_df <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_score, cyt_score, use="complete.obs", method="pearson"),
            .groups="drop")

corr_df$sample <- c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")# plot
 
p2 = ggplot(corr_df, aes(x=sample, y=corr, fill=sample)) +
  geom_col() +
  coord_flip() +
  labs(title="Correlation between cell_score and cyt_score \n(with spatial)",
       y="Pearson correlation", x="Sample") +
  theme(legend.position = "")

ggarrange(p1,p2)



# compute correlation per sample
corr_df <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_P1, nuc_P1, use="complete.obs", method="pearson"),
            .groups="drop")


corr_df$sample <- c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
# plot
p3 = ggplot(corr_df, aes(x=sample, y=corr, fill=sample)) +
  geom_col() +
  coord_flip() +
  labs(title="Correlation between cell_P1 and nuc_P1 \n(without spatial)",
       y="Pearson correlation", x="Sample") +
  theme(legend.position = "")


corr_df <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_score, nuc_score, use="complete.obs", method="pearson"),
            .groups="drop")

corr_df$sample <- c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")# plot
 
p4 = ggplot(corr_df, aes(x=sample, y=corr, fill=sample)) +
  geom_col() +
  coord_flip() +
  labs(title="Correlation between cell_score and nuc_score \n(with spatial)",
       y="Pearson correlation", x="Sample") +
  theme(legend.position = "")

ggarrange(p1,p2,p3,p4)



# Cytoplasmic comparison
corr_no_spatial <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_P1, cyt_P1, use="complete.obs", method="pearson"),
            .groups="drop") %>%
  mutate(type = "Without spatial")
corr_no_spatial$sample <- c("Brain (AD)","Brain (GL)","Breast Cancer","Lung",
                     "Lung (High-Res)","Lymph Node","Ovarian","Pancreas","Skin")
corr_spatial <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_score, cyt_score, use="complete.obs", method="pearson"),
            .groups="drop") %>%
  mutate(type = "With spatial")
corr_spatial$sample <- c("Brain (AD)","Brain (GL)","Breast Cancer","Lung",
                     "Lung (High-Res)","Lymph Node","Ovarian","Pancreas","Skin")
cyt_corr <- bind_rows(corr_no_spatial, corr_spatial)


p_cyt <- ggplot(cyt_corr, aes(x=sample, y=corr, fill=type)) +
  geom_col(position="dodge") +
  labs(
       y="Pearson correlation", x="") +
  scale_fill_manual(values = c("#D2665A", "#F2B28C")) + 
  theme_classic(base_size = 20) +
  theme(legend.title=element_blank(),
        axis.text.x = element_text(angle = 45, size = 12, hjust = 1, vjust = 1))

# Nuclear comparison
corr_no_spatial <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_P1, nuc_P1, use="complete.obs", method="pearson"),
            .groups="drop") %>%
  mutate(type = "Without spatial")
corr_no_spatial$sample <- c("Brain (AD)","Brain (GL)","Breast Cancer","Lung",
                     "Lung (High-Res)","Lymph Node","Ovarian","Pancreas","Skin")
corr_spatial <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_score, nuc_score, use="complete.obs", method="pearson"),
            .groups="drop") %>%
  mutate(type = "With spatial")
corr_spatial$sample <- c("Brain (AD)","Brain (GL)","Breast Cancer","Lung",
                     "Lung (High-Res)","Lymph Node","Ovarian","Pancreas","Skin")
nuc_corr <- bind_rows(corr_no_spatial, corr_spatial)

p_nuc <- ggplot(nuc_corr, aes(x=sample, y=corr, fill=type)) +
  geom_col(position="dodge") +
  scale_fill_manual(values = c("#D2665A", "#F2B28C"))+
  labs(
       y="", x="") +
  theme_classic(base_size = 20) +
  theme(legend.title=element_blank(),
        axis.text.x = element_text(angle = 45, size = 12, hjust = 1, vjust = 1))

ggarrange(p_cyt, p_nuc, ncol=2, common.legend=TRUE, legend="bottom")

ggsave("data/results/Final_Figures/Supplementary_figures/single_cell/correlation.pdf", width = 10)

# scatter plot  join the nuc and cytoplasm 
colnames(nuc_corr)[2] = "Nucleus"
colnames(cyt_corr)[2] = "Cytoplasm"
# inner join
setDT(nuc_corr)
setDT(cyt_corr)
data = cyt_corr[nuc_corr, on = c('sample', 'type')]
data_plot = melt(data, id.vars = c(1,3), measure.vars = c(2,4), variable.name = "Region")
wide <- dcast(
  data_plot,
  sample + variable ~ type,
  value.var = "value"
)

colnames(wide)[1:2] = c("Dataset","Region") 
ggplot(wide, aes(x=`With spatial`, y=`Without spatial`, color=Region, shape = Dataset)) + geom_point() +
  scale_color_manual(values = c("#D2665A", "#F2B28C")) + 
  scale_shape_manual(values = 1:9)

ggsave("data/results/Final_Figures/Supplementary_figures/single_cell/correlation.pdf")


metrics = fread("data/processed/metrics_all_combinations_trainings_10032025.csv")
metrics[ , geometric_mean := (roc_auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]


df_long <- metrics[,c(9,28,10,11, 12)]

spatial = df_long[condition == "with_spatial",]
no_spatial = df_long[condition == "without_spatial",]
#check
identical(spatial[,3:5], no_spatial[,3:5])

df_long = spatial[,3:5]
df_long$difference = spatial$geometric_mean - no_spatial$geometric_mean
df_long = na.omit(df_long)

df_long$test_name <- factor(df_long$test_name,
  levels = unique(df_long$test_name),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

ggplot(df_long, aes(x = test_name, y = difference, fill = test_name)) +
  geom_boxplot(outliers = TRUE) +
  scale_fill_manual(values = colors) + 
  # facet_wrap(~metric, scales = "free_y", ncol = 4) +
  labs( x = "", y = "Difference in geometric means") + 
  theme_classic(base_size = 20) +
  theme(
    axis.text.x = element_text(angle = 45, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 18)
  )
ggsave("data/results/Final_Figures/Supplementary_figures/single_cell/boxplot_difference.pdf", width =8)
```

## Exploring if the incorrect classifications are in low number of datasets


```{r}
dir = "data/processed/scores_training/classification_raw"
files = list.files(dir)[1:9]
file_list = list()

for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(dir, i))
  dat$correct = ifelse(dat$pred == dat$true, 1, 0)
  x = nrow(dat)/255
  dat$id = rep(1:x, 255)
  data = dcast(dat, id ~ group, fun.aggregate = NULL, value.var = "correct")
  data$sample = sub("classification_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = data
}

prediction = setDT(do.call(rbind, file_list))
rownames(prediction) = 1:nrow(prediction)
col = prediction[, lapply(.SD, sum), by = sample, .SDcols = c(2:256)]
percent <- prediction[, lapply(.SD, function(x) sum(x) *100 / .N), 
                      by = sample, .SDcols = 2:256]

counts <- prediction[, .N, by = sample]
counts$sample <- factor(counts$sample,
  levels = unique(counts$sample)
)

data = fread("data/processed/combinations_iters.csv")
data$sample = gsub("^spe_list_|_cellchat$", "",data$test_name)
column = melt(col, id.vars = 'sample', measure.vars = 2:256, variable.name = 'combo_id')
setDT(column)
setkey(column, combo_id, sample)
setkey(data, combo_id, sample)
merged_dt <- data[column]
merged_dt = merged_dt[counts, on = 'sample']
merged_dt[,percent:= value/N]
summary_df <- merged_dt %>%
  group_by(combo_size, sample) %>%
  summarise(
    mean_percent = mean(percent),
    median_percent = median(percent)
  ) %>%
  ungroup()

summary_df$combo_size = as.factor(summary_df$combo_size)
ggplot(summary_df, aes(x = combo_size, y = median_percent, color = sample, group = sample)) +
  geom_bar(size = 2) +
  labs(
    x = "Number of Combinations",
    y = "Mean Percent Correct",
    title = "Mean Percent Correct by Combo Size per Sample",
    color = "Sample"
  ) +
  theme_classic()

ggplot(summary_df, aes(x = sample, y = median_percent, fill = combo_size)) +
  geom_col(position = position_dodge(width = 0.8)) +
  labs(
    x = "Number of Combinations",
    y = "Median Percent Correct",
    title = "Median Percent Correct by Combo Size per Sample",
    fill = "Sample"
  ) +
  theme_classic()

```

