---
title: "Figure 4"
author: "rshreej"
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  html:
    toc: true
    toc-depth: 2
    toc-expand: true
    number-sections: true
    theme: united
    highlight: tango
    code-fold: true
    code-tools: true
    embed-resources: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r}
knitr::opts_knit$set(root.dir = here::here())
```

```{r}
#| message: false
#| include: false
#| warning: false
library(data.table)
library(circlize)
library(ggvenn)
library(ggridges)
library(cowplot)
library(pheatmap)
library(ComplexHeatmap)
library(fgsea)
library(msigdbr)
library(ggpubr)
library(clusterProfiler)
library(enrichplot) 
library(patchwork)
library(ggbiplot)
library(gridGraphics)
library(gridExtra)
library(igraph)
library(RColorBrewer)
library(ggvenn)
library(ggVennDiagram)
library(magick)
library(ragg)
library(tidyr)
library(dplyr)
library(ComplexHeatmap)
library(tidyverse)
```

```{r}
#| message: false
#| include: true
#| warning: false
theme_set(
  theme_classic(base_size = 14) +
    theme(
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 12, face = "bold"),
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
    )
)
```

#### Supplementary Figure 1: 

Hierarchical clustering dendrogram of Gene Ontology Cellular Component
terms for significantly interacting (p-value \< 0.05) ligands and
receptors detected across subcellular regions in all nine spatial
transcriptomics datasets. Dendrogram shading indicates broad ontology
categories; node color represents FDR-adjusted p-values of enriched
terms, and node size reflects the number of genes in each GO term.
Panels A and C represent clustering of significantly interacting ligands
detected in cytoplasm and nucleus respectively and panels B and D
represent clustering of significantly interacting receptors detected in
cytoplasm and nucleus respectively.

```{r}
#| message: false
#| warning: false
#| include: true
#| fig-width: 30
#| fig-height: 30 
#| label: gsea
# Gene set enrichment analysis ********************************************
data = fread("data/training_data/training_data.csv")
theme_set(theme_classic(base_size = 14))
# download the go:cc onotology terms 
gs_cc <- setDT(msigdbr(species = "Homo sapiens", category="C5", subcategory="GO:CC"))
TERM2GENE  <- gs_cc[,.(term = gs_name, gene = gene_symbol)]
unique_genes_df <- read_csv("data/training_data/raw_data.csv")
universe_genes <- c(unique_genes_df$ligand, unique_genes_df$receptor)%>%unique()

run_ora <- function(detected_genes, universe_genes, TERM2GENE,
                    minGS = 10, maxGS = 500, p_adj = "BH") {
  detected_genes <- intersect(unique(detected_genes), universe_genes)
  # if (length(detected_genes) < 5) {
  #   return(data.table())  
  # }
  ora <- enricher(gene          = detected_genes,
                  TERM2GENE     = TERM2GENE,
                  universe      = universe_genes,
                  pAdjustMethod = p_adj,
                  minGSSize     = minGS,
                  maxGSSize     = maxGS)
  as.data.table(ora@result)[order(p.adjust)]
}

options(enrichplot.colours = c("#1B9E77","#D81B60"))

plot_ora_tree <- function(ora_dt, top = 15, title = "") {
  if (!nrow(ora_dt)) return(invisible(NULL))

  # build enrichResult object
  ora_obj <- new("enrichResult", result = as.data.frame(ora_dt))

  # keep top categories (by adjusted p)
  if (!is.null(ora_obj@result) && nrow(ora_obj@result) > top) {
    keep_ids <- head(ora_obj@result[order(ora_obj@result$p.adjust), "ID"], top)
    ora_obj@result <- ora_obj@result[ora_obj@result$ID %in% keep_ids, , drop = FALSE]
  }

  # compute pairwise term similarity then treeplot
  # requires enrichplot >= 1.20 and DOSE loaded as a dep
  if (requireNamespace("enrichplot", quietly = TRUE)) {
    ora_sim <- enrichplot::pairwise_termsim(ora_obj)
    if ("treeplot" %in% getNamespaceExports("enrichplot")) {
      p <- enrichplot::treeplot(ora_sim, showCategory = top, fontsize = 10,
                                extend = 1,
                                label_format_tiplab = 30,
                                label_format_cladelab = 20,
                                leave_fontsize = 4,
                                clade_fontsize = 4,
                                offset =8,
                                legend_n = 0,
                                geneClusterPanel = "dotplot",
                                group_color = c("#f27042",
                                                "#fcda83", "#b62e48", 
                                                "#7ecba4", "#4965b0"
                                                )) + ggtitle(title)
      return(p)
    } else {
      # fallback: enrichment map if treeplot not available
      p <- enrichplot::emapplot(ora_sim, showCategory = top) + ggtitle(paste0(title, " (emap)"))
      return(p)
    }
  } else {
    stop("Package 'enrichplot' is required for tree/emap plots. Install with BiocManager::install('enrichplot').")
  }
}

# Suppose you have four enrichment results
padj_min <- min(c(res_lig_cyt$p.adjust,
                  res_lig_nuc$p.adjust,
                  res_rec_cyt$p.adjust,
                  res_rec_nuc$p.adjust))

padj_max <- max(c(res_lig_cyt$p.adjust,
                  res_lig_nuc$p.adjust,
                  res_rec_cyt$p.adjust,
                  res_rec_nuc$p.adjust))

gene_ratio_min <- 0
gene_ratio_max <- max(c(res_lig_cyt$Count,
                        res_lig_nuc$Count,
                        res_rec_cyt$Count,
                        res_rec_nuc$Count))


p1 = plot_ora_tree(res_lig_cyt, top = 17, title = "A. Ligands - Cytoplasm") +
  scale_color_gradientn(
    colors = c("#1B9E77", "#1B9E77","#D81B60", "#D81B60"),  # repeated colors
    values = scales::rescale(c(0, 0.05, 0.05, 1)),           # cutoff at 0.05
    limits = c(0, 1),                                        # full range
    name = "FDR-adjusted p-value"
  )+ 
      scale_size_continuous(limits = c(gene_ratio_min, gene_ratio_max), name = "Number of genes") +
      theme(legend.position = "none")
p2 = plot_ora_tree(res_lig_nuc, top = 17, title = "B. Ligands - Nucleus") + 
  scale_color_gradientn(
    colors = c("#1B9E77", "#1B9E77","#D81B60", "#D81B60"),  # repeated colors
    values = scales::rescale(c(0, 0.05, 0.05, 1)),           # cutoff at 0.05
    limits = c(0, 1),                                        # full range
    name = "FDR-adjusted p-value"
  ) + 
  scale_size_continuous(limits = c(gene_ratio_min, gene_ratio_max), name = "Number of genes") +
  theme(legend.position = "none")
p3 = plot_ora_tree(res_rec_cyt, top = 17, title = "C. Receptors - Cytoplasm") + 
  scale_color_gradientn(
    colors = c("#1B9E77", "#1B9E77","#D81B60", "#D81B60"),  # repeated colors
    values = scales::rescale(c(0, 0.05, 0.05, 1)),           # cutoff at 0.05
    limits = c(0, 1),                                        # full range
    name = "FDR-adjusted p-value"
  ) + 
  scale_size_continuous(limits = c(gene_ratio_min, gene_ratio_max), name = "Number of genes") +
  theme(legend.position = "none")
p4 = plot_ora_tree(res_rec_nuc, top = 17, title = "D. Receptors - Nucleus") + 
  scale_color_gradientn(
    colors = c("#1B9E77", "#1B9E77","#D81B60", "#D81B60"),  # repeated colors
    values = scales::rescale(c(0, 0.05, 0.05, 1)),           # cutoff at 0.05
    limits = c(0, 1),                                        # full range
    name = "FDR-adjusted p-value"
  ) + 
  scale_size_continuous(limits = c(gene_ratio_min, gene_ratio_max), name = "Number of genes") +
  theme(legend.position = "none")

combined <- (p1 | p2) / (p3 | p4) &
  theme(                       # applies to ALL subplots uniformly
    text = element_text(size = 20),
    axis.text.y = element_text(size = 0),
    axis.text.x = element_text(size = 0)
  )
combined
ggsave('data/results/Final_Figures/Final_figure1/overlap_test_tree.pdf', height = 15, width = 20)

p1 + theme(legend.position = "bottom")
ggsave('data/results/Final_Figures/Final_figure1/overlap_test_tree_legend.pdf', height = 15, width = 20)


```

#### Supplementary Figure 4: 

Principal component analysis (PCA) of CellChat-derived communication
score components across the whole cell (left), cytoplasm (middle), and
nucleus (right) for all the 9 subcellularly resolved spatial
transcriptomics datasets. Each point represents a ligand–receptor (LR)
pair and is colored by the P1 value, defined as the Hill function of LR
expression.

```{r}
#| message: false
#| include: true
#| warning: false
#| label: pca-plots
pc_df = fread("data/processed/pcs_df.csv")
# color by product of expression 
p1= ggplot(pc_df, aes(PC1, PC2, color = cell_expression)) + 
  geom_point(size = 1, alpha = 0.7) +
  theme_classic() +
  ylim(y = c(-3, 10)) + 
  xlim(-10,10) +
  scale_color_gradient(low="#fadc85",
                     high="#a31044", space ="Lab", trans = "log10",
    breaks = c(1e-7, 1e-5, 1e-3),
  labels = c(expression(10^-7), expression(10^-5), expression(10^-3))
)+
  labs(x = "PC1 (Explained variance: 43.7%)",
        y = "PC2 (Explained variance: 20%)"
       ) +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 20),
    axis.text = element_text(size = 20),
    axis.title = element_text(size = 20),
    panel.grid.minor = element_blank(),
    legend.title = element_text(size = 20)
  ) +
  guides(color=guide_colorbar(title="LR Expression(log10)"))

p2 = ggplot(pc_df, aes(PC1, PC2, color = cyt_expression)) + 
  geom_point(size = 1, alpha = 0.7) +
  theme_classic() +
  ylim(y = c(-3, 10)) + 
  xlim(-10,10) +
  scale_color_gradient(low="#fadc85",
                     high="#a31044", space ="Lab", trans = "log10") +
  labs(x = "PC1 (Explained variance: 43.7%)",
        y = "PC2 (Explained variance: 20%)"
       ) +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 20),
    axis.text = element_text(size = 20),
    axis.title = element_text(size = 20),
    panel.grid.minor = element_blank(),
    legend.title = element_text(size = 20)
  ) +
  guides(color=guide_colorbar(title="LR Expression(log10)"))

p3 = ggplot(pc_df, aes(PC1, PC2, color = nuc_expression)) + 
  geom_point(size = 1, alpha = 0.7) +
  theme_classic() +
  ylim(y = c(-3, 10)) + 
  xlim(-10,10) +
  scale_color_gradient(low="#fadc85",
                     high="#a31044", space ="Lab", trans = "log10") +
  labs(x = "PC1 (Explained variance: 43.7%)",
        y = "PC2 (Explained variance: 20%)"
       ) +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 20),
    axis.text = element_text(size = 20),
    axis.title = element_text(size = 20),
    panel.grid.minor = element_blank(),
    legend.title = element_text(size = 20)
  ) +
  guides(color=guide_colorbar(title="LR Expression(log10)"))
ggarrange(p1,p2,p3, ncol = 3, common.legend = TRUE, legend = "none")

ggsave("data/results/Final_Figures/Final_figure1/pca_plot_expr.pdf", width = 20)

legend <- get_legend(
  p1 +
    guides(color = guide_colorbar(
      title = "LR Expression (log10)",
      title.position = "bottom",
      title.hjust = 0.5,
      barwidth = 15,
      barheight = 1,
      direction = "horizontal"
    ))
)


ggsave("data/results/Final_Figures/Final_figure1/pca_legend.pdf", legend, width=4, height=1)
```

#### Supplementary Figure 2: 

Upset plots of the significantly communicating LR communicating between
each of the subcellular regions for all the 9 subcellularly resolved
spatial transcriptomics datasets. A. Brain (AD), B. Brain (GL), C.
Breast Cancer, D. Lung, E. Lung (High-Res), F. Lymph Node, G. Pancreas,
H. Ovarian, I. Skin. N - Nucleus, C - Cytoplasm, WC - whole cell, AD -
Alzheimer's Disease, GL - Glioblastoma

```{r}
#| message: false
#| include: true
#| warning: false
#| label: upset-plots
data <- fread("data/training_data/training_data.csv") 
data$unique_id = paste0(data$lr_pair, "_", data$source, "_", data$target)
venn_data = list()
plot_list = list()
for (i in unique(data$sample)){
  subset = data[sample == i,]
  data_list = list(
    `WC-->WC` = as.vector(subset[cell_pval <0.05 & cell_score >0,unique_id]),
    `C-->C` = as.vector(subset[cyt_pval <0.05 & cyt_score > 0,unique_id]),
    `N-->N` = as.vector(subset[nuc_pval <0.05& nuc_score > 0,unique_id])
  )
  venn_data[[i]] = data_list
}


files = list.files('data/processed/cyt_to_nuc') # cyt to nuc and nuc to cyt (run the bin/cytnuc_nuccyt.R)
cyt_to_nuc = list()
nuc_to_cyt = list()
for (i in files){
  filename = gsub(".rds", "", i)
  obj = readRDS(file.path('data/processed/cyt_to_nuc', i))
  scores <- as.data.frame(as.table(obj[["Prob_P2_P3_P4"]]))
  pval <- as.data.frame(as.table(obj[["PvalP2_P3_P4"]]))
  colnames(scores) = c("source", "target", "lr_pair", "score")
  colnames(pval) = c("source", "target", "lr_pair", "pval")
  df = setDT(inner_join(scores, pval, by = intersect(colnames(scores), colnames(pval))))
  df$unique_id = paste0(df$lr_pair, "_", df$source, "_", df$target)
  cyt_to_nuc[[filename]] = df[pval<0.05,unique_id]
  obj = readRDS(file.path('data/processed/nuc_to_cyt', i))
  scores <- as.data.frame(as.table(obj[["Prob_P2_P3_P4"]]))
  pval <- as.data.frame(as.table(obj[["PvalP2_P3_P4"]]))
  colnames(scores) = c("source", "target", "lr_pair", "score")
  colnames(pval) = c("source", "target", "lr_pair", "pval")
  df = setDT(inner_join(scores, pval, by = intersect(colnames(scores), colnames(pval))))
  df$unique_id = paste0(df$lr_pair, "_", df$source, "_", df$target)
  nuc_to_cyt[[filename]] = df[pval<0.05,unique_id]
}

# add cyt-nuc to the venn data
for (i in names(venn_data)){
  venn_data[[i]][['C-->N']] = cyt_to_nuc[[i]]
  venn_data[[i]][['N-->C']] = nuc_to_cyt[[i]]
}

names(venn_data) = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
upset_plots = list()
interactions = c(10, 11, 4, 17, 21, 15, 8, 7, 16)
save_dir = "data/results/Final_Figures/Supplementary_figures/upset_plots"
for (i in 1:length(names(venn_data))){
  filename = file.path(save_dir, paste0(names(venn_data)[[i]], ".pdf"))
  pdf(filename)
  list = venn_data[[names(venn_data)[i]]]
  venn = Venn(list)
  p = plot_upset(venn, min_size = 1, nintersects = interactions[i], 
                 top.bar.color = "#3C3B6E", 
                 sets.bar.color = "#B22234",
                 intersection.matrix.color = "darkgreen",
                 top.bar.numbers.size = 6,
                 relative_height = 4,
                 relative_width = 0.4
                 )
  
  p[["plotlist"]][[2]] = p[["plotlist"]][[2]] + ggtitle(names(venn_data)[i]) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 16),
      strip.text = element_text(size = 16)
    )
  p[["plotlist"]][[3]] = p[["plotlist"]][[3]] + theme(
      plot.title = element_text(size = 16, face = "bold"),
      axis.text = element_text(size = 16, angle = 45),
      axis.title = element_text(size = 16),
      strip.text = element_text(size = 16)
    )
  
  p[["plotlist"]][[1]] = p[["plotlist"]][[1]] + theme(
      plot.title = element_text(size = 16, face = "bold"),
      axis.text = element_text(size = 20),
      axis.title = element_text(size = 16),
      strip.text = element_text(size = 16)
    )
  
  print(p)
  dev.off()
}
```

#### Supplementary Figure 3: 

Distribution of communication scores in each 9 subcellularly resolved spatial transcriptomics datasets across the cell and its subcellular regions. Only the significantly communicating ligand receptor pairs (p-value \< 0.05) that were identified in all three regions (cell, cytoplasm and nucleus) are shown. A. Brain (AD), B. Brain (GL), C. Breast Cancer D. Lung, E. Lung (High-Res), F. Lymph Node, G. Pancreas, H. Ovarian, I. Skin.

```{r}
#| message: false
#| include: true
#| warning: false
#| label: boxplots
box_data <- data[(nuc_pval < 0.05) | (cyt_pval < 0.05) | (cell_pval < 0.05), .(cell_score, cell_P1, cyt_score, cyt_P1, nuc_score, nuc_P1, sample)]
box_data <- box_data %>%
  mutate(compartment = case_when(
    cyt_score > 0 & nuc_score > 0 & cell_score > 0 ~ "cell+cyt+nuc",
    cyt_score == 0 & nuc_score == 0 & cell_score > 0 ~ "cell",
    cyt_score > 0 & nuc_score > 0 & cell_score == 0 ~ "cyt+nuc",
    cyt_score == 0 & nuc_score > 0 & cell_score > 0 ~ "cell+nuc",
    cyt_score == 0 & nuc_score > 0 & cell_score == 0 ~ "nuc",
    cyt_score > 0 & nuc_score == 0 & cell_score == 0 ~ "cyt",
    cyt_score > 0 & nuc_score == 0 & cell_score > 0 ~ "cell+cyt",
    TRUE ~ "none"   # fallback in case none of the conditions match
  ))
box_data$sample = factor(box_data$sample, unique(box_data$sample), 
                         labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin"))
colnames(box_data) = c('cell', 'cell_P1', 'cytoplasm', 'cyt_P1', 'nucleus', 'nuc_P1', 'sample', 'compartment')
cell_cyt_data <- box_data[compartment == "cell+cyt"]
cell_nuc_data <- box_data[compartment == "cell+nuc"]
cell_cyt_nuc_data <- box_data[compartment == "cell+cyt+nuc"]

cell_cyt_long <- melt(cell_cyt_data[,.(cell, cytoplasm, sample)],measure.vars = c("cell", "cytoplasm"),variable.name = "Compartment",value.name = "Score")
cell_nuc_long <- melt(cell_nuc_data[,.(cell, nucleus, sample)],measure.vars = c("cell", "nucleus"),variable.name = "Compartment",value.name = "Score")
cell_cyt_nuc_long <- melt(cell_cyt_nuc_data[,.(cell, cytoplasm, nucleus, sample)],measure.vars = c("cell", "cytoplasm", "nucleus"),variable.name = "Compartment",value.name = "Score")
setDT(cell_cyt_long)
setDT(cell_nuc_long)
setDT(cell_cyt_nuc_long)
cell_cyt_long[, log_Score := log10(Score + 1e-10)] 
cell_nuc_long[, log_Score := log10(Score + 1e-10)] 
cell_cyt_nuc_long[, log_Score := log10(Score + 1e-10)] 

plotdir = "data/results/Final_Figures/Supplementary_figures/compare_ditributions/"
ggboxplot(
  cell_cyt_nuc_long, 
  x = "Compartment", 
  y = "log_Score",
  color = "black", 
  fill = "Compartment", 
  palette = c("#b089c0", "#97c9c8", "#ea9b8e"),  # colorblind-friendly
  add = "jitter", 
  add.params = list(alpha = 0.2, size = 0.3, width = 0.1)
) +
  facet_wrap(~ sample,ncol = 3) +
  stat_compare_means(
    comparisons = list(
      c("cell", "cytoplasm"),
      c("cell", "nucleus"),
      c("cytoplasm", "nucleus")
    ),
    method = "wilcox.test",
    size = 4
  ) +
  labs(
    y = expression(log[10]~"Communication scores"),
    x = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(color = "lightgray", fill = NA, size = 1), # box around each facet
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    legend.position = "none"
  )
ggsave(file.path(plotdir, 'cellvscytvsnuc.pdf'), height = 15)
```

### Supplementary Figure 5: 

A. Additional performance metrics of the best performing selected model D across the 9 subcellularly resolved spatial transcriptomics datasets not shown in the main Figure 2. B. Classification and regression metrics of the different base learners across the 9 subcellularly resolved spatial transcriptomics datasets. C. Classification and regression metrics of models with different feature engineering approaches selected across the 9 subcellularly resolved spatial transcriptomics datasets.

```{r}
#| message: false
#| include: true
#| warning: false
#| label: other_metrics 
ddir = "data/processed/plots_metrics" 
xgb_voting = fread(file.path(ddir, "comparison_baselearners/xgboost_voting_metrics.csv")) #used extract_metrics from utils.py
xgb_voting = xgb_voting[, c(1,2:7,11,13,14,17,20)]
class_metrics = melt(xgb_voting, id.vars = c('Dataset'), measure.vars = 2:12)

class_metrics$variable <- factor(class_metrics$variable,
  levels = unique(class_metrics$variable),
  labels = c("F1 (Class 0)", "Precision (Class 0)", "Recall (Class 0)",
  "F1 (Class 1)", "Precision (Class 1)", "Recall (Class 1)",
  "Accuracy", "Macro F1", "Macro Precision",
  "Cytoplasm RMSE", "Nucleus RMSE")
)
colors = c("#1f77b4","#3a99d9","#5cb0e0","#7fc7e6","#a3d5eb","#c7e3f0","#fdb462","#fec978","#fee0a3","#9467bd","#b087d0")
ggplot(class_metrics, aes(x = variable, y = value)) +
  geom_violin(width = 1, aes(fill = variable)) +
  geom_jitter(width = 0.1) + 
  scale_fill_manual(values = colors) + 
  geom_boxplot(color = "black", fill = "white", width = 0.05) +
  theme_classic(base_size = 20) + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = ""
    ) +
  labs( x = "", y = "Value")
ggsave("data/results/Final_Figures/Supplementary_figures/best_model_metrics/class_metrics.pdf", width = 12)

ddir = "data/processed/plots_metrics"
xgb = fread(file.path(ddir, "comparison_baselearners/xgboost_metrics.csv"))
rf = fread(file.path(ddir, "comparison_baselearners/rf_metrics.csv"))
xgb_rf = fread(file.path(ddir, "comparison_baselearners/xgboost_rf_metrics.csv"))
xgb_voting = fread(file.path(ddir, "comparison_baselearners/xgboost_voting_metrics.csv"))

xgb = xgb[, model:= "xgb"]
rf = rf[, model:= "rf"]
xgb_rf = xgb_rf[, model:= "xgb_rf"]
xgb_voting = xgb_voting[, model:= "xgb_voting"]

dt = rbind(xgb, rf, xgb_rf, xgb_voting)

heatmap_df = melt(dt, id.vars= c('model', 'Dataset'), measure.vars= c(12,15,16,18,19,21)) 
heatmap_df$variable <- factor(heatmap_df$variable,
  levels = c("auc","macro_recall","cyt_R²","nuc_R²",
             # "cyt_RMSE","nuc_RMSE",
             "cyt_NRMSE","nuc_NRMSE","geometric_mean"),
  labels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "Geometric Mean")
)

heatmap_df$Dataset <- factor(heatmap_df$Dataset,
  levels = unique(heatmap_df$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

heatmap_df$model <- factor(heatmap_df$model,
  levels = unique(heatmap_df$model),
  labels = c("Model 1", "Model 2", "Model 3", "Model 4")
)

# plot the heatmap
ggplot(heatmap_df, aes(x = variable, y = Dataset, fill = value)) +
  geom_tile(color = "black", linewidth = 0.1) + 
  scale_fill_gradientn (colours = c("white", "#cdedea", "#20b3a5")) + 
  labs(x = "", y = "Dataset") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  ) +
  coord_fixed() + 
  facet_wrap(~model, nrow = 1)

ggsave("data/results/Final_Figures/Supplementary_figures/heatmap_comparisons/heatmap_baselearners.pdf", width = 10)

ddir = "data/processed/plots_metrics"
a = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_ohe_not_separated_model2.csv"))
b = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_catboost_not_separated_model3.csv"))
c = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_predictive_features_combined_model4.csv"))
d = fread(file.path(ddir, "comparison_features/xgboost_voting_metrics_model1.csv"))

a = a[, model:= "Model A"]
b = b[, model:= "Model B"]
c = c[, model:= "Model C"]
d = d[, model:= "Model D"]

dt = rbind(a, b, c, d)

heatmap_df = melt(dt, id.vars= c('model', 'Dataset'), measure.vars= c(12,15,16,18,19,21)) 
heatmap_df$variable <- factor(heatmap_df$variable,
  levels = c("auc","macro_recall","cyt_R²","nuc_R²",
             # "cyt_RMSE","nuc_RMSE",
             "cyt_NRMSE","nuc_NRMSE","geometric_mean"),
  labels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "Geometric Mean")
)

heatmap_df$Dataset <- factor(heatmap_df$Dataset,
  levels = unique(heatmap_df$Dataset),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

heatmap_df$model <- factor(heatmap_df$model,
  levels = unique(heatmap_df$model),
  labels = c("Model A", "Model B", "Model C", "Model D")
)

# plot the heatmap
ggplot(heatmap_df, aes(x = variable, y = Dataset, fill = value)) +
  geom_tile(color = "black", linewidth = 0.1) + 
  scale_fill_gradientn (colours = c("white", "#cdc2ea", "#b585d1")) + 
  labs(x = "", y = "Dataset") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "right",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 12)
  ) +
  coord_fixed() + 
  facet_wrap(~model, nrow = 1)
ggsave("data/results/Final_Figures/Supplementary_figures/heatmap_comparisons/heatmap_fe_regression.pdf", width = 10)

```

#### Supplementary Figure 7: 

Assessing model stability by training each dataset using varying combinations of the remaining datasets. Each dataset was trained with 255 different combinations of the other eight datasets. A–B. Violin and Ridge plots showing the distribution of the number of correctly classified LR pairs across combinations for each dataset. C. Violin plots of additional performance metrics not reported in the main figure. D. Ridgeplots of the composite metric across datasets, illustrating variability when trained on different dataset combinations.

```{r}
#| message: false
#| include: true
#| warning: false
#| label: combo_dataset
row = fread("plot_supplementary7a")
ggplot(row, aes(x = sample, y = percent, fill = sample)) +
  geom_violin(width = 0.6) + 
  geom_boxplot(color = "black", fill = "white", width = 0.05) + 
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )  + 
  labs(x = "", y = "Number of correctly \n classified LR pairs") +
  scale_fill_manual(values = colors) + 
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 18),
      legend.text = element_text(size = 18),
      legend.title = element_text(size = 18),
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.text.x = element_text(angle = 45,  hjust = 1, vjust = 1),
      legend.position = "None"
    )
ggsave("data/results/Final_Figures/Supplementary_figures/counts_plots/lr_box_plot.pdf", width = 8)

dir = "data/processed/training_256/correct" # for each dataset, run_compartments.py was run and each lr_pair is scored if its correct or wrong
files = list.files(dir)
file_list = list()

for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(dir, i))
  x = nrow(dat)/255
  dat$id = rep(1:x, 255)
  data = dcast(dat, id ~ group, fun.aggregate = NULL, value.var = "correct")
  data$sample = sub("classification_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = data
}

prediction = setDT(do.call(rbind, file_list))
rownames(prediction) = 1:nrow(prediction)
col = prediction[, lapply(.SD, sum), by = sample, .SDcols = c(2:256)]
percent <- prediction[, lapply(.SD, function(x) sum(x) *100 / .N), 
                      by = sample, .SDcols = 2:256]

colsums = melt(col, id.vars = "sample")
percent = melt(percent, id.vars = "sample")
row <- prediction[, .(value = rowSums(.SD)), by = sample, .SDcols = 2:256]
percent <- prediction[, .(value = rowSums(.SD) * 100/ 256), 
                      by = sample, .SDcols = 2:256]

# p1 = ggplot(row, aes(x = value)) +
#   geom_histogram(binwidth = 5, fill = "#fb8073", color = "black") +
#   theme_classic(base_size = 12) +
#   labs(x = "Number of combinations", y = "count of LR") +
#   facet_wrap(~sample, scales = 'free')

row$sample <- factor(row$sample,
  levels = unique(row$sample),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)


colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )
ggplot(row, aes(x = value, y = sample, fill = sample)) +
  geom_density_ridges(alpha = 0.7, color = "black", scale = 1) +
  theme_classic(base_size = 16) + xlim(0,255) + 
  labs(x = "Number of correctly \n classified LR pairs", y = "Datasets") +
  scale_fill_manual(values = colors) + # customize colors
  theme(legend.position = "none") 
ggsave("data/results/Final_Figures/Supplementary_figures/counts_plots/lr_counts_plot.pdf")



# other metrics 
class_dir <- "data/processed/scores_training/classification_metrics/" # metrics from run_compartments.py calculated using extract_metrics in utils.py
files = list.files(class_dir)
file_list = list()

for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(class_dir, i))
  dat$sample = sub("classification_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}

# join with the metrics with perclass metrics 
metrics = do.call(rbind, file_list)
perclass_metrics = fread("data/processed/scores_training/final_classification_metrics.csv")
colnames(perclass_metrics)[11] = "group"
perclass_metrics$sample = sub("classification_spe_list_(.*)_cellchat\\.csv$", "\\1", perclass_metrics$dataset)
perclass_metrics = perclass_metrics[,-c(9,10,12)]
metrics = metrics[perclass_metrics, on = .(sample, group)]
metrics_further = metrics


# Regression metrics 
cyt_reg_dir <- "data/processed/scores_training/cyt_metrics" # metrics from run_compartments.py calculated using extract_metrics in utils.py
nuc_reg_dir <- "data/processed/scores_training/nuc_metrics"# metrics from run_compartments.py calculated using extract_metrics in utils.py
files = list.files(cyt_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(cyt_reg_dir, i))
  dat$sample = sub("cyt_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
cyt_metrics = do.call(rbind, file_list)
colnames(cyt_metrics)[1:3] = paste0("cyt_",colnames(cyt_metrics)[1:3])

files = list.files(nuc_reg_dir)
file_list = list()
for (i in files){
  # read the files and store in a list
  dat <- fread(file.path(nuc_reg_dir, i))
  dat$sample = sub("nuc_metrics_spe_list_(.*)_cellchat\\.csv", "\\1", i)
  file_list[[i]] = dat
}
nuc_metrics = do.call(rbind, file_list)
colnames(nuc_metrics)[1:3] = paste0("nuc_",colnames(nuc_metrics)[1:3])

regression_metrics = cyt_metrics[nuc_metrics, on = .(group, sample)]
regression_metrics$`cyt_R²` = ifelse(regression_metrics$`cyt_R²`<0, 0, regression_metrics$`cyt_R²`)
regression_metrics$`nuc_R²` = ifelse(regression_metrics$`nuc_R²`<0, 0, regression_metrics$`nuc_R²`)
regression_metrics_further = regression_metrics

regression_metrics = melt(regression_metrics, id.vars = c('sample', "group"), measure.vars = c(1:3, 6:8))
regression_metrics$facets <- with(regression_metrics,
  ifelse(variable %in% levels(regression_metrics$variable)[c(1,4)], "R²",
  ifelse(variable %in% levels(regression_metrics$variable)[c(3,6)], "NRMSE", 'RMSE'))
)

regression_metrics$facets = factor(regression_metrics$facets, levels = c("R²", "NRMSE", 'RMSE'))



metrics = metrics_further[regression_metrics_further, on = .(group, sample)]
combo_metrics = metrics[, c(9,7,8,12,10,11,1,3,13,17,19)]
dt = melt(combo_metrics)
dt$variable <- factor(dt$variable,
  levels = unique(dt$variable),
  labels = c("F1 (Class 0)", "Precision (Class 0)", "Recall (Class 0)",
  "F1 (Class 1)", "Precision (Class 1)", "Recall (Class 1)",
  "Accuracy", "Macro F1", "Macro Precision",
  "Cytoplasm RMSE", "Nucleus RMSE")
)

colors = c("#1f77b4","#3a99d9","#5cb0e0","#7fc7e6","#a3d5eb","#c7e3f0","#fdb462","#fec978","#fee0a3","#9467bd","#b087d0")
ggplot(dt, aes(x = variable, y = value)) +
  geom_violin(width = 1, aes(fill = variable)) + 
  scale_fill_manual(values = colors) + 
  geom_boxplot(color = "black", fill = "white", width = 0.2) +
  theme_classic(base_size = 18) +
    theme(
      axis.text = element_text(size = 16),
      axis.title = element_text(size = 18),
      legend.position = "",
      plot.title = element_text(size = 18, hjust = 0.5),
      axis.text.x = element_text(angle = 45,  hjust = 1, vjust = 1)
    ) +
  labs( x = "", y = "Value") + coord_cartesian(ylim=c(0,1)) 
ggsave("data/results/Final_Figures/Supplementary_figures/counts_plots/combo_metrics.pdf", width = 8)

# split the composite metric alone
combo_metrics = metrics[, c(4,5,6,14,15,18,17,20)]
combo_metrics[ , geometric_mean := abs(roc_auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]


combo_metrics = combo_metrics[,c(3,9)]


combo_metrics$sample <- factor(combo_metrics$sample,
  levels = unique(combo_metrics$sample),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)


colors = c("#f57b6c", "#d4dda6", "#78c2e3", "#f2a8da", "#f3b773", "#84c3b7", "#fbe79e", "#b8aeea", "#88d7da" )
ggplot(combo_metrics, aes(x = geometric_mean, y = sample, fill = sample)) +
  geom_density_ridges(alpha = 0.7, color = "black", scale = 1) +
  theme_classic(base_size = 20) + xlim(0,1) + 
  labs(x = "Composite metric", y = "Datasets") +
  scale_fill_manual(values = colors) + # customize colors
  theme(legend.position = "none") 


ggsave("data/results/Final_Figures/Supplementary_figures/counts_plots/composite_metric.pdf")
```

#### Supplementary Figure 6: 
Performance metrics of the models trained with and without spatial features across different combinations of datasets. There were a total of 4590 models trained, with each dataset (n = 9) trained across 255 combinations of other datasets.

```{r}
#| message: false
#| include: true
#| warning: false
#| label: combo_metrics 
metrics = fread("data/processed/metrics_all_combinations_trainings_10032025.csv") # metrics from runnining experiments/run_combinatiions.py and extract_metrics from utils.py
df_long <- metrics[,c(16,24,1,6,3,8,9,12)]
colnames(df_long) = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "condition",
             "combo_size")
df_long <- df_long %>%
   pivot_longer(cols = colnames(df_long)[c(1:6)],
                names_to = "metric", values_to = "value")

df_summary <- df_long %>%
  group_by(combo_size, condition, metric) %>%
  summarise(mean_val = mean(value, na.rm = TRUE),
            sd_val = sd(value, na.rm = TRUE)/sqrt(n()), .groups = "drop")

df_summary$metric = factor(df_summary$metric,
                           levels = c("AUC",
             "Recall (Macro)",
             "R² – Cytoplasm",
             "R² – Nucleus",
             # "RMSE – Cytoplasm",
             # "RMSE – Nucleus",
             "NRMSE – Cytoplasm",
             "NRMSE – Nucleus",
             "condition",
             "combo_size")
                             )

colors = c('with_spatial' = "#b715b7", "without_spatial" = "#14bf81")
ggplot(df_summary, aes(x = combo_size, y = mean_val, color = condition, group = condition)) +
  geom_line() +
  geom_point() +
  scale_color_manual(
  name   = "Model type",
  values = colors,
  labels = c(
    with_spatial    = "With spatial information",
    without_spatial = "Without spatial information"
  )
) + 
  geom_errorbar(aes(ymin = mean_val - sd_val, ymax = mean_val + sd_val), width = 0.2) +
  # facet_wrap(~metric, scales = "free_y", ncol = 4) +
  labs( x = "Number of training datasets", y = "Value") + 
  facet_wrap(~metric, scales = "free",ncol =3) + theme_classic(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 90, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
    axis.text.y = element_text(size = 12),
    strip.background = element_blank(), strip.placement = "outside"
  ) 

ggsave("data/results/Final_Figures/Supplementary_figures/best_model_metrics/spatial_features.pdf", height = 10)
```

#### Supplementary Figure 8: 
Assessing spatial characteristics of ligand–receptor communication scores across datasets. A. Correlation of region-specific communication scores with the whole-cell score, comparing models trained with and without spatial features. B. Distribution of pairwise differences in model predictions across all combinations of training datasets for a given candidate dataset. Brain (AD) and Lung datasets, which exhibit high spatial correlation, are included. C. Network plot showing the summed ligand–receptor (LR) scores in the predicted cytoplasm and nucleus when the model is applied to single-cell data.

```{r}
#| message: false
#| include: true
#| warning: false
#| label: single_cell 
# load output from experiments/run_singlecell.py

df =  fread("data/training_data.csv/training_data.csv") 
# Cytoplasmic comparison
corr_no_spatial <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_P1, cyt_P1, use="complete.obs", method="pearson"),
            .groups="drop") %>%
  mutate(type = "Without spatial")
corr_no_spatial$sample <- c("Brain (AD)","Brain (GL)","Breast Cancer","Lung",
                     "Lung (High-Res)","Lymph Node","Ovarian","Pancreas","Skin")
corr_spatial <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_score, cyt_score, use="complete.obs", method="pearson"),
            .groups="drop") %>%
  mutate(type = "With spatial")
corr_spatial$sample <- c("Brain (AD)","Brain (GL)","Breast Cancer","Lung",
                     "Lung (High-Res)","Lymph Node","Ovarian","Pancreas","Skin")
cyt_corr <- bind_rows(corr_no_spatial, corr_spatial)

# Nuclear comparison
corr_no_spatial <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_P1, nuc_P1, use="complete.obs", method="pearson"),
            .groups="drop") %>%
  mutate(type = "Without spatial")
corr_no_spatial$sample <- c("Brain (AD)","Brain (GL)","Breast Cancer","Lung",
                     "Lung (High-Res)","Lymph Node","Ovarian","Pancreas","Skin")
corr_spatial <- df %>%
  group_by(sample) %>%
  summarise(corr = cor(cell_score, nuc_score, use="complete.obs", method="pearson"),
            .groups="drop") %>%
  mutate(type = "With spatial")
corr_spatial$sample <- c("Brain (AD)","Brain (GL)","Breast Cancer","Lung",
                     "Lung (High-Res)","Lymph Node","Ovarian","Pancreas","Skin")
nuc_corr <- bind_rows(corr_no_spatial, corr_spatial)

# scatter plot  join the nuc and cytoplasm 
colnames(nuc_corr)[2] = "Nucleus"
colnames(cyt_corr)[2] = "Cytoplasm"
# inner join
setDT(nuc_corr)
setDT(cyt_corr)
data = cyt_corr[nuc_corr, on = c('sample', 'type')]
data_plot = melt(data, id.vars = c(1,3), measure.vars = c(2,4), variable.name = "Region")
wide <- dcast(
  data_plot,
  sample + variable ~ type,
  value.var = "value"
)

colnames(wide)[1:2] = c("Dataset","Region") 
ggplot(wide, aes(x=`With spatial`, y=`Without spatial`, color=Region, shape = Dataset)) + geom_point() +
  scale_color_manual(values = c("#D2665A", "#F2B28C")) + 
  scale_shape_manual(values = 1:9)

ggsave("data/results/Final_Figures/Supplementary_figures/single_cell/correlation.pdf")


metrics = fread("data/processed/metrics_all_combinations_trainings_10032025.csv") # extract_metrics using utils.py from the run_compartments.py
metrics[ , geometric_mean := (roc_auc 
                                 * (1 - ((cyt_NRMSE + nuc_NRMSE)/2)) 
                                    * macro_recall 
                                    * ((`cyt_R²` + `nuc_R²`)/2)
                                 )**(1/4)]


df_long <- metrics[,c(9,28,10,11, 12)]

spatial = df_long[condition == "with_spatial",]
no_spatial = df_long[condition == "without_spatial",]
#check
identical(spatial[,3:5], no_spatial[,3:5])

df_long = spatial[,3:5]
df_long$difference = spatial$geometric_mean - no_spatial$geometric_mean
df_long = na.omit(df_long)

df_long$test_name <- factor(df_long$test_name,
  levels = unique(df_long$test_name),
  labels = c("Brain (AD)",
             "Brain (GL)",
             "Breast Cancer",
             "Lung",
             "Lung (High-Res)",
             "Lymph Node",
             "Ovarian",
             "Pancreas",
             "Skin")
)

ggplot(df_long, aes(x = test_name, y = difference, fill = test_name)) +
  geom_boxplot(outliers = TRUE) +
  scale_fill_manual(values = colors) + 
  # facet_wrap(~metric, scales = "free_y", ncol = 4) +
  labs( x = "", y = "Difference in geometric means") + 
  theme_classic(base_size = 20) +
  theme(
    axis.text.x = element_text(angle = 45, size = 12, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "",
    strip.text = element_text(face = "bold", size = 10, hjust = 0.5),
axis.text.y = element_text(size = 18)
  )
ggsave("data/results/Final_Figures/Supplementary_figures/single_cell/boxplot_difference.pdf", width =8)

cytoplasm = fread("data/results/predictions/scatter_data_cyt_single_cell.csv")
nucleus = fread("data/results/predictions/scatter_data_nuc_single_cell.csv")
count_function = function(data){
  shift_val = min(0, min(data$pred))
  data$score = data$pred - shift_val
  # compute the network plot
  links = data.table(
    source= data$source,
    target= data$target,
    importance= data$score
    )
  edges_count <- links[, .(count = .N), by = .(source, target)]
  edges_sum   <- links[, .(sum = sum(importance, na.rm = TRUE)), 
                     by = .(source, target)]
  nodes <- unique(c(links$source, links$target))
  all_pairs <- CJ(source = nodes, target = nodes, unique=TRUE)
  matrix <- merge(all_pairs, edges_count, by=c("source","target"), all.x=TRUE)
  matrix <- merge(matrix, edges_sum,   by=c("source","target"), all.x=TRUE)
  matrix[is.na(count), count := 0]
  matrix[is.na(sum),   sum   := 0]
}
# for each of the group, shift the prediction and then construct adjacency and from the network plot
data = cytoplasm
# shift the scale 
shift_val = min(0, min(data$pred))
data$score = data$pred - shift_val
# compute the network plot
links = data.table(
  source= data$source,
  target= data$target,
  importance= data$score
  )
edges_count <- links[, .(weight = .N), by = .(source, target)]
edges_sum   <- links[, .(weight = sum(importance, na.rm = TRUE)), 
                   by = .(source, target)]
nodes = unique(c(data$source, data$target))
network <- graph_from_data_frame(d=edges_count, vertices=nodes, directed=F) 
coul  <- brewer.pal(9, "Set1") 
my_color = colorRampPalette(coul)(20)
pdf("data/results/Final_Figures/Final_figure4/scnetworks.pdf", width = 15)
par(mfrow = c(1,2))
# plot(network, vertex.color=coul, edge.width=0.2*(E(network)$weight), main = "Cyt Network (counts)", layout=layout.circle)
# plots_count = recordPlot()
network <- graph_from_data_frame(d=edges_sum, vertices=nodes, directed=F) 
plot(network, vertex.color=coul, edge.width=0.2*(E(network)$weight), main = "Cyt Network (sum)", layout=layout.circle)
plots_sum = recordPlot()

data = nucleus
# shift the scale 
shift_val = min(0, min(data$pred))
data$score = data$pred - shift_val
# compute the network plot
links = data.table(
  source= data$source,
  target= data$target,
  importance= data$score
  )
edges_count <- links[, .(weight = .N), by = .(source, target)]
edges_sum   <- links[, .(weight = sum(importance, na.rm = TRUE)), 
                   by = .(source, target)]
nodes = unique(c(data$source, data$target))
network <- graph_from_data_frame(d=edges_count, vertices=nodes, directed=F) 
coul  <- brewer.pal(9, "Set1") 
my_color = colorRampPalette(coul)(20)
# plot(network, vertex.color=coul, edge.width=0.2*(E(network)$weight), main = "Nuc Network (counts)", layout=layout.circle)
# plots_count = recordPlot()
network <- graph_from_data_frame(d=edges_sum, vertices=nodes, directed=F) 
plot(network, vertex.color=coul, edge.width=0.2*(E(network)$weight), main = "Nuc Network (sum):", layout=layout.circle)
plots_sum = recordPlot()
dev.off()


```

